### 2019-08-07 paper/main.tex
initial checkin

\documentclass[modern]{aastex63}

\shorttitle{RR Lyrae in the Pal 5 stream}
\shortauthors{People et al.}

\begin{document}

\title{RR Lyrae in the Palomar 5 stellar stream}

\author{Order TBD}

\author{Cecilia Mateu}

\author[0000-0003-0872-7098]{Adrian~M.~Price-Whelan}
\affiliation{Center for Computational Astrophysics, Flatiron Institute,
             Simons Foundation, 162 Fifth Avenue, New York, NY 10010, USA}
\affiliation{Department of Astrophysical Sciences,
             Princeton University, Princeton, NJ 08544, USA}
% \email{aprice-whelan@flatironinstitute.org}
% \correspondingauthor{Adrian M. Price-Whelan}

\author{more people}

\begin{abstract}
Stuff!
\end{abstract}

\keywords{}

\section{Introduction} \label{sec:intro}

Words.

\section{Data} \label{sec:data}

Words.

% \subsection{Visit velocity error calibration} \label{sec:visitcalib}

\section{Results} \label{sec:results}

\software{
    Astropy \citep{astropy, astropy:2018},
    emcee \citep{emcee},
    gala \citep{gala},
    IPython \citep{ipython}
}

\bibliographystyle{aasjournal}
\bibliography{refs}

\end{document}

### 2019-08-20 paper/main.tex

@@
-\documentclass[modern]{aastex63}
+% \documentclass[modern]{aastex63}
+\documentclass[twocolumn]{aastex63}
 
@@
 \shortauthors{People et al.}
+\newcommand{\sa}[1]{{\color{teal} SP: #1}}
+
+\input{paper/defs.tex}
 
@@
 
-\title{RR Lyrae in the Palomar 5 stellar stream}
-
-\author{Order TBD}
-
-\author{Cecilia Mateu}
+\title{Kinematics of the Palomar 5 stellar stream from RR Lyrae stars}
 
@@
 
-\author{more people}
+\author[0000-0002-6330-2394]{Cecilia~Mateu}
+\affiliation{Departamento de Astronom\'ia, Facultad de Ciencias, Universidad de la Rep\'ublica, Igu\'a 4225, 14000, Montevideo, Uruguay}
+
+\author{Giuliano~Iorio}
+\affiliation{Institute of Astronomy, University of Cambridge, Madingley Road, Cambridge CB3 0HA, UK}
+
+\author{Sarah~Pearson}
+\affiliation{Center for Computational Astrophysics, Flatiron Institute,
+             Simons Foundation, 162 Fifth Avenue, New York, NY 10010, USA}
+
+\author{Vasily~Belokurov}
+\affiliation{Institute of Astronomy, University of Cambridge, Madingley Road, Cambridge CB3 0HA, UK}
+
+% \author{more people?}
 
 \begin{abstract}
-Stuff!
+\todo{This is just placeholder text!}
+% Context
+Stellar streams are important.
+The Palomar 5 stellar stream is a prototypical example: It connects back to the progenitor cluster.
+RR Lyrae-type (RRL) stars are useful because they are standard candles and their distances can be inferred with small errors ($\sim7\%$)
+% Aims
+While the Pal 5 cluster is known to host a small number (5--7) of RRL stars, only two RRL stars have since been found in the long tidal tails that form the Pal 5 stream.
+% Methods
+Here we use a large catalog of RRL stars released as a part of Gaia DR2 to search for RRL stars in the Pal 5 stream.
+By building a probabilistic model of the cluster and stream in proper motion and distance, we find XX RRL consistent with being members of the stream.
+% Results
+The number of Pal 5 RRL is therefore XX times larger in the tails than in the remnant cluster: This changes our interpretation of ...blah.
+% Conclusions
+Stuff.
 \end{abstract}
 
-\keywords{}
+\keywords{globular clusters: individual: Palomar 5 -- 
+Galaxy: halo --
+stars: variables: RR Lyrae --
+Galaxy: structure}
 
@@
 
-Words.
+Globular clusters are destroyed as they orbit within the Milky Way.
+The $\sim$150 bound globular clusters we presently see throughout the Galaxy \citep{Harris2010} are therefore thought to be surviving relics of a much larger initial population, most of which were destroyed (primarily) through a combination of relaxation / evaporation and gravitational shocking from the disk and bulge \citep[e.g.,][]{Gnedin:1997, Gnedin:2014}.
+As stellar systems are destroyed---i.e., as stars are tidally stripped from their progenitor---the tidal debris forms tails of matter that both lead and trail the remnant cluster, producing a stellar stream that may persist even after the progenitor is fully destroyed \citep[e.g.,][]{Johnston:1996, Eyre:2011}.
+
+Stellar streams are useful objects because they almost\footnote{More precisely, the phase-space distribution of stars along a stream encodes the orbital history of the progenitor system \citep[e.g.,][]{Sanders:2013}.} delineate the past and future trajectory of their progenitor system, offering information about orbits that can be used to infer the mass distribution from the single kinematic snapshot of the Galaxy that we are afforded \citep[e.g.,][]{Johnston:1999, Binney:2008, Eyre:2009, Sanders:2013, PriceWhelan:2014, Bonaca:2018}.
+Many globular cluster stellar streams have been discovered over the last few decades \citep[see, e.g.,][]{Grillmair:2016, Shipp:2018}, in large part because of large-area, deep, multi-band imaging surveys such as the Sloan Digital Sky Survey \citep[SDSS;][]{York:2000}, Pan-STARRS PS1 survey \citep{Chambers:2016}, and Dark Energy Survey \citep{DES:2016}.
+A subset of the known streams have been precisely characterized and further studied \citep[e.g.,][]{PriceWhelan:2018, Malhan:2018, Shipp:2019} using kinematic data from the recent data release 2 (\DR{2}) of the \Gaia\ mission \citep{Gaia:2016, Gaia:2018}.
+Of the currently known $\sim$60 candidate stellar streams found throughout the Milky Way, few have known progenitor systems.
+Conversely, of the $\sim$150 known globular clusters, only $\lesssim 20\%$ have purported tidal tails \citep[e.g.,][]{Leon:2000, Kundu:2019}, but even these are mostly low density and tenuous \citep[as might be expected, e.g.,][]{Balbinot:2018}.
+A prominent exception to these statements is the Palomar 5 (Pal 5) globular cluster and stream.
+
+The Pal 5 stream---at a heliocentric distance of $d \sim 20~\kpc$---was discovered using early SDSS imaging, which reached well below the main sequence turnoff of the stream over a large area surrounding the cluster \citep{Odenkirchen:2001, Rockosi:2002}.
+Since then, radial velocities have been measured for a handful of giant branch stars in the cluster and along the stream \citep{Odenkirchen:2002, Odenkirchen:2009, Ibata:2017}, and deeper imaging data has been used to map the tails at higher signal-to-noise \citep{Bernard:2016, Ibata:2016, Bonaca:2019}.
+The detailed but still limited kinematic data for Pal 5 has made it a canonical example of tidal destruction and a useful tool for constraining Galactic structure.
+For example, using the sky track and radial velocity information, the stream has been used to measure the enclosed mass of the Milky Way \citep{Kuepper:2015, Bovy:2016}.
+Using deep photometry, the density variations along the stream have been used to place limits on the abundance of massive substructures in the Galactic halo \citep{Erkal:2017}.
+
+However, a number of peculiar aspects of the stream still remain unexplained in detail.
+For example, the leading and trailing tails have significantly different total star counts and lengths \citep{Dehnen2004, Bernard:2016}, and the stream track, width, and density vary significantly over the full extent of the stream \citep{Ibata:2016, Bonaca:2019}.
+Given that Pal 5's pericenter is well within the Galactic disk ($r_{\textrm{peri}} \sim 7$--$8~\kpc$), these features could be a sign of perturbations from molecular clouds \citep[e.g.,][]{Amorisco:2016}, interaction with the Galactic bar \citep[e.g.,][]{Pearson:2017}, signatures of past dark matter subhalo encounters \citep[e.g.,][]{Erkal:2017}, or some combination of all of these effects.
+Mapping the stream in all six phase-space dimensions---i.e., combining deep photometry with radial velocity, distance, and proper motion measurements---will enable and motivate improved dynamical models of the stream and more precise constraints on the Galactic mass distribution \citep[e.g.,][]{PriceWhelan:2013}.
+However, at its present distance, main sequence stars are barely detected and have large astrometric uncertainties in \Gaia\ \DR{2}, and other stellar tracers (e.g., giant branch stars) are not numerous enough or precise enough distance indicators to provide useful relative distance information along the stream.
+Enter RR Lyrae-type stars.
+
+\todo{Cecilia: write this paragraph? you probably have some canned text?}
+RR Lyrae stars (RRLs) are pulsating horizontal branch stars that are intrinsically more luminous than main sequence stars, but about $\sim 10,000$ time less numerous. 
+HB = good distance indicator.
+The horizontal branch in Pal 5 is bright enough to be observed by Gaia, meaning we can measure kinematic properties of the full stream...
+\citep{Ibata:2017} found some RRL in pal 5, but no proper motions.
+This is useful for modeling...
+Many stream candidates found in RRL alone \citep{Mateu:2018}.
+
+In this Article, we construct a probabilistic model for kinematic properties of RRLs in the vicinity of the Pal 5 stream to determine membership probabilities and thus map the distance and proper motion trends in the stream.
+In \sectionname~\ref{sec:data}, we describe the source RR Lyrae catalog we use to search for Pal 5 members.
+In \sectionname~\ref{sec:membership}, we implement a probabilistic membership model to identify candidate members of the stream using distance and astrometric information.
+In \sectionname~\ref{sec:results}, we ...
 
@@
 
-Words.
+% From \Gaia~DR2 we have combined the two available RRL catalogues: the Specific Objects Study \citep[SOS][]{Clementini2018} and vari-classifier \citep[VC][]{Holl2018,Rimoldini2018}.
+The resulting VC+SOS catalogue contains a total of 228\,904, with all-sky coverage down to the mission's limiting magnitude $G=20.7$.
+After filtering out stars with contaminated photometry \verb+phot_bp_rp_excess_factor+>2 or null, as recommended in \citep{Rimoldini2018}, the VC+SOS catalogue contains a total of 175\,164 RR Lyrae.
+For the 122\,747 in SOS, which have undergone extensive light curve fitting, periods, amplitudes and other fitting parameters have been provided by \citet{Clementini2018}.
+The remaining 52\,417 stars (from the VC catalogue) did not have enough data points in \Gaia~DR2 to go into the SOS pipeline, so periods and other fitting parameters are not yet available for these stars.
 
-% \subsection{Visit velocity error calibration} \label{sec:visitcalib}
+To search for RRLs in the vicinity of Pal 5, we use a catalog of RRL stars from the PS1 survey that spans $\sim$3/4 of the sky (all sky north of declination $\delta > -30\degr$) and contains 61\,795 RRLs identified from their variability \citep{Sesar2017b}.
+In detail, the \emph{bona fide} RRLs in the catalog were classified using multi-epoch data in the $grizy$ photometric bands, with a typical number of 12 observations per filter, and has a limiting magnitude $r\sim21$ (corresponding to $G~21$ for the typical colors of RRLs).
+Periods and amplitudes in all five bands are available for all stars in this catalog. 
+We have chosen to use this catalog instead of the recently released \Gaia~RRL catalogs \citep[VariClassifier and Specific Object Studies][]{Holl2018, Rimoldini2018, Clementini2018} as these are subject to higher incompleteness in this particular area of the sky due to the \Gaia\ scanning law.
+Nevertheless, the vast majority (61\,625) of PS1's \emph{bona fide} RRLs have astrometric information available in \Gaia~\DR{2}.
+
+\citet{Sesar2017b} provide distances to the RRLs based on an $i$-band Period-Luminosity-Metallicity (PLZ) relation. 
+This has two key advantages when compared to optical bands: The PLZ relation in the $i$-band only weakly depends on metallicity, and the impact of dust extinction is less severe at longer wavelengths.
+However, distances reported for the \rrc~stars in Table~5 of \citet{Sesar2017b} have a systematic offset with respect to the \typeab~stars. 
+This offset comes from having applied the same PLZ to \typeab~and \typec~stars. 
+This is corrected by using the `fundamentalized' period in the PLZ relation for the \rrc~stars, which is computed as $\log{P_F} = \log P + 0.126$ \citep[following][]{Braga2016}.
+In what follows, we use the $i$-band PLZ distances for the RRLs.
+\citet{Sesar2017b} estimate their overall distance precision to be 3\%, of which $\sim2$\% and $\sim1$\% correspond to systematic and random uncertainties, respectively.
+
+We cross-match the full PS1 RRL catalog with \Gaia~DR2 \citep{Gaia:2018}, using a 3" sky separation tolerance, in order to retrieve astrometric information for these stars. 
+Although not all PS1 RRLs have been identified as RRLs from \Gaia\ photometry, all of the PS1 RRL are present in the main point source catalog.
+
+%The combined \Gaia+PS1 catalogue has a total of XXX RRL stars.
+
+% \subsection{RRL Distances}
+
+% We assume a single intrinsic magnitude $M_G=+0.63$ for all RRLs, corresponding to the mean halo metallicity $\FeH=-1.5$ according to the calibration by \citet{Muraveva2018}, and compute distances to all stars based on their $G$-band intensity-averaged magnitudes: we use \verb+int_average_g+ for stars in VC+SOS and the intensity-averaged $g$ reported by \citet{Sesar2017b} converted to the G-band using the transformation of Eq.~XX from \citet{Evans2018}.
+To correct for the effect of dust extinction in the apparent magnitudes, we obtain the colour excess $E(B-V)$ for each RRL from the \citet{Schlegel1998} maps, and compute the V-band extinction as $A_V=2.742E(B-V)$ assuming an $R_V$=3.1 \citep{Schlafly2011}.
+Finally, the extinction is transformed to the $G$ band using the calibration from Carrasco (2018, private communication):
+% \begin{equation}
+% \begin{split}
+% f(\Gbp-\Grp, A_V) \equiv & a + b \ (\Gbp-\Grp) + c \ (\Gbp-\Grp)^2 \\
+%                          & +\ d \  A_V + e\  A_V \ (\Gbp-\Grp)
+% \end{split}
+% \end{equation}
+% \begin{equation}
+%  A_G = f(\Gbp-\Grp, A_V) \cdot A_V
+% \end{equation}
+% where $a=0.97883$, $b=-0.14365$, $c=0.011077$, $d=0.034842$, and $e=-0.0041448$; assuming a mean colour $BP-RP=0.7$ for RRLs.
+
+% \todo{Describe uncertainties and justification of relative precision vs. absolute error}
+
+\begin{figure*}[th]
+\begin{center}
+\includegraphics[width=\textwidth]{paper/proper-motion.pdf}
+\caption{\textbf{Left}: Proper motions of RRLs in the region $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$ with distances in the range $18 < d < 25~\kpc$.
+\textbf{Middle}: The same, but for RRLs within $1~\degr$ of the mean sky track of Pal 5 \citep{Bonaca:2019}.
+The over-density of stars near $(-2.7, -2.7)$ is the stream, and the two colored points show the Pal 5 cluster proper motions from \citet{Vasiliev:2019} and \citet{Fritz:2015}, respectively.
+\textbf{Right}: The same, but for RRLs excluding $2~\degr$ around the Pal 5 sky track.}
+\label{fig:pm}
+\end{center}
+\end{figure*}
+
+\section{Determining stream membership} \label{sec:membership}
+
+In order to search for new RRL stars in the Pal 5 stellar stream, we construct a probabilistic model of the stream and background stellar distribution (simultaneously) using proper motion $(\mu_\alpha, \mu_\delta)$ and distance $d$ data for all RRL in the sky window $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$, where $(\alpha, \delta)$ are right ascension and declination.
+We \emph{do not} include the known spatial extent or morphology of the Pal 5 stream in our membership model for the stream: We instead later use the sky positions to validate the modeling procedure and selection criteria we then use to define a sample of probable Pal 5 stream RRLs.
+This model ultimately enables us to compute stream membership probabilities for RRL stars in a way that fully incorporates \Gaia\ error properties, while simultaneously modeling and marginalizing over uncertainties in the background RRL distribution.
+
+To start, we use the sky track and width of Pal 5 from \citet{Bonaca:2019} to mask out a $\sim4^\circ$-wide region around the stream---and an extrapolation of the known extent of the stream---in order to define a ``background'' region that should be devoid of Pal 5 member stars.
+\figurename~\ref{fig:pm} demonstrates that Pal 5 stream members appear in the RRL catalog: This figure shows the \Gaia\ \DR{2} proper motions for all RRLs in the sky window defined above within the distance range $18 < d < 25~\kpc$, then the same for RRLs in a tight selection around the observed stream track, and then finally the same for RRLs \emph{excluding} the stream region.
+We use RRL stars in the background region to construct an error-deconvolved density model in the 3D space of proper motion components and distance.
+We use proper motions and uncertainties from \Gaia---using the provided covariance matrix between proper motion components---and distances and uncertainties as derived above (\sectionname~\ref{sec:data}).
+We use \project{extreme deconvolution} \citep[\acronym{XD};][]{Bovy:XD} to optimize the parameters of a Gaussian Mixture Model (GMM) representation of the (deconvolved) background RRL density.
+We first fit the background model (with \acronym{XD}) using between $3$--$12$ mixture components and evaluate the Bayesian Information Criterion (BIC) for the optimal parameters in each case.
+From this, we find that the BIC is minimized for 6 mixture components: We therefore use 6 mixture components and fix the GMM model parameters to the \acronym{XD}-optimized values and use this as our background model.
+We evaluate the background likelihood of all stars in the full sky region (defined above) and refer to this as $p_{{\rm bg}, n}$ (for the $n$th star) below.
+
+To represent the Pal 5 stream, we use a single Gaussian component with a fixed dispersion of $0.05~\masyr$ for the proper motion components ($\approx 5~\kms$ at the distance of Pal 5) and $0.2~\kpc$ in distance.
+However, we allow the mean to vary independently in each component as a function of $\phi_1$, a longitude coordinate (along the stream) as defined in \citet{Bonaca:2019}.
+Based on the Pal 5 stream models from \citet{Bonaca:2019}, over the range of $\phi_1 \in (-20^\circ, 15^\circ$, the mean stream trends in $\mu_\alpha$, $\mu_\delta$, and distance can be well-approximated (within our observational uncertainties) by a quadratic function in $\phi_1$.
+The mean of our stream component, $\bs{x}$, is therefore given by
+\begin{equation}
+    \bs{x} = \begin{pmatrix}
+        \mu_\alpha^* + b_\alpha \, \phi_1 + c_\alpha \, \phi_1^2\\
+        \mu_\delta^* + b_\delta \, \phi_1 + c_\delta \, \phi_1^2\\
+        d^* + b_d \, \phi_1 + c_d \, \phi_1 \label{eq:meanpoly}
+    \end{pmatrix}
+\end{equation}
+where $\bs{\theta} = (\mu_\alpha^*, \mu_\delta^*, d^*, b_\alpha, b_\delta, b_d, c_\alpha, c_\delta, c_d)$ are free parameters.
+
+Given the above background model, and this model for the stream track, the full likelihood for an RRL star, $n$, with data $D_n = (\mu_\alpha, \mu_\delta, d)_n$ is represented as a mixture of the two components with mixture weight $f$,
+\begin{equation}
+    p(D_n \given \bs{\theta}, f) = f \, \mathcal{N}(D_n \given \bs{x}, \mat{C}) + (1-f)\,p_{{\rm bg}, n}
+\end{equation}
+where $\mathcal{N}(\cdot \given \bs{x}, \mat{C})$ is the multivariate normal distribution with mean $\bs{x}$ and covariance matrix $\mat{C}$.
+
+We use Markov Chain Monte Carlo (MCMC) to generate samples over the parameters $(\bs{\theta}, f)$ using the posterior probability
+\begin{equation}
+    p(\bs{\theta}, f \given \{D_n\}) \propto p(\bs{\theta}, f) \, \prod_n^N p(D_n \given \bs{\theta}, f)
+\end{equation}
+where $p(\bs{\theta}, f)$ is the prior probability distribution over the parameters.
+We assume that our prior is separable in all of our parameters: We use a uniform prior for the mixture weight, $f$, such that $p(f) = \mathcal{U}(0, 1)$, Gaussian priors on the mean proper motion and distance using the Pal 5 cluster proper motion measurements from \citet{Vasiliev:2019} and the Pal 5 distance from \citet{Kuepper:2015}.
+We use broad uniform priors on the linear and quadratic polynomial coefficients defined in \equationname~\ref{eq:meanpoly}.
+We use the affine-invariant ensemble MCMC sampler \texttt{emcee} \citep{emcee} to sample from the posterior probability distribution given above.
+We run \texttt{emcee} with 80 walkers for an initial 1024 steps to ``burn-in'' the sampler, then reset and restart the sampler for another 2048 steps.
+We compute the Gelman-Rubin convergence statistic $\mathcal{R}$ \citep{Gelman:1992} to verify that $\mathcal{R} < 1.1$ for all walkers and assume that the chains have converged.
+We then thin the chains by saving only every 128th step, leaving us with 1,280 posterior samples in our membership model parameters, $\bs{\theta}$.
+For each star, we also compute posterior membership probabilities using the samples following \citet{DFM:blog}.
+
+\figurename~\ref{fig:trackmembers} shows a summary of the inferred membership model properties in sky position (shown in Pal 5 coordinates $\phi_1, \phi_2$), distance, and proper motion components.
+
+\begin{figure*}[t]
+\begin{center}
+\includegraphics[width=\textwidth]{paper/tracks.pdf}
+\caption{todo}
+\label{fig:trackmembers}
+\end{center}
+\end{figure*}
 
 \section{Results} \label{sec:results}
+
+\begin{figure*}[t]
+\begin{center}
+\includegraphics[width=\textwidth]{paper/members.pdf}
+\caption{Sky positions of RRL...todo}
+\label{fig:members}
+\end{center}
+\end{figure*}
+
+\subsection{RR Lyrae stars known to be associated with Pal 5 or it's tails}
+
+Pal~5 has 5 known RRL stars (V1-V5), all of type \typec, as reported in the \citet{Clement2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{SawyerHogg1973}.
+Star V2 has a  systemic radial velocity of  $-65\pm16$~\kms~measured by \citet{Vivas2005}.
+
+In their analysis of the halo density profile \citet{Vivas2006} report an overdensity, dubbed Group~6, containing 6 previously unidentified RRLs.
+Based on the distance to the RRLs they claim many of these could be associated to Pal~5's tidal tails.
+The QUEST IDs of the stars in their Group~6 are: 403, 405, 407, 411, 416 and 422.
+The first two stars --403 and 405-- are located at distances of  $9\farcm8$ and $11'$ from the cluster's center, which \citet{Vivas2004} notes places them within the cluster's tidal radius, assuming the value of $r_t\sim16'$ reported by \citet{Odenkirchen:2002}.
+\todo{Update this value}.
+\citet{Vivas2006} also note that star 393, not in Group~6, could also be associated to the cluster's tails, given it's distance and location along the known extent of the tails.
+
+In summary,  previous studies report 7 known RRLs associated to the cluster itself (1 \typeab~and 6 \typec) plus 5 more reported as possibly associated to the tails (4 \typeab~and 2 \typec).
+
+% 
+\begin{table*}[t]
+\caption{RRL members of the Pal 5 stream.}\label{t:rrl_members}
+\begin{footnotesize}
+\begin{tabular}{llllllll}
+\toprule
+\Gaia~DR2               & Other & Period & Amp-$g$ & Type & D & member & sep  \\
+\verb+source_id+        & ID   & (d)    & (mag)   &     &(kpc)& prob & (") \\
+\midrule
+4418920808077110784 &  V5 & 0.337934& 0.46 & RRc  & 20.73 & 0.9982 &   1.9 \\
+4418920846732620032 &     & 0.643239& 0.27 & RRab & 21.71 & 0.9963 &   2.0 \\
+4418914863842345856 &  V1 & 0.293229& 0.56 & RRc  & 20.36 & 0.9696 &   2.0 \\
+4418726027016125056 &  V3 & 0.329948& 0.51 & RRc  & 20.65 & 0.9978 &   3.8 \\
+4418725889577171328 &  V4 & 0.286366& 0.59 & RRc  & 20.28 & 0.9915 &   4.5 \\
+4418731829516302848 &     & 0.28164 & 0.3  & RRc  & 18.14 & 0.9297 &   4.8 \\
+4418913218870688768 &  V2 & 0.332473& 0.4  & RRc  & 20.33 & 0.9977 &   5.0 \\
+4418734165978521728 &     & 0.380859& 0.52 & RRc  & 22.75 & 0.9850 &   7.6 \\
+4418724034151291776 & 403 & 0.551705& 1.14 & RRab & 21.23 & 0.9987 &   9.7 \\
+4418732791593809152 & 405 & 0.329670& 0.45 & RRc  & 20.74 & 0.9959 &  11.0 \\
+4419052204012341760 &     & 0.530006& 1.01 & RRab & 20.84 & 0.9960 &  44.5 \\
+4418142117622280192 & 393 & 0.462671& 1.36 & RRab & 19.95 & 0.9060 & 118.7 \\
+6339498589346000768 &     & 0.320110& 0.46 & RRc  & 18.89 & 0.9934 & 217.7 \\
+6339499379619987200 &     & 0.329577& 0.47 & RRc  & 19.35 & 0.9966 & 218.6 \\
+6339478312804685824 &     & 0.465883& 1.12 & RRab & 19.91 & 0.9970 & 242.8 \\
+4421078432143578496 &     & 0.520749& 1.11 & RRab & 21.98 & 0.9981 & 246.1 \\
+6339398155830632448 &     & 0.688316& 0.75 & RRab & 19.82 & 0.9982 & 258.9 \\
+4427253907921168000 &     & 0.496334& 1.35 & RRab & 21.81 & 0.9905 & 362.5 \\
+4427220338456828416 &     & 0.600796& 0.81 & RRab & 21.98 & 0.9974 & 384.4 \\
+4427234700827469952 &     & 0.517911& 1.24 & RRab & 21.94 & 0.9979 & 392.3 \\
+6337302795905934336 &     & 0.519462& 1.02 & RRab & 17.29 & 0.9808 & 404.5 \\
+6337233350579111680 &     & 0.542383& 0.96 & RRab & 18.98 & 0.9962 & 450.6 \\
+4427646704153765632 &     & 0.385552& 0.37 & RRc  & 19.90 & 0.9429 & 456.5 \\
+4424705647988505600 &     & 0.335502& 0.41 & RRc  & 20.87 & 0.9832 & 513.0 \\
+4426221707021159424 &     & 0.533313& 1.14 & RRab & 21.97 & 0.9728 & 550.1 \\
+\bottomrule
+\end{tabular}
+\end{footnotesize}
+\end{table*}
+
+\subsubsection{Notes on individual stars}
+
+Star 422 is reported as an \rrc~in \citet{Vivas2004}, with a period of 0.61925~d and $V$ amplitude 0.51~mag.
+In PS1 it is classified as an \rrc~with a period of 0.3832276153~d.
+The period reported in \citet{Vivas2004} is an alias of this one, more likely to be the true period given the variability amplitude\footnote{\citet{Vivas2004} also cautions that derived periods for QUEST \rrc~stars are known to be prone to aliasing due to the observing cadence of $\sim1$~d.}.
+Therefore, we keep PS1's classification henceforth for RRL 422.
+
+Star \verb+source_id=6339498589346000768+ is reported both in PS1 and \Gaia~as an \rrc~, with a period 0.4714315854~d in PS1 and 0.320110123108~d in \Gaia~SOS, the two are a pair of 1~d aliases \citep[see e.g.]{Lafler1965}.
+In what follows we adopt the shorter period reported in \Gaia~as the correct one as it is the more probable one for an \rrc~star.
+
+Star \verb+source_id=4418731829516302848+ is reported in PS1 and \Gaia~as an \rrab~with a period around 0.646 and a similarly low amplitude $A_G=0.2$ and $A_g=0.3$ in both surveys.
+This star is reported in CRTS \citep{Drake2014} (ID CSS\_J151623.2-000831) as an \rrc~with a period 0.28164~d.
+We adopt the CRTS period for this star in what follows, as it is based on many more observations ($\gtrsim200$) than \Gaia~and PS1's (about a dozen per filter in each case) and the \rrc~classification and short period are more likely for such a short amplitude star.
+
+\subsection{RR Lyrae population }
+
+Based on the paucity of RRL stars of type~\typeab~in the cluster, Pal~5  would have been expected to be classified as an Oosterhoff type II cluster (ref-XX, no one really seems to say this explicitly), at odds with the cluster's known spectroscopic metallicity of $\FeH=-1.32$ \citep{Kuijsen2018}, as OoII clusters are on average more metal-poor than OoI clusters.
+
+We find 15 RRLs in the cluster tails, $\sim11$ of \type~ab and $\sim4$ of \type~c.
+Combined with the 10~\rrc~in the cluster itself, there are 25 RRLs in total associated to Pal~5 and its tails with membership probabilities larger than 0.5.
+Figure~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
+The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram.
+The distribution of the \typeab~RRL clearly indicate Pal~5 to be an Oosterhoff I cluster, consistently with their mean period of 0.565~d(XX).
+Our findings of the updated ratio of \rrab~to~\rrc, mean period and Period-Amplitude distribution of the \rrab, all support an OoI classification for Pal~5, consistent with the cluster's metallicity.
+
+Out of our newly found \rrab's three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars.
+This type of RRL is only found in relatively metal-rich globular clusters  \citep[$\Feh>-1.5$][]{Monelli2017} and the Sagittarius dwarf galaxy, but not in any other dwarf spheroidal galaxy.
+\todo{CM:( any other point to be made about this?)}.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=\columnwidth]{paper/rrls_PA.pdf}
+\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field.}\label{fig:PA_diagram}
+\end{center}
+\end{figure}
+
+\subsection{Stellar Mass and Luminosity}
+
+%We can provide an estimate on the cluster's luminosity based on the observed number of RRLs. For that, the first step is to assess how complete our observed sample is.
+
+\citet{Sesar2017b} provide an estimate of their catalogue's completeness as a function of distance, at high galactic latitude, based on simulations.
+They estimate the mean completeness to be 92\% for \rrab~and 79\% for \rrc, up to a distance of 40~kpc, well beyond the distance of Pal~5 and its tails.
+At the average ratio of 3 \rrc~stars per every 10 \rrab~\citep{Layden1995}, this represents a mean completeness of 89\%.
+We can also produce an estimate per line of sight using the methodology described in \citet{Rybizki2018}, which requires two independent catalogues to assess the completeness of both in a probabilistic manner.
+Using the \Gaia~DR2 VariClassifier \citet{Holl2018,Rimoldini2018} and Specific Objects Study \citep{Clementini2018} catalogues, we find a median completeness of 92\% with a standard deviation of 16\%  with no evident spatial trend across the field of Figure~\ref{fig:members}.
+
+%we obtain the completeness map shown in Figure~\ref{fig:completeness_map}, for RRLs of type \typeab~and \typec.
+The mean completeness estimated this way is 92\% with a standard deviation of 16\% and, as the figure shows, there is no evident spatial trend.
+The two estimates are remarkably consistent and, in what follows, we will adopt a completeness of 90\%-XX.
+
+To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab RRL, based on data from Galactic dwarf galaxies and globular clusters.
+Here we use it including the data for the \typec~stars.
+Correcting the observed number of 25~RRLs by the median completeness of 92\%, we get a total expected number of 28~RRLs, for which find an absolute magnitude $M_V=-8.1\pm 1.1$.
+
+% \begin{figure}
+% \begin{center}
+% \includegraphics[width=\columnwidth]{paper/ps1_completeness.pdf}
+% \caption{PS1 RRL estimated completeness. The Pal 5 stream's track is shown with the solid line.}\label{fig:completeness_map}
+% \end{center}
+% \end{figure}
+
+\todo{CM: estimate luminosity based on number of RRLs. Compare to estimate from $M_V-\Feh$ relation. Derive stellar mass from stellar luminosity. Compare to dynamical mass estimates.}
+
+- there are $\sim4$-XX times as many RRLs in the tails as there are in the cluster.  
+- how much mass did we already know to be in the tails compared to the cluster?
+(\sa{according to Ibata 2016 the mass in the present day cluster is 4297 $\pm~ 98$ M$_{\odot}$, and the streams excluding the cluster have $\sim$7903 M$_{\odot}$.
+So under a factor of 2 more stars in the tails than in the remnant cluster}.
+The cluter is thought to have had an initial mass of $\sim47000$ M$_{\odot}$).
+Probably not as much as a factor of 4.
+This makes it unexpected to have found this many RRLs in the tails.
+Based on prior knowledge of the mass loss undergone by the cluster we would have expected this to be XX (Sarah?/Adrian?) \sa{Not sure, the lowest mass stars should have been lost first...}.
+Similarly,
+
+\subsection{Implications for stellar population}
+
+\todo{Constraints on initial mass of Pal 5}
+
+\todo{CM:} Inferred luminosity based on the number of RRLs is XX. Compare to inferred luminosity based on the $M_V-\FeH$ relation.
+
+\subsection{Cluster distance and stream track}
 
@@
 \bibliographystyle{aasjournal}
-\bibliography{refs}
+\bibliography{refs,cm1,cm2}
 

### 2019-08-21 paper/main.tex

@@
 However, at its present distance, main sequence stars are barely detected and have large astrometric uncertainties in \Gaia\ \DR{2}, and other stellar tracers (e.g., giant branch stars) are not numerous enough or precise enough distance indicators to provide useful relative distance information along the stream.
-Enter RR Lyrae-type stars.
+% Enter RR Lyrae-type stars.
 
-\todo{Cecilia: write this paragraph? you probably have some canned text?}
-RR Lyrae stars (RRLs) are pulsating horizontal branch stars that are intrinsically more luminous than main sequence stars, but about $\sim 10,000$ time less numerous. 
-HB = good distance indicator.
-The horizontal branch in Pal 5 is bright enough to be observed by Gaia, meaning we can measure kinematic properties of the full stream...
+\todo{Cecilia: write this paragraph? you probably have some canned text? - (In progress, CM.}
+RR Lyrae stars (RRLs) are pulsating horizontal branch giants intrinsically more luminous than main sequence stars, but about $\sim 10,000$ time less numerous.
+They can be reliably identified based on their photometric variability alone, with little to no contamination from other types of variable stars, and are well known standard candles.
+Their distances can be estimated based on photometric data alone by means of their period-luminosity relation, with distance errors that go from $\sim8\%$ in the optical to as low as 3\% in the infrared \citep{Neely2017,Sesar2017a}.
+At its current distance, Pal~5's horizontal branch is bright enough $\G\sim17.3$ to be observed by \Gaia, so it's full extent can be traced and kinematic properties measured with RRLs.
+%meaning we can measure kinematic properties of the full stream...
 \citep{Ibata:2017} found some RRL in pal 5, but no proper motions.
@@
 \includegraphics[width=\textwidth]{paper/tracks.pdf}
-\caption{todo}
+\caption{RRL stars in the vicinity of the Pal 5 stream (with $\textrm{prob.} > 0.01$), colored by membership probability, and under-plotted with the inferred stream trends used in the membership model.
+\textbf{Upper left}: Sky positions (in the Pal 5 stream coordinate frame) of RRLs, with arrows showing the \Gaia\ \DR{2} proper motion vector direction.
+\textbf{Upper right}: Distance as a function of Pal 5 longitude, $\phi_1$, for the RRLs.
+The shaded (orange) band shows the 16--84th percentile confidence region for the inferred distance trend of the stream, and the solid line shows the median posterior sample, all assuming a quadratic function of $\phi_1$.
+\textbf{Lower panels}: Same as upper right, but for the proper motion in right ascension, $\mu_\alpha$, and declination, $\mu_\delta$.
+}
 \label{fig:trackmembers}
@@
 
+\subsection{RR Lyrae stars associated with Pal 5 and its tails}
+
+\todo{How many we find, what their types are}
+
+\subsection{Kinematics of the stream and cluster}
+
+\todo{Summarize mean dist and pm from RRLs, and trend parameters}
+
+\todo{Make a table with summary information}
+
 \subsection{RR Lyrae stars known to be associated with Pal 5 or it's tails}
+
+\todo{move some of this to discussion?}
 

### 2019-08-21 paper/main.tex
whitespace and fix some errors

@@
 
-\input{paper/defs.tex}
+\input{defs.tex}
+\graphicspath{paper/}
 
@@
 \begin{center}
-\includegraphics[width=\textwidth]{paper/proper-motion.pdf}
+\includegraphics[width=\textwidth]{proper-motion.pdf}
 \caption{\textbf{Left}: Proper motions of RRLs in the region $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$ with distances in the range $18 < d < 25~\kpc$.
@@
 \begin{center}
-\includegraphics[width=\textwidth]{paper/tracks.pdf}
+\includegraphics[width=\textwidth]{tracks.pdf}
 \caption{RRL stars in the vicinity of the Pal 5 stream (with $\textrm{prob.} > 0.01$), colored by membership probability, and under-plotted with the inferred stream trends used in the membership model.
@@
 \begin{center}
-\includegraphics[width=\textwidth]{paper/members.pdf}
+\includegraphics[width=\textwidth]{members.pdf}
 \caption{Sky positions of RRL...todo}
@@
 
-We find 15 RRLs in the cluster tails, $\sim11$ of \type~ab and $\sim4$ of \type~c.
+We find 15 RRLs in the cluster tails, $\sim11$ of \typeab\ and $\sim4$ of \typec.
 Combined with the 10~\rrc~in the cluster itself, there are 25 RRLs in total associated to Pal~5 and its tails with membership probabilities larger than 0.5.
@@
 Out of our newly found \rrab's three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars.
-This type of RRL is only found in relatively metal-rich globular clusters  \citep[$\Feh>-1.5$][]{Monelli2017} and the Sagittarius dwarf galaxy, but not in any other dwarf spheroidal galaxy.
+This type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$][]{Monelli2017} and the Sagittarius dwarf galaxy, but not in any other dwarf spheroidal galaxy.
 \todo{CM:( any other point to be made about this?)}.
@@
 \begin{center}
-\includegraphics[width=\columnwidth]{paper/rrls_PA.pdf}
+\includegraphics[width=\columnwidth]{rrls_PA.pdf}
 \caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field.}\label{fig:PA_diagram}
@@
 
-To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab RRL, based on data from Galactic dwarf galaxies and globular clusters.
+To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters.
 Here we use it including the data for the \typec~stars.
@@
 % \begin{center}
-% \includegraphics[width=\columnwidth]{paper/ps1_completeness.pdf}
+% \includegraphics[width=\columnwidth]{ps1_completeness.pdf}
 % \caption{PS1 RRL estimated completeness. The Pal 5 stream's track is shown with the solid line.}\label{fig:completeness_map}
@@
 
-\todo{CM: estimate luminosity based on number of RRLs. Compare to estimate from $M_V-\Feh$ relation. Derive stellar mass from stellar luminosity. Compare to dynamical mass estimates.}
+\todo{CM: estimate luminosity based on number of RRLs. Compare to estimate from $M_V-\FeH$ relation. Derive stellar mass from stellar luminosity. Compare to dynamical mass estimates.}
 

### 2019-08-21 paper/main.tex
bibs

@@
 
-\keywords{globular clusters: individual: Palomar 5 --
-Galaxy: halo --
-stars: variables: RR Lyrae --
+\keywords{globular clusters: individual: Palomar 5 ---
+Galaxy: halo ---
+stars: variables: RR Lyrae ---
 Galaxy: structure}
@@
 Globular clusters are destroyed as they orbit within the Milky Way.
-The $\sim$150 bound globular clusters we presently see throughout the Galaxy \citep{Harris2010} are therefore thought to be surviving relics of a much larger initial population, most of which were destroyed (primarily) through a combination of relaxation / evaporation and gravitational shocking from the disk and bulge \citep[e.g.,][]{Gnedin:1997, Gnedin:2014}.
+The $\sim$150 bound globular clusters we presently see throughout the Galaxy \citep{Harris:2010} are therefore thought to be surviving relics of a much larger initial population, most of which were destroyed (primarily) through a combination of relaxation / evaporation and gravitational shocking from the disk and bulge \citep[e.g.,][]{Gnedin:1997, Gnedin:2014}.
 As stellar systems are destroyed---i.e., as stars are tidally stripped from their progenitor---the tidal debris forms tails of matter that both lead and trail the remnant cluster, producing a stellar stream that may persist even after the progenitor is fully destroyed \citep[e.g.,][]{Johnston:1996, Eyre:2011}.
@@
 However, a number of peculiar aspects of the stream still remain unexplained in detail.
-For example, the leading and trailing tails have significantly different total star counts and lengths \citep{Dehnen2004, Bernard:2016}, and the stream track, width, and density vary significantly over the full extent of the stream \citep{Ibata:2016, Bonaca:2019}.
+For example, the leading and trailing tails have significantly different total star counts and lengths \citep{Dehnen:2004, Bernard:2016}, and the stream track, width, and density vary significantly over the full extent of the stream \citep{Ibata:2016, Bonaca:2019}.
 Given that Pal 5's pericenter is well within the Galactic disk ($r_{\textrm{peri}} \sim 7$--$8~\kpc$), these features could be a sign of perturbations from molecular clouds \citep[e.g.,][]{Amorisco:2016}, interaction with the Galactic bar \citep[e.g.,][]{Pearson:2017}, signatures of past dark matter subhalo encounters \citep[e.g.,][]{Erkal:2017}, or some combination of all of these effects.
@@
 They can be reliably identified based on their photometric variability alone, with little to no contamination from other types of variable stars, and are well known standard candles.
-Their distances can be estimated based on photometric data alone by means of their period-luminosity relation, with distance errors that go from $\sim8\%$ in the optical to as low as 3\% in the infrared \citep{Neely2017,Sesar2017a}.
+Their distances can be estimated based on photometric data alone by means of their period-luminosity relation, with distance errors that go from $\sim8\%$ in the optical to as low as 3\% in the infrared \citep{Neeley:2017, Sesar:2017a}.
 At its current distance, Pal~5's horizontal branch is bright enough $\G\sim17.3$ to be observed by \Gaia, so it's full extent can be traced and kinematic properties measured with RRLs.
@@
 
-To search for RRLs in the vicinity of Pal 5, we use a catalog of RRL stars from the PS1 survey that spans $\sim$3/4 of the sky (all sky north of declination $\delta > -30\degr$) and contains 61\,795 RRLs identified from their variability \citep{Sesar2017b}.
+To search for RRLs in the vicinity of Pal 5, we use a catalog of RRL stars from the PS1 survey that spans $\sim$3/4 of the sky (all sky north of declination $\delta > -30\degr$) and contains 61\,795 RRLs identified from their variability \citep{Sesar:2017b}.
 In detail, the \emph{bona fide} RRLs in the catalog were classified using multi-epoch data in the $grizy$ photometric bands, with a typical number of 12 observations per filter, and has a limiting magnitude $r\sim21$ (corresponding to $G~21$ for the typical colors of RRLs).
@@
 
-\citet{Sesar2017b} provide distances to the RRLs based on an $i$-band Period-Luminosity-Metallicity (PLZ) relation.
+\citet{Sesar:2017b} provide distances to the RRLs based on an $i$-band Period-Luminosity-Metallicity (PLZ) relation.
 This has two key advantages when compared to optical bands: The PLZ relation in the $i$-band only weakly depends on metallicity, and the impact of dust extinction is less severe at longer wavelengths.
-However, distances reported for the \rrc~stars in Table~5 of \citet{Sesar2017b} have a systematic offset with respect to the \typeab~stars.
+However, distances reported for the \rrc~stars in Table~5 of \citet{Sesar:2017b} have a systematic offset with respect to the \typeab~stars.
 This offset comes from having applied the same PLZ to \typeab~and \typec~stars.
@@
 In what follows, we use the $i$-band PLZ distances for the RRLs.
-\citet{Sesar2017b} estimate their overall distance precision to be 3\%, of which $\sim2$\% and $\sim1$\% correspond to systematic and random uncertainties, respectively.
+\citet{Sesar:2017b} estimate their overall distance precision to be 3\%, of which $\sim2$\% and $\sim1$\% correspond to systematic and random uncertainties, respectively.
 
@@
 
-Based on the paucity of RRL stars of type~\typeab~in the cluster, Pal~5  would have been expected to be classified as an Oosterhoff type II cluster (ref-XX, no one really seems to say this explicitly), at odds with the cluster's known spectroscopic metallicity of $\FeH=-1.32$ \citep{Kuijsen2018}, as OoII clusters are on average more metal-poor than OoI clusters.
+Based on the paucity of RRL stars of type~\typeab~in the cluster, Pal~5  would have been expected to be classified as an Oosterhoff type II cluster (ref-XX, no one really seems to say this explicitly), at odds with the cluster's known spectroscopic metallicity of $\FeH=-1.32$ \citep{TODO}, as OoII clusters are on average more metal-poor than OoI clusters.
 
@@
 Out of our newly found \rrab's three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars.
-This type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$][]{Monelli2017} and the Sagittarius dwarf galaxy, but not in any other dwarf spheroidal galaxy.
+This type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$][]{Monelli:2017} and the Sagittarius dwarf galaxy, but not in any other dwarf spheroidal galaxy.
 \todo{CM:( any other point to be made about this?)}.
@@
 
-To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters.
+To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu:2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters.
 Here we use it including the data for the \typec~stars.
@@
 So under a factor of 2 more stars in the tails than in the remnant cluster}.
-The cluter is thought to have had an initial mass of $\sim47000$ M$_{\odot}$).
+The cluster is thought to have had an initial mass of $\sim47000$ M$_{\odot}$).
 Probably not as much as a factor of 4.
@@
 \bibliographystyle{aasjournal}
-\bibliography{refs,cm1,cm2}
+\bibliography{refs}
 

### 2019-08-21 paper/main.tex
shuffling

@@
 \begin{abstract}
-\todo{This is just placeholder text!}
 % Context
-Stellar streams are important.
-The Palomar 5 stellar stream is a prototypical example: It connects back to the progenitor cluster.
-RR Lyrae-type (RRL) stars are useful because they are standard candles and their distances can be inferred with small errors ($\sim7\%$)
+Thin stellar streams, formed from the tidal disruption of globular clusters, are important gravitational tools, sensitive to both global and small-scale properties of dark matter.
+The Palomar 5 stellar stream (Pal 5) is an exemplar stream within the Milky Way: Its $\approx 20\degr$ stream connects back to the progenitor cluster, and it has been used to study the shape and total mass of, along with substructure in, the dark matter distribution of the Galaxy.
+However, ...most details of the stream are not fully explained.
 % Aims
-While the Pal 5 cluster is known to host a small number (5--7) of RRL stars, only two RRL stars have since been found in the long tidal tails that form the Pal 5 stream.
+...we want to measure distance and kinematic tracks along the stream to inform models.
+Here we use a large catalog of RR Lyrae-type stars (RRLs) with astrometric data from the \Gaia\ mission to probabilistically identify RRLs in the Pal 5 stream.
+RRLs are useful because they are intrinsically-luminous standard candles and their distances can be inferred with small relative precision ($\sim3\%$).
+% While the Pal 5 cluster is known to host a small number (5--7) of RRL stars, only two RRL stars have since been found in the long tidal tails that form the Pal 5 stream.
 % Methods
-Here we use a large catalog of RRL stars released as a part of Gaia DR2 to search for RRL stars in the Pal 5 stream.
+% Results
 By building a probabilistic model of the cluster and stream in proper motion and distance, we find XX RRL consistent with being members of the stream.
-% Results
 The number of Pal 5 RRL is therefore XX times larger in the tails than in the remnant cluster: This changes our interpretation of ...blah.
@@
 To search for RRLs in the vicinity of Pal 5, we use a catalog of RRL stars from the PS1 survey that spans $\sim$3/4 of the sky (all sky north of declination $\delta > -30\degr$) and contains 61\,795 RRLs identified from their variability \citep{Sesar:2017b}.
-In detail, the \emph{bona fide} RRLs in the catalog were classified using multi-epoch data in the $grizy$ photometric bands, with a typical number of 12 observations per filter, and has a limiting magnitude $r\sim21$ (corresponding to $G~21$ for the typical colors of RRLs).
+In detail, the \emph{bona fide} RRLs in the catalog were classified using multi-epoch data in the $grizy$ photometric bands, with a typical number of 12 observations per filter, and has a limiting magnitude $r\sim21$ (corresponding to $G \sim 21$ for the typical colors of RRLs).
 Periods and amplitudes in all five bands are available for all stars in this catalog.
@@
 Nevertheless, the vast majority (61\,625) of PS1's \emph{bona fide} RRLs have astrometric information available in \Gaia~\DR{2}.
+
+\begin{figure*}[th]
+\begin{center}
+\includegraphics[width=\textwidth]{proper-motion.pdf}
+\caption{\textbf{Left}: Proper motions of RRLs in the region $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$ with distances in the range $18 < d < 25~\kpc$.
+\textbf{Middle}: The same, but for RRLs within $1~\degr$ of the mean sky track of Pal 5 \citep{Bonaca:2019}.
+The over-density of stars near $(-2.7, -2.7)$ is the stream, and the two colored points show the Pal 5 cluster proper motions from \citet{Vasiliev:2019} and \citet{Fritz:2015}, respectively.
+\textbf{Right}: The same, but for RRLs excluding $2~\degr$ around the Pal 5 sky track.}
+\label{fig:pm}
+\end{center}
+\end{figure*}
 
@@
 
-\begin{figure*}[th]
-\begin{center}
-\includegraphics[width=\textwidth]{proper-motion.pdf}
-\caption{\textbf{Left}: Proper motions of RRLs in the region $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$ with distances in the range $18 < d < 25~\kpc$.
-\textbf{Middle}: The same, but for RRLs within $1~\degr$ of the mean sky track of Pal 5 \citep{Bonaca:2019}.
-The over-density of stars near $(-2.7, -2.7)$ is the stream, and the two colored points show the Pal 5 cluster proper motions from \citet{Vasiliev:2019} and \citet{Fritz:2015}, respectively.
-\textbf{Right}: The same, but for RRLs excluding $2~\degr$ around the Pal 5 sky track.}
-\label{fig:pm}
-\end{center}
-\end{figure*}
-
 \section{Determining stream membership} \label{sec:membership}
@@
 
+\begin{figure*}[t]
+\begin{center}
+\includegraphics[width=\textwidth]{tracks.pdf}
+\caption{RRL stars in the vicinity of the Pal 5 stream (with $\textrm{prob.} > 0.01$), colored by membership probability, and under-plotted with the inferred stream trends used in the membership model.
+\textbf{Upper left}: Sky positions (in the Pal 5 stream coordinate frame) of RRLs, with arrows showing the \Gaia\ \DR{2} proper motion vector direction.
+\textbf{Upper right}: Distance as a function of Pal 5 longitude, $\phi_1$, for the RRLs.
+The shaded (orange) band shows the 16--84th percentile confidence region for the inferred distance trend of the stream, and the solid line shows the median posterior sample, all assuming a quadratic function of $\phi_1$.
+\textbf{Lower panels}: Same as upper right, but for the proper motion in right ascension, $\mu_\alpha$, and declination, $\mu_\delta$.
+}
+\label{fig:trackmembers}
+\end{center}
+\end{figure*}
+
 To represent the Pal 5 stream, we use a single Gaussian component with a fixed dispersion of $0.05~\masyr$ for the proper motion components ($\approx 5~\kms$ at the distance of Pal 5) and $0.2~\kpc$ in distance.
 However, we allow the mean to vary independently in each component as a function of $\phi_1$, a longitude coordinate (along the stream) as defined in \citet{Bonaca:2019}.
-Based on the Pal 5 stream models from \citet{Bonaca:2019}, over the range of $\phi_1 \in (-20^\circ, 15^\circ$, the mean stream trends in $\mu_\alpha$, $\mu_\delta$, and distance can be well-approximated (within our observational uncertainties) by a quadratic function in $\phi_1$.
+Based on the Pal 5 stream models from \citet{Bonaca:2019}, over the range of $\phi_1 \in (-20^\circ, 15^\circ)$, the mean stream trends in $\mu_\alpha$, $\mu_\delta$, and distance can be well-approximated (within our observational uncertainties) by a quadratic function in $\phi_1$.
 The mean of our stream component, $\bs{x}$, is therefore given by
@@
 
+\begin{figure*}[t]
+\begin{center}
+\includegraphics[width=\textwidth]{members.pdf}
+\caption{Sky positions of RRL...todo}
+\label{fig:members}
+\end{center}
+\end{figure*}
+
 We use Markov Chain Monte Carlo (MCMC) to generate samples over the parameters $(\bs{\theta}, f)$ using the posterior probability
@@
 
+\section{Results} \label{sec:results}
+
+\subsection{RR Lyrae stars associated with Pal 5 and its tails}
+
+We find a total of 24 RR Lyrae stars with membership probability $> 0.5$ that lie within $1\degr$ of the stream track \citep[again using the track from][]{Bonaca:2019}.
+Of those, 13 are classified as type \typeab\ and 11 are classified as type \typec\ RRLs.
+We find a total of 10 RRLs within the estimated Jacobi radius of the cluster ($\approx 12'$), meaning that 15 RRLs are spread between the leading and trailing tails.
+\figurename~\ref{fig:members} shows the sky positions of all RRL with membership probability $>0.01$ over-plotted on the density of filtered main sequence stars from \citet{Bonaca:2019}.
+We note that there are a few RRLs outside of the main stream track that appear to be at a consistent distance with consistent proper motions (e.g., the two stars near $230\degr, 0\degr$) that are not included in the tally above.
+No theoretical models of Pal 5 have predicted tidal debris at such locations, so follow-up spectroscopy to confirm or rule out their origin would be very informative.
+
+\subsection{Kinematics of the stream and cluster}
+
+We find a mean Pal 5 cluster distance of $20.9 \pm 0.2~\kpc$ using the PS1 RRLs and the membership model described in \sectionname~\ref{sec:membership}.
+This is significantly closer than past measurements of the cluster distance that have used \todo{summarize past distance measurements and why this could be so different...}.
+We also detect a distance trend from the closer leading tail (at $\approx 19~\kpc$) to the more distant trailing tail (at $\approx 22~\kpc$).
+While the proper motion in right ascension appears consistent with constant or only a mild gradient over the extent of the stream, we see a significant gradient in the declination component, with a difference of about $\approx 1~\masyr$ from trailing to leading tail.
+% \todo{Make a table with summary information?}
+
+\subsection{RR Lyrae population }
+
+Based on the paucity of RRL stars of type~\typeab~in the cluster, Pal~5  would have been expected to be classified as an Oosterhoff type II cluster (ref-XX, no one really seems to say this explicitly), at odds with the cluster's known spectroscopic metallicity of $\FeH=-1.32$ \citep{TODO}, as OoII clusters are on average more metal-poor than OoI clusters.
+
+We find 15 RRLs in the cluster tails, $\sim11$ of \typeab\ and $\sim4$ of \typec.
+Combined with the 10~\rrc~in the cluster itself, there are 25 RRLs in total associated to Pal~5 and its tails with membership probabilities larger than 0.5.
+Figure~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
+The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram.
+The distribution of the \typeab~RRL clearly indicate Pal~5 to be an Oosterhoff I cluster, consistently with their mean period of 0.565~d(XX).
+Our findings of the updated ratio of \rrab~to~\rrc, mean period and Period-Amplitude distribution of the \rrab, all support an OoI classification for Pal~5, consistent with the cluster's metallicity.
+
+Out of our newly found \rrab's three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars.
+This type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$][]{Monelli:2017} and the Sagittarius dwarf galaxy, but not in any other dwarf spheroidal galaxy.
+\todo{CM:( any other point to be made about this?)}.
+
 \begin{figure*}[t]
 \begin{center}
-\includegraphics[width=\textwidth]{tracks.pdf}
-\caption{RRL stars in the vicinity of the Pal 5 stream (with $\textrm{prob.} > 0.01$), colored by membership probability, and under-plotted with the inferred stream trends used in the membership model.
-\textbf{Upper left}: Sky positions (in the Pal 5 stream coordinate frame) of RRLs, with arrows showing the \Gaia\ \DR{2} proper motion vector direction.
-\textbf{Upper right}: Distance as a function of Pal 5 longitude, $\phi_1$, for the RRLs.
-The shaded (orange) band shows the 16--84th percentile confidence region for the inferred distance trend of the stream, and the solid line shows the median posterior sample, all assuming a quadratic function of $\phi_1$.
-\textbf{Lower panels}: Same as upper right, but for the proper motion in right ascension, $\mu_\alpha$, and declination, $\mu_\delta$.
-}
-\label{fig:trackmembers}
+\includegraphics[width=\textwidth]{rrls_PA.pdf}
+\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field.}
+\label{fig:PA_diagram}
 \end{center}
@@
 
-\section{Results} \label{sec:results}
+\subsection{Stellar Mass and Luminosity}
 
-\begin{figure*}[t]
-\begin{center}
-\includegraphics[width=\textwidth]{members.pdf}
-\caption{Sky positions of RRL...todo}
-\label{fig:members}
-\end{center}
-\end{figure*}
+%We can provide an estimate on the cluster's luminosity based on the observed number of RRLs. For that, the first step is to assess how complete our observed sample is.
 
-\subsection{RR Lyrae stars associated with Pal 5 and its tails}
+\citet{Sesar:2017b} provide an estimate of their catalogue's completeness as a function of distance, at high galactic latitude, based on simulations.
+They estimate the mean completeness to be 92\% for \rrab~and 79\% for \rrc, up to a distance of 40~kpc, well beyond the distance of Pal~5 and its tails.
+At the average ratio of 3 \rrc~stars per every 10 \rrab~\citep{Layden1995}, this represents a mean completeness of 89\%.
+We can also produce an estimate per line of sight using the methodology described in \citet{Rybizki2018}, which requires two independent catalogues to assess the completeness of both in a probabilistic manner.
+Using the \Gaia~DR2 VariClassifier \citet{Holl2018,Rimoldini2018} and Specific Objects Study \citep{Clementini2018} catalogues, we find a median completeness of 92\% with a standard deviation of 16\%  with no evident spatial trend across the field of Figure~\ref{fig:members}.
 
-\todo{How many we find, what their types are}
+%we obtain the completeness map shown in Figure~\ref{fig:completeness_map}, for RRLs of type \typeab~and \typec.
+The mean completeness estimated this way is 92\% with a standard deviation of 16\% and, as the figure shows, there is no evident spatial trend.
+The two estimates are remarkably consistent and, in what follows, we will adopt a completeness of 90\%-XX.
 
-\subsection{Kinematics of the stream and cluster}
+To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu:2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters.
+Here we use it including the data for the \typec~stars.
+Correcting the observed number of 25~RRLs by the median completeness of 92\%, we get a total expected number of 28~RRLs, for which find an absolute magnitude $M_V=-8.1\pm 1.1$.
 
-\todo{Summarize mean dist and pm from RRLs, and trend parameters}
+% \begin{figure}
+% \begin{center}
+% \includegraphics[width=\columnwidth]{ps1_completeness.pdf}
+% \caption{PS1 RRL estimated completeness. The Pal 5 stream's track is shown with the solid line.}\label{fig:completeness_map}
+% \end{center}
+% \end{figure}
 
-\todo{Make a table with summary information}
+\todo{CM: estimate luminosity based on number of RRLs. Compare to estimate from $M_V-\FeH$ relation. Derive stellar mass from stellar luminosity. Compare to dynamical mass estimates.}
+
+- there are $\sim4$-XX times as many RRLs in the tails as there are in the cluster.
+- how much mass did we already know to be in the tails compared to the cluster?
+(\sa{according to Ibata 2016 the mass in the present day cluster is 4297 $\pm~ 98$ M$_{\odot}$, and the streams excluding the cluster have $\sim$7903 M$_{\odot}$.
+So under a factor of 2 more stars in the tails than in the remnant cluster}.
+The cluster is thought to have had an initial mass of $\sim47000$ M$_{\odot}$).
+Probably not as much as a factor of 4.
+This makes it unexpected to have found this many RRLs in the tails.
+Based on prior knowledge of the mass loss undergone by the cluster we would have expected this to be XX (Sarah?/Adrian?) \sa{Not sure, the lowest mass stars should have been lost first...}.
+Similarly,
+
+\section{Discussion}
 
@@
 
-\subsection{RR Lyrae population }
+\section{Conclusions} \label{sec:conclusions}
 
-Based on the paucity of RRL stars of type~\typeab~in the cluster, Pal~5  would have been expected to be classified as an Oosterhoff type II cluster (ref-XX, no one really seems to say this explicitly), at odds with the cluster's known spectroscopic metallicity of $\FeH=-1.32$ \citep{TODO}, as OoII clusters are on average more metal-poor than OoI clusters.
-
-We find 15 RRLs in the cluster tails, $\sim11$ of \typeab\ and $\sim4$ of \typec.
-Combined with the 10~\rrc~in the cluster itself, there are 25 RRLs in total associated to Pal~5 and its tails with membership probabilities larger than 0.5.
-Figure~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
-The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram.
-The distribution of the \typeab~RRL clearly indicate Pal~5 to be an Oosterhoff I cluster, consistently with their mean period of 0.565~d(XX).
-Our findings of the updated ratio of \rrab~to~\rrc, mean period and Period-Amplitude distribution of the \rrab, all support an OoI classification for Pal~5, consistent with the cluster's metallicity.
-
-Out of our newly found \rrab's three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars.
-This type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$][]{Monelli:2017} and the Sagittarius dwarf galaxy, but not in any other dwarf spheroidal galaxy.
-\todo{CM:( any other point to be made about this?)}.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=\columnwidth]{rrls_PA.pdf}
-\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field.}\label{fig:PA_diagram}
-\end{center}
-\end{figure}
-
-\subsection{Stellar Mass and Luminosity}
-
-%We can provide an estimate on the cluster's luminosity based on the observed number of RRLs. For that, the first step is to assess how complete our observed sample is.
-
-\citet{Sesar2017b} provide an estimate of their catalogue's completeness as a function of distance, at high galactic latitude, based on simulations.
-They estimate the mean completeness to be 92\% for \rrab~and 79\% for \rrc, up to a distance of 40~kpc, well beyond the distance of Pal~5 and its tails.
-At the average ratio of 3 \rrc~stars per every 10 \rrab~\citep{Layden1995}, this represents a mean completeness of 89\%.
-We can also produce an estimate per line of sight using the methodology described in \citet{Rybizki2018}, which requires two independent catalogues to assess the completeness of both in a probabilistic manner.
-Using the \Gaia~DR2 VariClassifier \citet{Holl2018,Rimoldini2018} and Specific Objects Study \citep{Clementini2018} catalogues, we find a median completeness of 92\% with a standard deviation of 16\%  with no evident spatial trend across the field of Figure~\ref{fig:members}.
-
-%we obtain the completeness map shown in Figure~\ref{fig:completeness_map}, for RRLs of type \typeab~and \typec.
-The mean completeness estimated this way is 92\% with a standard deviation of 16\% and, as the figure shows, there is no evident spatial trend.
-The two estimates are remarkably consistent and, in what follows, we will adopt a completeness of 90\%-XX.
-
-To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu:2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters.
-Here we use it including the data for the \typec~stars.
-Correcting the observed number of 25~RRLs by the median completeness of 92\%, we get a total expected number of 28~RRLs, for which find an absolute magnitude $M_V=-8.1\pm 1.1$.
-
-% \begin{figure}
-% \begin{center}
-% \includegraphics[width=\columnwidth]{ps1_completeness.pdf}
-% \caption{PS1 RRL estimated completeness. The Pal 5 stream's track is shown with the solid line.}\label{fig:completeness_map}
-% \end{center}
-% \end{figure}
-
-\todo{CM: estimate luminosity based on number of RRLs. Compare to estimate from $M_V-\FeH$ relation. Derive stellar mass from stellar luminosity. Compare to dynamical mass estimates.}
-
-- there are $\sim4$-XX times as many RRLs in the tails as there are in the cluster.
-- how much mass did we already know to be in the tails compared to the cluster?
-(\sa{according to Ibata 2016 the mass in the present day cluster is 4297 $\pm~ 98$ M$_{\odot}$, and the streams excluding the cluster have $\sim$7903 M$_{\odot}$.
-So under a factor of 2 more stars in the tails than in the remnant cluster}.
-The cluster is thought to have had an initial mass of $\sim47000$ M$_{\odot}$).
-Probably not as much as a factor of 4.
-This makes it unexpected to have found this many RRLs in the tails.
-Based on prior knowledge of the mass loss undergone by the cluster we would have expected this to be XX (Sarah?/Adrian?) \sa{Not sure, the lowest mass stars should have been lost first...}.
-Similarly,
-
-\subsection{Implications for stellar population}
-
-\todo{Constraints on initial mass of Pal 5}
-
-\todo{CM:} Inferred luminosity based on the number of RRLs is XX. Compare to inferred luminosity based on the $M_V-\FeH$ relation.
-
-\subsection{Cluster distance and stream track}
+\todo{what did we do}
 

### 2019-08-21 paper/main.tex
Updates from Overleaf

@@
 Thin stellar streams, formed from the tidal disruption of globular clusters, are important gravitational tools, sensitive to both global and small-scale properties of dark matter.
-The Palomar 5 stellar stream (Pal 5) is an exemplar stream within the Milky Way: Its $\approx 20\degr$ stream connects back to the progenitor cluster, and it has been used to study the shape and total mass of, along with substructure in, the dark matter distribution of the Galaxy.
-However, ...most details of the stream are not fully explained.
+The Palomar 5 stellar stream (Pal 5) is an exemplar stream within the Milky Way: Its $\approx 20\degr$ tidal tails connect back to the progenitor cluster, and it has been used to study the shape and total mass of, along with substructure in, the dark matter distribution of the Galaxy.
+However, most details of the phase-space distribution of the stream are not fully explained, and dynamical models that use the stream for other inferences are therefore incomplete.
 % Aims
-...we want to measure distance and kinematic tracks along the stream to inform models.
-Here we use a large catalog of RR Lyrae-type stars (RRLs) with astrometric data from the \Gaia\ mission to probabilistically identify RRLs in the Pal 5 stream.
+Here we aim to measure distance and kinematic properties along the Pal 5 stream in order to motivate improved models of the Pal 5 stream.
+We use a large catalog of RR Lyrae-type stars (RRLs) with astrometric data from the \Gaia\ mission to probabilistically identify RRLs in the Pal 5 stream.
 RRLs are useful because they are intrinsically-luminous standard candles and their distances can be inferred with small relative precision ($\sim3\%$).
-% While the Pal 5 cluster is known to host a small number (5--7) of RRL stars, only two RRL stars have since been found in the long tidal tails that form the Pal 5 stream.
 % Methods
 % Results
-By building a probabilistic model of the cluster and stream in proper motion and distance, we find XX RRL consistent with being members of the stream.
-The number of Pal 5 RRL is therefore XX times larger in the tails than in the remnant cluster: This changes our interpretation of ...blah.
+By building a probabilistic model of the Pal 5 cluster and stream in proper motion and distance, we find 25 RRL consistent with being members of the cluster (10) and stream (15).
+Using these RRL, we detect gradients in distance and proper motion along the stream, and provide an updated measurement of the distance to the Pal 5 cluster using the RRL.
 % Conclusions
-Stuff.
+We provide a catalog of Pal 5 RRLs with inferred membership probabilities for future modeling work.
 \end{abstract}
@@
 The $\sim$150 bound globular clusters we presently see throughout the Galaxy \citep{Harris:2010} are therefore thought to be surviving relics of a much larger initial population, most of which were destroyed (primarily) through a combination of relaxation / evaporation and gravitational shocking from the disk and bulge \citep[e.g.,][]{Gnedin:1997, Gnedin:2014}.
-As stellar systems are destroyed---i.e., as stars are tidally stripped from their progenitor---the tidal debris forms tails of matter that both lead and trail the remnant cluster, producing a stellar stream that may persist even after the progenitor is fully destroyed \citep[e.g.,][]{Johnston:1996, Eyre:2011}.
+As stellar systems are destroyed---i.e., as stars are tidally stripped from their progenitor---the tidal debris forms tails of matter that both lead and trail the remnant cluster, producing a stellar stream that may persist even after the progenitor is fully destroyed \citep[e.g.,][]{Johnston:1996}.
 
-Stellar streams are useful objects because they almost\footnote{More precisely, the phase-space distribution of stars along a stream encodes the orbital history of the progenitor system \citep[e.g.,][]{Sanders:2013}.} delineate the past and future trajectory of their progenitor system, offering information about orbits that can be used to infer the mass distribution from the single kinematic snapshot of the Galaxy that we are afforded \citep[e.g.,][]{Johnston:1999, Binney:2008, Eyre:2009, Sanders:2013, PriceWhelan:2014, Bonaca:2018}.
+Stellar streams are useful objects because they almost\footnote{More precisely, the phase-space distribution of stars along a stream encodes the orbital history of the progenitor system \citep[e.g.,][]{Sanders:2013}.} delineate the past and future trajectory of their progenitor system, offering information about orbits that can be used to infer the mass distribution from the single kinematic snapshot of the Galaxy that we are afforded \citep[e.g.,][]{Johnston:1999, Sanders:2013, PriceWhelan:2014, Bonaca:2018}.
+% Binney:2008, Eyre:2009,
 Many globular cluster stellar streams have been discovered over the last few decades \citep[see, e.g.,][]{Grillmair:2016, Shipp:2018}, in large part because of large-area, deep, multi-band imaging surveys such as the Sloan Digital Sky Survey \citep[SDSS;][]{York:2000}, Pan-STARRS PS1 survey \citep{Chambers:2016}, and Dark Energy Survey \citep{DES:2016}.

### 2019-08-21 paper/main.tex
Updates from Overleaf

@@
 
-\author{Sarah~Pearson}
+\author[0000-0003-0256-5446]{Sarah~Pearson}
 \affiliation{Center for Computational Astrophysics, Flatiron Institute,
@@
 Thin stellar streams, formed from the tidal disruption of globular clusters, are important gravitational tools, sensitive to both global and small-scale properties of dark matter.
-The Palomar 5 stellar stream (Pal 5) is an exemplar stream within the Milky Way: Its $\approx 20\degr$ tidal tails connect back to the progenitor cluster, and it has been used to study the shape and total mass of, along with substructure in, the dark matter distribution of the Galaxy.
+The Palomar 5 stellar stream (Pal 5) is an exemplar stream within the Milky Way: Its $\approx 20\degr$ tidal tails connect back to the progenitor cluster, and the stream has been used to study the shape and total mass of, along with substructure in, the dark matter distribution of the Galaxy.
 However, most details of the phase-space distribution of the stream are not fully explained, and dynamical models that use the stream for other inferences are therefore incomplete.
 % Aims
-Here we aim to measure distance and kinematic properties along the Pal 5 stream in order to motivate improved models of the Pal 5 stream.
+Here we aim to measure distance and kinematic properties along the Pal 5 stream in order to motivate improved models of the system. %Pal 5 stream.
 We use a large catalog of RR Lyrae-type stars (RRLs) with astrometric data from the \Gaia\ mission to probabilistically identify RRLs in the Pal 5 stream.
@@
 \todo{Cecilia: write this paragraph? you probably have some canned text? - (In progress, CM.}
-RR Lyrae stars (RRLs) are pulsating horizontal branch giants intrinsically more luminous than main sequence stars, but about $\sim 10,000$ time less numerous.
-They can be reliably identified based on their photometric variability alone, with little to no contamination from other types of variable stars, and are well known standard candles.
-Their distances can be estimated based on photometric data alone by means of their period-luminosity relation, with distance errors that go from $\sim8\%$ in the optical to as low as 3\% in the infrared \citep{Neeley:2017, Sesar:2017a}.
-At its current distance, Pal~5's horizontal branch is bright enough $\G\sim17.3$ to be observed by \Gaia, so it's full extent can be traced and kinematic properties measured with RRLs.
+
+RR Lyrae stars (RRLs) are pulsating horizontal branch giants intrinsically more luminous than main sequence turn-off stars.
+Although about $\sim 10,000$ time less numerous than main sequence stars, RRLs are extremely useful probes of the Galactic halo as they can be observed to large distances \citep{Medina2018,Sesar2017c}; they are reliably identified based on their photometric variability, with little to no contamination from other types of variable stars \citep[e.g.]{Holl2018,Drake2017,Mateu2012}; and are well established as standard candles.
+Their distances can be estimated based on photometric data alone, with errors that go from $\sim8\%$ in the optical to as low as 3\% in the infrared \citep{Neeley:2017, Sesar:2017a}.
+
+Previous studies have found RRLs in Pal5's tails: the first two were found by \citet{Vivas2005} in the QUEST survey; more recently, \citep{Ibata:2017} found about ten new RRLs using the catalogue of PS1 candidate RRLs from \citet{Hernitschek2015}.
+At its current distance, Pal~5's horizontal branch is bright enough ($\G\sim17.3$) to be observed by \Gaia.
+Combining \Gaia~DR2 proper motions with the newer PS1 RRL catalogue from \citet{Sesar2017b} --with improved classification probabilities and light curve periods--  will allows to make precise distance and proper motion measurements along the full extent of the Pal~5 stream.
+Both, key ingredients for dynamical modelling of the Pal~5 stream...
 %meaning we can measure kinematic properties of the full stream...
@@
 
+%---yes, I rewrite things many times. C.---
+%RR Lyrae stars (RRLs) are pulsating horizontal branch giants intrinsically more luminous than main sequence stars, but about $\sim 10,000$ time less numerous.
+At its current distance, Pal~5's horizontal branch is bright enough ($\G\sim17.3$) to be observed by \Gaia, so it's full extent can be traced and it's kinematic properties measured using RRLs.
+These stars can be reliably identified based on their photometric variability, with little to no contamination from other types of variable stars \citep[e.g.]{Holl2018,Drake2017,Mateu2012}, and are well known standard candles.
+Their distances can be estimated based on photometric data alone by means of their period-luminosity relation, with distance errors that go from $\sim8\%$ in the optical to as low as 3\% in the infrared \citep{Neeley:2017, Sesar:2017a}.
+Previous studies have found RRLs in Pal5's tails: In the QUEST survey, \citet{Vivas2005} found 2 RRLs; while  \citep{Ibata:2017} found 12 RRLs using the catalogue of PS1 candidate RRLs from \citet{Hernitschek2015}.
+Combining \Gaia~DR2 proper motions with the newer PS1 RRL catalogue from \citet{Sesar2017b} --with improved classification probabilities and light curve periods--  will allow for precise distance and proper motion measurements along the full extent of the Pal~5 stream.
 In this Article, we construct a probabilistic model for kinematic properties of RRLs in the vicinity of the Pal 5 stream to determine membership probabilities and thus map the distance and proper motion trends in the stream.

### 2019-08-21 paper/main.tex
Updates from Overleaf

@@
 
-\todo{Cecilia: write this paragraph? you probably have some canned text? - (In progress, CM.}
-
 RR Lyrae stars (RRLs) are pulsating horizontal branch giants intrinsically more luminous than main sequence turn-off stars.
-Although about $\sim 10,000$ time less numerous than main sequence stars, RRLs are extremely useful probes of the Galactic halo as they can be observed to large distances \citep{Medina2018,Sesar2017c}; they are reliably identified based on their photometric variability, with little to no contamination from other types of variable stars \citep[e.g.]{Holl2018,Drake2017,Mateu2012}; and are well established as standard candles.
+Although about $\sim 10,000$ time less numerous than main sequence stars, RRLs are extremely useful probes of the Galactic halo as they are well established as standard candles; they can be observed to large distances \citep{Medina2018,Sesar2017c}; and are reliably identified based on their photometric variability, with little to no contamination from other types of variable stars \citep[e.g.]{Holl2018,Drake2017,Mateu2012}.
 Their distances can be estimated based on photometric data alone, with errors that go from $\sim8\%$ in the optical to as low as 3\% in the infrared \citep{Neeley:2017, Sesar:2017a}.
 
-Previous studies have found RRLs in Pal5's tails: the first two were found by \citet{Vivas2005} in the QUEST survey; more recently, \citep{Ibata:2017} found about ten new RRLs using the catalogue of PS1 candidate RRLs from \citet{Hernitschek2015}.
+Previous studies have found RRLs in Pal5's tails: the first two were found by \citet{Vivas2005} in the QUEST survey; and more recently \citep{Ibata:2017} found about ten new RRLs using the catalogue of PS1 candidate RRLs from \citet{Hernitschek2015}.
 At its current distance, Pal~5's horizontal branch is bright enough ($\G\sim17.3$) to be observed by \Gaia.
-Combining \Gaia~DR2 proper motions with the newer PS1 RRL catalogue from \citet{Sesar2017b} --with improved classification probabilities and light curve periods--  will allows to make precise distance and proper motion measurements along the full extent of the Pal~5 stream.
+Therefore, combining \Gaia~DR2 proper motions with the newer PS1 RRL catalogue from \citet{Sesar2017b} --with improved classification probabilities and light curve periods--  will allows to make precise distance and proper motion measurements along the full extent of the Pal~5 stream.
 Both, key ingredients for dynamical modelling of the Pal~5 stream...
@@
 This makes it unexpected to have found this many RRLs in the tails.
-Based on prior knowledge of the mass loss undergone by the cluster we would have expected this to be XX (Sarah?/Adrian?) \sa{Not sure, the lowest mass stars should have been lost first...}.
-Similarly,
+Based on prior knowledge of the mass loss undergone by the cluster we would have expected this to be XX (Sarah?/Adrian?) Similarly,
 
@@
 
+\acknowledgments
+This work was performed in part during the Gaia19 workshop and the 2019 Santa Barbara Gaia Sprint (also supported by the Heising-Simons Foundation), both hosted by the Kavli Institute for Theoretical Physics at the University of California, Santa Barbara.
+The Flatiron Institute is supported by the Simons Foundation.
+CM acknowledges funding from the MIA program at Universidad de la Republica, Uruguay.
+
 \software{

### 2019-08-21 paper/main.tex
minor

--- 2019-08-21 paper/main.tex
@@
 \keywords{globular clusters: individual: Palomar 5 ---
+stars: variables: RR Lyrae ---
 Galaxy: halo ---
-stars: variables: RR Lyrae ---
 Galaxy: structure}
@@
 
-Stellar streams are useful objects because they almost\footnote{More precisely, the phase-space distribution of stars along a stream encodes the orbital history of the progenitor system \citep[e.g.,][]{Sanders:2013}.} delineate the past and future trajectory of their progenitor system, offering information about orbits that can be used to infer the mass distribution from the single kinematic snapshot of the Galaxy that we are afforded \citep[e.g.,][]{Johnston:1999, Sanders:2013, PriceWhelan:2014, Bonaca:2018}.
+Stellar streams are useful objects because they almost \citep[see][]{Sanders:2013} delineate the past and future trajectory of their progenitor system, offering information about orbits that can be used to infer the mass distribution from the single kinematic snapshot of the Galaxy that we are afforded \citep[e.g.,][]{Johnston:1999, Sanders:2013, PriceWhelan:2014, Bonaca:2018}.
 % Binney:2008, Eyre:2009,
@@
 However, at its present distance, main sequence stars are barely detected and have large astrometric uncertainties in \Gaia\ \DR{2}, and other stellar tracers (e.g., giant branch stars) are not numerous enough or precise enough distance indicators to provide useful relative distance information along the stream.
-% Enter RR Lyrae-type stars.
+% Enter: RR Lyrae-type stars.
 

### 2019-08-22 paper/main.tex
Updates from Overleaf

@@
 RR Lyrae stars (RRLs) are pulsating horizontal branch giants intrinsically more luminous than main sequence turn-off stars.
-Although about $\sim 10,000$ time less numerous than main sequence stars, RRLs are extremely useful probes of the Galactic halo as they are well established as standard candles; they can be observed to large distances \citep{Medina2018,Sesar2017c}; and are reliably identified based on their photometric variability, with little to no contamination from other types of variable stars \citep[e.g.]{Holl2018,Drake2017,Mateu2012}.
+Although about $\sim 10,000$ time less numerous than main sequence stars, RRLs are extremely useful probes of the Galactic halo as they are well established as standard candles; they can be observed to large distances \citep{Medina:2018,Sesar:2017c}; and are reliably identified based on their photometric variability, with little to no contamination from other types of variable stars \citep[e.g.]{Holl:2018,Drake:2017,Mateu:2012}.
 Their distances can be estimated based on photometric data alone, with errors that go from $\sim8\%$ in the optical to as low as 3\% in the infrared \citep{Neeley:2017, Sesar:2017a}.
 
-Previous studies have found RRLs in Pal5's tails: the first two were found by \citet{Vivas2005} in the QUEST survey; and more recently \citep{Ibata:2017} found about ten new RRLs using the catalogue of PS1 candidate RRLs from \citet{Hernitschek2015}.
+Previous studies have found RRLs in Pal5's tails: the first two were found by \citet{Vivas:2001} in the QUEST survey; and more recently \citep{Ibata:2017} found about ten new RRLs using the catalogue of PS1 candidate RRLs from \citet{Hernitschek:2016}.
 At its current distance, Pal~5's horizontal branch is bright enough ($\G\sim17.3$) to be observed by \Gaia.
-Therefore, combining \Gaia~DR2 proper motions with the newer PS1 RRL catalogue from \citet{Sesar2017b} --with improved classification probabilities and light curve periods--  will allows to make precise distance and proper motion measurements along the full extent of the Pal~5 stream.
+Therefore, combining \Gaia~DR2 proper motions with the newer PS1 RRL catalogue from \citet{Sesar:2017b} --with improved classification probabilities and light curve periods--  will allows to make precise distance and proper motion measurements along the full extent of the Pal~5 stream.
 Both, key ingredients for dynamical modelling of the Pal~5 stream...

### 2019-08-22 paper/main.tex
trimmming and reorg edits

@@
 
-% \author{more people?}
-
 \begin{abstract}
@@
 Thin stellar streams, formed from the tidal disruption of globular clusters, are important gravitational tools, sensitive to both global and small-scale properties of dark matter.
-The Palomar 5 stellar stream (Pal 5) is an exemplar stream within the Milky Way: Its $\approx 20\degr$ tidal tails connect back to the progenitor cluster, and the stream has been used to study the shape and total mass of, along with substructure in, the dark matter distribution of the Galaxy.
+The Palomar 5 stellar stream (Pal 5) is an exemplar stream within the Milky Way: Its $\sim 20\degr$ tidal tails connect back to the progenitor cluster, and the stream has been used to study the shape and total mass of, along with substructure in, the dark matter distribution of the Galaxy.
 However, most details of the phase-space distribution of the stream are not fully explained, and dynamical models that use the stream for other inferences are therefore incomplete.
@@
 
+\begin{figure*}[t!]
+\begin{center}
+\includegraphics[width=\textwidth]{proper-motion.pdf}
+\caption{\textbf{Left}: Proper motions of RRLs in the region $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$ with distances in the range $18 < d < 25~\kpc$.
+\textbf{Middle}: The same, but for RRLs within $1~\degr$ of the mean sky track of Pal 5 \citep{Bonaca:2019}.
+The over-density of stars near $(-2.7, -2.7)$ is the stream, and the two colored points show the Pal 5 cluster proper motions from \citet{Vasiliev:2019} and \citet{Fritz:2015}, respectively.
+\textbf{Right}: The same, but for RRLs excluding $2~\degr$ around the Pal 5 sky track.}
+\label{fig:pm}
+\end{center}
+\end{figure*}
+
 % From \Gaia~DR2 we have combined the two available RRL catalogues: the Specific Objects Study \citep[SOS][]{Clementini2018} and vari-classifier \citep[VC][]{Holl2018,Rimoldini2018}.
@@
 
-\begin{figure*}[th]
-\begin{center}
-\includegraphics[width=\textwidth]{proper-motion.pdf}
-\caption{\textbf{Left}: Proper motions of RRLs in the region $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$ with distances in the range $18 < d < 25~\kpc$.
-\textbf{Middle}: The same, but for RRLs within $1~\degr$ of the mean sky track of Pal 5 \citep{Bonaca:2019}.
-The over-density of stars near $(-2.7, -2.7)$ is the stream, and the two colored points show the Pal 5 cluster proper motions from \citet{Vasiliev:2019} and \citet{Fritz:2015}, respectively.
-\textbf{Right}: The same, but for RRLs excluding $2~\degr$ around the Pal 5 sky track.}
-\label{fig:pm}
-\end{center}
-\end{figure*}
-
 \citet{Sesar:2017b} provide distances to the RRLs based on an $i$-band Period-Luminosity-Metallicity (PLZ) relation.
@@
 Although not all PS1 RRLs have been identified as RRLs from \Gaia\ photometry, all of the PS1 RRL are present in the main point source catalog.
+\figurename~\ref{fig:pm} demonstrates that Pal 5 stream members appear in the RRL catalog: This figure shows the \Gaia\ \DR{2} proper motions for all RRLs in the sky window defined above within the distance range $18 < d < 25~\kpc$, then the same for RRLs in a tight selection around the observed stream track, and then finally the same for RRLs \emph{excluding} the stream region.
 
-%The combined \Gaia+PS1 catalogue has a total of XXX RRL stars.
+\citet{Sesar:2017b} provide an estimate of their catalog's completeness as a function of distance, at high galactic latitude, based on simulations.
+They estimate the mean completeness to be 92\% for \rrab~and 79\% for \rrc, up to a distance of 40~kpc, well beyond the distance of Pal~5 and its tails.
+At the average ratio of 3 \rrc~stars per every 10 \rrab~\citep{Layden1995}, this represents a mean completeness of 89\%.
+We can also produce an estimate per line of sight using the methodology described in \citet{Rybizki2018}, which requires two independent catalogues to assess the completeness of both in a probabilistic manner.
+Using the \Gaia~DR2 VariClassifier \citet{Holl2018,Rimoldini2018} and Specific Objects Study \citep{Clementini2018} catalogues, we find a median completeness of 92\% with a standard deviation of 16\%  with no evident spatial trend across the field of Figure~\ref{fig:members}.
 
@@
 To start, we use the sky track and width of Pal 5 from \citet{Bonaca:2019} to mask out a $\sim4^\circ$-wide region around the stream---and an extrapolation of the known extent of the stream---in order to define a ``background'' region that should be devoid of Pal 5 member stars.
-\figurename~\ref{fig:pm} demonstrates that Pal 5 stream members appear in the RRL catalog: This figure shows the \Gaia\ \DR{2} proper motions for all RRLs in the sky window defined above within the distance range $18 < d < 25~\kpc$, then the same for RRLs in a tight selection around the observed stream track, and then finally the same for RRLs \emph{excluding} the stream region.
 We use RRL stars in the background region to construct an error-deconvolved density model in the 3D space of proper motion components and distance.
@@
 \figurename~\ref{fig:trackmembers} shows a summary of the inferred membership model properties in sky position (shown in Pal 5 coordinates $\phi_1, \phi_2$), distance, and proper motion components.
+\tablename~\ref{t:rrl_members} shows an abbreviated version of our catalog of RRL in the region around Pal 5, with PS1 and \Gaia data and membership probabilities appended.
 
-\section{Results} \label{sec:results}
-
-\subsection{RR Lyrae stars associated with Pal 5 and its tails}
-
-We find a total of 24 RR Lyrae stars with membership probability $> 0.5$ that lie within $1\degr$ of the stream track \citep[again using the track from][]{Bonaca:2019}.
-Of those, 13 are classified as type \typeab\ and 11 are classified as type \typec\ RRLs.
-We find a total of 10 RRLs within the estimated Jacobi radius of the cluster ($\approx 12'$), meaning that 15 RRLs are spread between the leading and trailing tails.
-\figurename~\ref{fig:members} shows the sky positions of all RRL with membership probability $>0.01$ over-plotted on the density of filtered main sequence stars from \citet{Bonaca:2019}.
-We note that there are a few RRLs outside of the main stream track that appear to be at a consistent distance with consistent proper motions (e.g., the two stars near $230\degr, 0\degr$) that are not included in the tally above.
-No theoretical models of Pal 5 have predicted tidal debris at such locations, so follow-up spectroscopy to confirm or rule out their origin would be very informative.
-
-\subsection{Kinematics of the stream and cluster}
-
-We find a mean Pal 5 cluster distance of $20.9 \pm 0.2~\kpc$ using the PS1 RRLs and the membership model described in \sectionname~\ref{sec:membership}.
-This is significantly closer than past measurements of the cluster distance that have used \todo{summarize past distance measurements and why this could be so different...}.
-We also detect a distance trend from the closer leading tail (at $\approx 19~\kpc$) to the more distant trailing tail (at $\approx 22~\kpc$).
-While the proper motion in right ascension appears consistent with constant or only a mild gradient over the extent of the stream, we see a significant gradient in the declination component, with a difference of about $\approx 1~\masyr$ from trailing to leading tail.
-% \todo{Make a table with summary information?}
-
-\subsection{RR Lyrae population }
-
-Based on the paucity of RRL stars of type~\typeab~in the cluster, Pal~5  would have been expected to be classified as an Oosterhoff type II cluster (ref-XX, no one really seems to say this explicitly), at odds with the cluster's known spectroscopic metallicity of $\FeH=-1.32$ \citep{TODO}, as OoII clusters are on average more metal-poor than OoI clusters.
-
-We find 15 RRLs in the cluster tails, $\sim11$ of \typeab\ and $\sim4$ of \typec.
-Combined with the 10~\rrc~in the cluster itself, there are 25 RRLs in total associated to Pal~5 and its tails with membership probabilities larger than 0.5.
-Figure~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
-The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram.
-The distribution of the \typeab~RRL clearly indicate Pal~5 to be an Oosterhoff I cluster, consistently with their mean period of 0.565~d(XX).
-Our findings of the updated ratio of \rrab~to~\rrc, mean period and Period-Amplitude distribution of the \rrab, all support an OoI classification for Pal~5, consistent with the cluster's metallicity.
-
-Out of our newly found \rrab's three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars.
-This type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$][]{Monelli:2017} and the Sagittarius dwarf galaxy, but not in any other dwarf spheroidal galaxy.
-\todo{CM:( any other point to be made about this?)}.
-
-\begin{figure*}[t]
-\begin{center}
-\includegraphics[width=\textwidth]{rrls_PA.pdf}
-\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field.}
-\label{fig:PA_diagram}
-\end{center}
-\end{figure*}
-
-\subsection{Stellar Mass and Luminosity}
-
-%We can provide an estimate on the cluster's luminosity based on the observed number of RRLs. For that, the first step is to assess how complete our observed sample is.
-
-\citet{Sesar:2017b} provide an estimate of their catalogue's completeness as a function of distance, at high galactic latitude, based on simulations.
-They estimate the mean completeness to be 92\% for \rrab~and 79\% for \rrc, up to a distance of 40~kpc, well beyond the distance of Pal~5 and its tails.
-At the average ratio of 3 \rrc~stars per every 10 \rrab~\citep{Layden1995}, this represents a mean completeness of 89\%.
-We can also produce an estimate per line of sight using the methodology described in \citet{Rybizki2018}, which requires two independent catalogues to assess the completeness of both in a probabilistic manner.
-Using the \Gaia~DR2 VariClassifier \citet{Holl2018,Rimoldini2018} and Specific Objects Study \citep{Clementini2018} catalogues, we find a median completeness of 92\% with a standard deviation of 16\%  with no evident spatial trend across the field of Figure~\ref{fig:members}.
-
-%we obtain the completeness map shown in Figure~\ref{fig:completeness_map}, for RRLs of type \typeab~and \typec.
-The mean completeness estimated this way is 92\% with a standard deviation of 16\% and, as the figure shows, there is no evident spatial trend.
-The two estimates are remarkably consistent and, in what follows, we will adopt a completeness of 90\%-XX.
-
-To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu:2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters.
-Here we use it including the data for the \typec~stars.
-Correcting the observed number of 25~RRLs by the median completeness of 92\%, we get a total expected number of 28~RRLs, for which find an absolute magnitude $M_V=-8.1\pm 1.1$.
-
-% \begin{figure}
-% \begin{center}
-% \includegraphics[width=\columnwidth]{ps1_completeness.pdf}
-% \caption{PS1 RRL estimated completeness. The Pal 5 stream's track is shown with the solid line.}\label{fig:completeness_map}
-% \end{center}
-% \end{figure}
-
-\todo{CM: estimate luminosity based on number of RRLs. Compare to estimate from $M_V-\FeH$ relation. Derive stellar mass from stellar luminosity. Compare to dynamical mass estimates.}
-
-- there are $\sim4$-XX times as many RRLs in the tails as there are in the cluster.
-- how much mass did we already know to be in the tails compared to the cluster?
-(\sa{according to Ibata 2016 the mass in the present day cluster is 4297 $\pm~ 98$ M$_{\odot}$, and the streams excluding the cluster have $\sim$7903 M$_{\odot}$.
-So under a factor of 2 more stars in the tails than in the remnant cluster}.
-The cluster is thought to have had an initial mass of $\sim47000$ M$_{\odot}$).
-Probably not as much as a factor of 4.
-This makes it unexpected to have found this many RRLs in the tails.
-Based on prior knowledge of the mass loss undergone by the cluster we would have expected this to be XX (Sarah?/Adrian?) Similarly,
-
-\section{Discussion}
-
-\subsection{RR Lyrae stars known to be associated with Pal 5 or it's tails}
-
-\todo{move some of this to discussion?}
-
-Pal~5 has 5 known RRL stars (V1-V5), all of type \typec, as reported in the \citet{Clement2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{SawyerHogg1973}.
-Star V2 has a  systemic radial velocity of  $-65\pm16$~\kms~measured by \citet{Vivas2005}.
-
-In their analysis of the halo density profile \citet{Vivas2006} report an overdensity, dubbed Group~6, containing 6 previously unidentified RRLs.
-Based on the distance to the RRLs they claim many of these could be associated to Pal~5's tidal tails.
-The QUEST IDs of the stars in their Group~6 are: 403, 405, 407, 411, 416 and 422.
-The first two stars --403 and 405-- are located at distances of  $9\farcm8$ and $11'$ from the cluster's center, which \citet{Vivas2004} notes places them within the cluster's tidal radius, assuming the value of $r_t\sim16'$ reported by \citet{Odenkirchen:2002}.
-\todo{Update this value}.
-\citet{Vivas2006} also note that star 393, not in Group~6, could also be associated to the cluster's tails, given it's distance and location along the known extent of the tails.
-
-In summary,  previous studies report 7 known RRLs associated to the cluster itself (1 \typeab~and 6 \typec) plus 5 more reported as possibly associated to the tails (4 \typeab~and 2 \typec).
-
-%
 \begin{table*}[t]
@@
 
-\subsubsection{Notes on individual stars}
+\section{Results} \label{sec:results}
 
-Star 422 is reported as an \rrc~in \citet{Vivas2004}, with a period of 0.61925~d and $V$ amplitude 0.51~mag.
-In PS1 it is classified as an \rrc~with a period of 0.3832276153~d.
-The period reported in \citet{Vivas2004} is an alias of this one, more likely to be the true period given the variability amplitude\footnote{\citet{Vivas2004} also cautions that derived periods for QUEST \rrc~stars are known to be prone to aliasing due to the observing cadence of $\sim1$~d.}.
-Therefore, we keep PS1's classification henceforth for RRL 422.
+\subsection{RR Lyrae stars associated with Pal 5 and its tails}
 
-Star \verb+source_id=6339498589346000768+ is reported both in PS1 and \Gaia~as an \rrc~, with a period 0.4714315854~d in PS1 and 0.320110123108~d in \Gaia~SOS, the two are a pair of 1~d aliases \citep[see e.g.]{Lafler1965}.
-In what follows we adopt the shorter period reported in \Gaia~as the correct one as it is the more probable one for an \rrc~star.
+We find a total of 24 RR Lyrae stars with membership probability $> 0.5$ that lie within $1\degr$ of the stream track \citep[again using the track from][]{Bonaca:2019}.
+Of those, 13 are classified as type \typeab\ and 11 are classified as type \typec\ RRLs.
+We find a total of 10 RRLs within the estimated Jacobi radius of the cluster ($\approx 12'$), meaning that 15 RRLs are spread between the leading and trailing tails.
+\figurename~\ref{fig:members} shows the sky positions of all RRL with membership probability $>0.01$ over-plotted on the density of filtered main sequence stars from \citet{Bonaca:2019}.
+We note that there are a few RRLs outside of the main stream track that appear to be at a consistent distance with consistent proper motions (e.g., the two stars near $\alpha = 230\degr$, $\delta = 0\degr$) that are not included in the tally above.
+No theoretical models of Pal 5 have predicted tidal debris at such locations, so follow-up spectroscopy to confirm or rule out their origin would be very informative.
 
-Star \verb+source_id=4418731829516302848+ is reported in PS1 and \Gaia~as an \rrab~with a period around 0.646 and a similarly low amplitude $A_G=0.2$ and $A_g=0.3$ in both surveys.
-This star is reported in CRTS \citep{Drake2014} (ID CSS\_J151623.2-000831) as an \rrc~with a period 0.28164~d.
-We adopt the CRTS period for this star in what follows, as it is based on many more observations ($\gtrsim200$) than \Gaia~and PS1's (about a dozen per filter in each case) and the \rrc~classification and short period are more likely for such a short amplitude star.
+\subsection{Kinematics of the cluster and stream}
+
+Past distance measurements to the Pal 5 cluster have ranged from $20.9~\kpc$ \citep{Dotter:2011} to $23.2~\kpc$ \citep{Harris:1996}, with a previous measurement from a small number of RR Lyrae stars in between at $22.3~\kpc$ \citep{Vivas:2006}.
+We find a mean Pal 5 cluster distance of $20.9 \pm 0.2~\kpc$ using the PS1 RRLs and the membership model described in \sectionname~\ref{sec:membership}
+This is closer than most past measurements of the cluster distance, but consistent with the measurement from \citet{Dotter:2011}.
+We find a proper motion $(\mu_\alpha\,\cos\delta, \mu_\delta) = (-2.69, -2.70) \pm (0.04, 0.03)~\masyr$, consistent with the measurement in \citet{Vasiliev:2019}.
+The larger proper motion (relative to \citealt{Fritz:2015} and other earlier measurements) and the closer distance means that the tangential velocity of Pal 5 is $\approx 30~\kms$ larger than most previous models have assumed.
+
+We also detect a distance trend from the closer leading tail ($d \approx 19~\kpc$ at $\phi_1 \approx 5\degr$) to the more distant trailing tail ($d \approx 22~\kpc$ at $\phi_1 \approx -5\degr$).
+While the proper motion in right ascension appears consistent with constant or only a mild gradient over the extent of the stream, we see a significant gradient in the declination component, with a difference of about $\approx 1~\masyr$ from trailing to leading tail.
+% \todo{Make a table with summary information?}
+
+\subsection{RR Lyrae population }
+
+% Based on the prior existence of  paucity of type~\typeab\ RRL stars of in the cluster, Pal~5  would have been expected to be classified as an Oosterhoff type II cluster (ref-XX, no one really seems to say this explicitly), at odds with the cluster's known spectroscopic metallicity of $\FeH=-1.32$ \citep{TODO}, as OoII clusters are on average more metal-poor than OoI clusters.
+
+Assuming a Jacobi radius of $12^\prime$, we find 15 RRLs in the cluster tails, 11 classified as type~\typeab\ and 4 type~\typec.
+Combined with the 2 type~\typeab\ and 7 type~\typec\ RRLs in the cluster itself, there are 24 RRLs in total associated to Pal~5 and its tails with membership probabilities larger than 0.5.
+In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.2$.
+\figurename~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
+The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram.
+The distribution of the \typeab~RRL clearly indicate Pal~5 to be an Oosterhoff I cluster, consistent with their mean period of $0.565~d$ \todo{(XX) ?}.
+Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an OoI classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.32$ \citep{TODO}.
+Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars.
+This type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017} and the Sagittarius dwarf galaxy, but not in any other dwarf spheroidal galaxy.
+\todo{CM:( any other point to be made about this?)}.
+
+\begin{figure*}[t]
+\begin{center}
+\includegraphics[width=\textwidth]{rrls_PA.pdf}
+\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field.}
+\label{fig:PA_diagram}
+\end{center}
+\end{figure*}
+
+\subsection{Stellar Mass and Luminosity}
+
+%We can provide an estimate on the cluster's luminosity based on the observed number of RRLs. For that, the first step is to assess how complete our observed sample is.
+
+%we obtain the completeness map shown in Figure~\ref{fig:completeness_map}, for RRLs of type \typeab~and \typec.
+The mean completeness estimated this way is 92\% with a standard deviation of 16\% and, as the figure shows, there is no evident spatial trend.
+The two estimates are remarkably consistent and, in what follows, we will adopt a completeness of 90\%-XX.
+
+To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu:2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters, here including the RR\typec~stars.
+Correcting the observed number of 24~RRLs by the median completeness of 92\%, we get a total expected number of 26~RRLs, for which find an absolute magnitude $M_V=-8.1\pm 1.1$ \todo{CM: update this number based on 26}.
+We therefore find that the initial mass of Pal 5 \todo{...}.
+
+% \begin{figure}
+% \begin{center}
+% \includegraphics[width=\columnwidth]{ps1_completeness.pdf}
+% \caption{PS1 RRL estimated completeness. The Pal 5 stream's track is shown with the solid line.}\label{fig:completeness_map}
+% \end{center}
+% \end{figure}
+
+% \todo{CM: estimate luminosity based on number of RRLs. Compare to estimate from $M_V-\FeH$ relation. Derive stellar mass from stellar luminosity. Compare to dynamical mass estimates.}
+
+% - there are $\sim4$-XX times as many RRLs in the tails as there are in the cluster.
+% - how much mass did we already know to be in the tails compared to the cluster?
+(\sa{according to Ibata 2016 the mass in the present day cluster is 4297 $\pm~ 98$ M$_{\odot}$, and the streams excluding the cluster have $\sim$7903 M$_{\odot}$.
+So under a factor of 2 more stars in the tails than in the remnant cluster}.
+The cluster is thought to have had an initial mass of $\sim47000$ M$_{\odot}$).
+Probably not as much as a factor of 4.
+This makes it unexpected to have found this many RRLs in the tails.
+Based on prior knowledge of the mass loss undergone by the cluster we would have expected this to be XX (Sarah?/Adrian?) Similarly,
+
+\section{Discussion}
+
+The Pal~5 cluster has 5 previous known RRL stars (V1-V5), all of type \typec, as reported in the \citet{Clement2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{SawyerHogg1973}.
+Star V2 has a  systemic radial velocity of  $-65\pm16$~\kms~measured by \citet{Vivas2005}.
+
+In their analysis of the halo density profile \citet{Vivas2006} report an overdensity, dubbed Group~6, containing 6 previously unidentified RRLs.
+Based on the distance to the RRLs they claim many of these could be associated to Pal~5's tidal tails.
+The QUEST IDs of the stars in their Group~6 are: 403, 405, 407, 411, 416 and 422.
+The first two stars --403 and 405-- are located at distances of  $9\farcm8$ and $11'$ from the cluster's center, which \citet{Vivas2004} notes places them within the cluster's tidal radius, assuming the value of $r_t\sim16'$ reported by \citet{Odenkirchen:2002}.
+\todo{Update this value}.
+\citet{Vivas2006} also note that star 393, not in Group~6, could also be associated to the cluster's tails, given it's distance and location along the known extent of the tails.
+
+In summary,  previous studies report 7 known RRLs associated to the cluster itself (1 \typeab~and 6 \typec) plus 5 more reported as possibly associated to the tails (4 \typeab~and 2 \typec).
+
+\todo{update this with a comparison to Ibata+2017?}
 
@@
 }
+\clearpage
+\appendix
+
+\section{Notes on individual stars}
+
+Star 422 is reported as an \rrc~in \citet{Vivas2004}, with a period of 0.61925~d and $V$ amplitude 0.51~mag.
+In PS1 it is classified as an \rrc~with a period of 0.3832276153~d.
+The period reported in \citet{Vivas2004} is an alias of this one, more likely to be the true period given the variability amplitude\footnote{\citet{Vivas2004} also cautions that derived periods for QUEST \rrc~stars are known to be prone to aliasing due to the observing cadence of $\sim1$~d.}.
+Therefore, we keep PS1's classification henceforth for RRL 422.
+
+Star \verb+source_id=6339498589346000768+ is reported both in PS1 and \Gaia~as an \rrc~, with a period 0.4714315854~d in PS1 and 0.320110123108~d in \Gaia~SOS, the two are a pair of 1~d aliases \citep[see e.g.]{Lafler1965}.
+In what follows we adopt the shorter period reported in \Gaia~as the correct one as it is the more probable one for an \rrc~star.
+
+Star \verb+source_id=4418731829516302848+ is reported in PS1 and \Gaia~as an \rrab~with a period around 0.646 and a similarly low amplitude $A_G=0.2$ and $A_g=0.3$ in both surveys.
+This star is reported in CRTS \citep{Drake2014} (ID CSS\_J151623.2-000831) as an \rrc~with a period 0.28164~d.
+We adopt the CRTS period for this star in what follows, as it is based on many more observations ($\gtrsim200$) than \Gaia~and PS1's (about a dozen per filter in each case) and the \rrc~classification and short period are more likely for such a short amplitude star.
 

### 2019-08-22 paper/main.tex
Updates from Overleaf

@@
 The Palomar 5 stellar stream (Pal 5) is an exemplar stream within the Milky Way: Its $\sim 20\degr$ tidal tails connect back to the progenitor cluster, and the stream has been used to study the shape and total mass of, along with substructure in, the dark matter distribution of the Galaxy.
+%SP the above sentence is a bit long
 However, most details of the phase-space distribution of the stream are not fully explained, and dynamical models that use the stream for other inferences are therefore incomplete.
@@
 We use a large catalog of RR Lyrae-type stars (RRLs) with astrometric data from the \Gaia\ mission to probabilistically identify RRLs in the Pal 5 stream.
+%SP throughout the manuscript, check if you mean plural when using RRL vs RRLs.
 RRLs are useful because they are intrinsically-luminous standard candles and their distances can be inferred with small relative precision ($\sim3\%$).
@@
 % Results
-By building a probabilistic model of the Pal 5 cluster and stream in proper motion and distance, we find 25 RRL consistent with being members of the cluster (10) and stream (15).
-Using these RRL, we detect gradients in distance and proper motion along the stream, and provide an updated measurement of the distance to the Pal 5 cluster using the RRL.
+By building a probabilistic model of the Pal 5 cluster and stream in proper motion and distance, we find 25 RRLs consistent with being members of the cluster (10) and stream (15).
+Using these RRLs, we detect gradients in distance and proper motion along the stream, and provide an updated measurement of the distance to the Pal 5 cluster using the RRLs.
 % Conclusions
@@
 
-Previous studies have found RRLs in Pal5's tails: the first two were found by \citet{Vivas:2001} in the QUEST survey; and more recently \citep{Ibata:2017} found about ten new RRLs using the catalogue of PS1 candidate RRLs from \citet{Hernitschek:2016}.
+Previous studies have found RRLs in Pal5's tails: the first two were found by \citet{Vivas:2001} in the QUEST survey; and more recently \citep{Ibata:2017} found about ten new RRLs using the catalog of PS1 candidate RRLs from \citet{Hernitschek:2016}.
 At its current distance, Pal~5's horizontal branch is bright enough ($\G\sim17.3$) to be observed by \Gaia.
-Therefore, combining \Gaia~DR2 proper motions with the newer PS1 RRL catalogue from \citet{Sesar:2017b} --with improved classification probabilities and light curve periods--  will allows to make precise distance and proper motion measurements along the full extent of the Pal~5 stream.
+Therefore, combining \Gaia~DR2 proper motions with the newer PS1 RRL catalog from \citet{Sesar:2017b} --with improved classification probabilities and light curve periods--  will allows to make precise distance and proper motion measurements along the full extent of the Pal~5 stream.
 Both, key ingredients for dynamical modelling of the Pal~5 stream...
@@
 Their distances can be estimated based on photometric data alone by means of their period-luminosity relation, with distance errors that go from $\sim8\%$ in the optical to as low as 3\% in the infrared \citep{Neeley:2017, Sesar:2017a}.
-Previous studies have found RRLs in Pal5's tails: In the QUEST survey, \citet{Vivas2005} found 2 RRLs; while  \citep{Ibata:2017} found 12 RRLs using the catalogue of PS1 candidate RRLs from \citet{Hernitschek2015}.
-Combining \Gaia~DR2 proper motions with the newer PS1 RRL catalogue from \citet{Sesar2017b} --with improved classification probabilities and light curve periods--  will allow for precise distance and proper motion measurements along the full extent of the Pal~5 stream.
+Previous studies have found RRLs in Pal5's tails: In the QUEST survey, \citet{Vivas2005} found 2 RRLs; while  \citep{Ibata:2017} found 12 RRLs using the catalog of PS1 candidate RRLs from \citet{Hernitschek2015}.
+Combining \Gaia~DR2 proper motions with the newer PS1 RRL catalog from \citet{Sesar2017b} --with improved classification probabilities and light curve periods--  will allow for precise distance and proper motion measurements along the full extent of the Pal~5 stream.
 In this Article, we construct a probabilistic model for kinematic properties of RRLs in the vicinity of the Pal 5 stream to determine membership probabilities and thus map the distance and proper motion trends in the stream.
@@
 
-% From \Gaia~DR2 we have combined the two available RRL catalogues: the Specific Objects Study \citep[SOS][]{Clementini2018} and vari-classifier \citep[VC][]{Holl2018,Rimoldini2018}.
-The resulting VC+SOS catalogue contains a total of 228\,904, with all-sky coverage down to the mission's limiting magnitude $G=20.7$.
-After filtering out stars with contaminated photometry \verb+phot_bp_rp_excess_factor+>2 or null, as recommended in \citep{Rimoldini2018}, the VC+SOS catalogue contains a total of 175\,164 RR Lyrae.
-For the 122\,747 in SOS, which have undergone extensive light curve fitting, periods, amplitudes and other fitting parameters have been provided by \citet{Clementini2018}.
-The remaining 52\,417 stars (from the VC catalogue) did not have enough data points in \Gaia~DR2 to go into the SOS pipeline, so periods and other fitting parameters are not yet available for these stars.
+\todo{This part was rewritten to reflect the catalog we ended up using. CM.}
 
-To search for RRLs in the vicinity of Pal 5, we use a catalog of RRL stars from the PS1 survey that spans $\sim$3/4 of the sky (all sky north of declination $\delta > -30\degr$) and contains 61\,795 RRLs identified from their variability \citep{Sesar:2017b}.
-In detail, the \emph{bona fide} RRLs in the catalog were classified using multi-epoch data in the $grizy$ photometric bands, with a typical number of 12 observations per filter, and has a limiting magnitude $r\sim21$ (corresponding to $G \sim 21$ for the typical colors of RRLs).
-Periods and amplitudes in all five bands are available for all stars in this catalog.
-We have chosen to use this catalog instead of the recently released \Gaia~RRL catalogs \citep[VariClassifier and Specific Object Studies][]{Holl2018, Rimoldini2018, Clementini2018} as these are subject to higher incompleteness in this particular area of the sky due to the \Gaia\ scanning law.
-Nevertheless, the vast majority (61\,625) of PS1's \emph{bona fide} RRLs have astrometric information available in \Gaia~\DR{2}.
+The PanSTARRS-1 (PS1) catalog of RRLs \citep{Sesar:2017b} contains 229K stars spanning $\sim$3/4 of the sky (all sky north of declination $\delta > -30\degr$), with multi-epoch $grizy$ data up to a limiting magnitude $r\sim21$ (corresponding to $G \sim 21$ for the typical colors of RRLs).
+These stars were identified as RRLs using machine-learning methods based on a template fitting algorithm that can find pulsation periods overcoming the sparse sampling of the PS1 survey ($\sim12$ epochs per filter).
 
+Two all-sky RR Lyrae catalogs have also been released as part of \Gaia~\DR{2} \citep[VariClassifier and Specific Object Studies][]{Holl2018, Rimoldini2018, Clementini2018}.
+At present, however, these catalogs are subject to more significant and spatially-varying incompleteness in this particular area of the sky due to the \Gaia\ scanning law \citep[see][]{Rimoldini2018}.
+
+Out of the 229K RRLs in the PS1 catalog, $\sim60$K are also reported as RRLs by \Gaia~\DR{2}, the vast majority of which (49K) are part of the subset defined by \citet{Sesar:2017b} as \emph{bona fide}.
+This subset is comprised of 61K RRLs above classification score thresholds set by \citet{Sesar:2017b} to ensure its high purity ($>90$\%) and completeness ($>80$\%).
+An extra $\sim6$K stars with lower classification scores in PS1 were also found to be RRLs by \Gaia~\DR{2} and have reported pulsation periods in the SOS catalog.
+Most of these RRLs have matching classifications and consistent periods in the two surveys, which increases our confidence that these are true RRLs.
+
+For our present work, therefore, we use the superset of 68\,085 stars consisting of the 61\,795 \emph{bona fide} PS1 RRLs, plus the 6\,459 non-\emph{bona fide} PS1 stars also reported in the SOS \Gaia~\DR{2} catalog.
+Out of these, 68\,085 RRLs (99.8\%) have astrometric information available in \Gaia~\DR{2} and all---by construction---have photometric data and pulsation periods from PS1.
+
+%distances---
 \citet{Sesar:2017b} provide distances to the RRLs based on an $i$-band Period-Luminosity-Metallicity (PLZ) relation.
@@
 At the average ratio of 3 \rrc~stars per every 10 \rrab~\citep{Layden1995}, this represents a mean completeness of 89\%.
-We can also produce an estimate per line of sight using the methodology described in \citet{Rybizki2018}, which requires two independent catalogues to assess the completeness of both in a probabilistic manner.
-Using the \Gaia~DR2 VariClassifier \citet{Holl2018,Rimoldini2018} and Specific Objects Study \citep{Clementini2018} catalogues, we find a median completeness of 92\% with a standard deviation of 16\%  with no evident spatial trend across the field of Figure~\ref{fig:members}.
+We can also produce an estimate per line of sight using the methodology described in \citet{Rybizki2018}, which requires two independent catalogs to assess the completeness of both in a probabilistic manner.
+Using the \Gaia~DR2 VariClassifier \citet{Holl2018,Rimoldini2018} and Specific Objects Study \citep{Clementini2018} catalogs, we find a median completeness of 92\% with a standard deviation of 16\%  with no evident spatial trend across the field of Figure~\ref{fig:members}.
 
@@
 \includegraphics[width=\textwidth]{members.pdf}
-\caption{Sky positions of RRL...todo}
+\caption{Sky positions of RR Lyrae stars with Pal 5 membership probabilities $>0.01$, colored by membership probability (large markers).
+The background greyscale shows the surface density of CMD-filtered main sequence stars from \citet{Bonaca:2019}.}
 \label{fig:members}
@@
 To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu:2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters, here including the RR\typec~stars.
-Correcting the observed number of 24~RRLs by the median completeness of 92\%, we get a total expected number of 26~RRLs, for which find an absolute magnitude $M_V=-8.1\pm 1.1$ \todo{CM: update this number based on 26}.
+The $\log{N_{RR}-M_V$ has quite a large scatter so, particularly for such low numbers of RRLs, it will only allow us to give rough limits on the cluster's absolute magnitude.
+Correcting the observed number of 26~RRLs by the median completeness of 92\%, we get a total expected number of 28~RRLs, for which we find an absolute magnitude $M_V=-8.1\pm 1.1$, which correspond to the mode and 16/84th percentiles of the posterior probability PDF.
 We therefore find that the initial mass of Pal 5 \todo{...}.
@@
 The Flatiron Institute is supported by the Simons Foundation.
-CM acknowledges funding from the MIA program at Universidad de la Republica, Uruguay.
+CM acknowledges funding from the MIA program at Universidad de la Republica, Uruguay, and is grateful for the hospitality and support of the Flatiron Institute where part of this research was carried out.
 

### 2019-08-22 paper/main.tex
disc

--- 2019-08-22 paper/main.tex
@@
 
-\begin{table*}[t]
-\caption{RRL members of the Pal 5 stream.}\label{t:rrl_members}
-\begin{footnotesize}
-\begin{tabular}{llllllll}
-\toprule
-\Gaia~DR2               & Other & Period & Amp-$g$ & Type & D & member & sep  \\
-\verb+source_id+        & ID   & (d)    & (mag)   &     &(kpc)& prob & (") \\
-\midrule
-4418920808077110784 &  V5 & 0.337934& 0.46 & RRc  & 20.73 & 0.9982 &   1.9 \\
-4418920846732620032 &     & 0.643239& 0.27 & RRab & 21.71 & 0.9963 &   2.0 \\
-4418914863842345856 &  V1 & 0.293229& 0.56 & RRc  & 20.36 & 0.9696 &   2.0 \\
-4418726027016125056 &  V3 & 0.329948& 0.51 & RRc  & 20.65 & 0.9978 &   3.8 \\
-4418725889577171328 &  V4 & 0.286366& 0.59 & RRc  & 20.28 & 0.9915 &   4.5 \\
-4418731829516302848 &     & 0.28164 & 0.3  & RRc  & 18.14 & 0.9297 &   4.8 \\
-4418913218870688768 &  V2 & 0.332473& 0.4  & RRc  & 20.33 & 0.9977 &   5.0 \\
-4418734165978521728 &     & 0.380859& 0.52 & RRc  & 22.75 & 0.9850 &   7.6 \\
-4418724034151291776 & 403 & 0.551705& 1.14 & RRab & 21.23 & 0.9987 &   9.7 \\
-4418732791593809152 & 405 & 0.329670& 0.45 & RRc  & 20.74 & 0.9959 &  11.0 \\
-4419052204012341760 &     & 0.530006& 1.01 & RRab & 20.84 & 0.9960 &  44.5 \\
-4418142117622280192 & 393 & 0.462671& 1.36 & RRab & 19.95 & 0.9060 & 118.7 \\
-6339498589346000768 &     & 0.320110& 0.46 & RRc  & 18.89 & 0.9934 & 217.7 \\
-6339499379619987200 &     & 0.329577& 0.47 & RRc  & 19.35 & 0.9966 & 218.6 \\
-6339478312804685824 &     & 0.465883& 1.12 & RRab & 19.91 & 0.9970 & 242.8 \\
-4421078432143578496 &     & 0.520749& 1.11 & RRab & 21.98 & 0.9981 & 246.1 \\
-6339398155830632448 &     & 0.688316& 0.75 & RRab & 19.82 & 0.9982 & 258.9 \\
-4427253907921168000 &     & 0.496334& 1.35 & RRab & 21.81 & 0.9905 & 362.5 \\
-4427220338456828416 &     & 0.600796& 0.81 & RRab & 21.98 & 0.9974 & 384.4 \\
-4427234700827469952 &     & 0.517911& 1.24 & RRab & 21.94 & 0.9979 & 392.3 \\
-6337302795905934336 &     & 0.519462& 1.02 & RRab & 17.29 & 0.9808 & 404.5 \\
-6337233350579111680 &     & 0.542383& 0.96 & RRab & 18.98 & 0.9962 & 450.6 \\
-4427646704153765632 &     & 0.385552& 0.37 & RRc  & 19.90 & 0.9429 & 456.5 \\
-4424705647988505600 &     & 0.335502& 0.41 & RRc  & 20.87 & 0.9832 & 513.0 \\
-4426221707021159424 &     & 0.533313& 1.14 & RRab & 21.97 & 0.9728 & 550.1 \\
-\bottomrule
-\end{tabular}
-\end{footnotesize}
-\end{table*}
-
-\section{Results} \label{sec:results}
+\section{Results and Discussion} \label{sec:results}
 
@@
 
-Assuming a Jacobi radius of $12^\prime$, we find 15 RRLs in the cluster tails, 11 classified as type~\typeab\ and 4 type~\typec.
+Assuming a Jacobi radius of $12^\prime$ (computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.4\times10^4~\msun$ and using the Milky Way mass model in \citealt{gala}), we find 15 RRLs in the cluster tails, 11 classified as type~\typeab\ and 4 type~\typec.
 Combined with the 2 type~\typeab\ and 7 type~\typec\ RRLs in the cluster itself, there are 24 RRLs in total associated to Pal~5 and its tails with membership probabilities larger than 0.5.
@@
 
-\section{Discussion}
+\subsection{Previously known Pal 5 RR Lyrae}
 
-The Pal~5 cluster has 5 previous known RRL stars (V1-V5), all of type \typec, as reported in the \citet{Clement2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{SawyerHogg1973}.
-Star V2 has a  systemic radial velocity of  $-65\pm16$~\kms~measured by \citet{Vivas2005}.
+The Pal~5 cluster has 5 previous known RRL stars (named V1-V5), all of type \typec, as reported in the \citet{Clement:2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{SawyerHogg:1973}.
+In their analysis of the halo density profile with RRLs, \citet{Vivas:2006} report an overdensity (``Group~6'') that they claim could be associated with Pal~5's tidal tails.
+% The QUEST IDs of the stars in their Group~6 are: 403, 405, 407, 411, 416 and 422.
+% The first two stars --403 and 405-- are located at distances of  $9\farcm8$ and $11'$ from the cluster's center, which \citet{Vivas2004} notes places them within the cluster's tidal radius, assuming the value of $r_t\sim16'$ reported by \citet{Odenkirchen:2002}.
+Two of these stars (IDs 403 and 405) are separated by just $\sim10'$ and $11'$ from the cluster's center, which places them within the cluster's Jacobi radius \citep{Vivas:2004}, and an additional star (star 393) outside of the Jacobi radius but at a similar distance could be associated with the stream \citep{Vivas:2006}.
+More recently, \citet{Ibata:2017} use a prior iteration of the PS1 RRL catalog \citep{Hernitschek:2016} to identify \todo{XX} candidate members of the Pal 5 cluster and tails based on sky position and distance alone.
+However, without a published catalog, we cannot compare our sample to theirs.
 
-In their analysis of the halo density profile \citet{Vivas2006} report an overdensity, dubbed Group~6, containing 6 previously unidentified RRLs.
-Based on the distance to the RRLs they claim many of these could be associated to Pal~5's tidal tails.
-The QUEST IDs of the stars in their Group~6 are: 403, 405, 407, 411, 416 and 422.
-The first two stars --403 and 405-- are located at distances of  $9\farcm8$ and $11'$ from the cluster's center, which \citet{Vivas2004} notes places them within the cluster's tidal radius, assuming the value of $r_t\sim16'$ reported by \citet{Odenkirchen:2002}.
-\todo{Update this value}.
-\citet{Vivas2006} also note that star 393, not in Group~6, could also be associated to the cluster's tails, given it's distance and location along the known extent of the tails.
-
-In summary,  previous studies report 7 known RRLs associated to the cluster itself (1 \typeab~and 6 \typec) plus 5 more reported as possibly associated to the tails (4 \typeab~and 2 \typec).
-
-\todo{update this with a comparison to Ibata+2017?}
+% In summary,  previous studies report 7 known RRLs associated to the cluster itself (1 \typeab~and 6 \typec) plus 5 more reported as possibly associated to the tails (4 \typeab~and 2 \typec).
 
@@
 
-\todo{what did we do}
+Using a catalog of RR Lyrae-type stars identified using multi-epoch, multi-band photometry from the Pan-STARRS PS1 survey and cross-matched to \Gaia\ \DR{2}, we XX.
 
@@
 
+\begin{table*}[t]
+\caption{RRL members of the Pal 5 stream.}\label{t:rrl_members}
+\begin{footnotesize}
+\begin{tabular}{llllllll}
+\toprule
+\Gaia~DR2               & Other & Period & Amp-$g$ & Type & D & member & sep  \\
+\verb+source_id+        & ID   & (d)    & (mag)   &     &(kpc)& prob & (") \\
+\midrule
+4418920808077110784 &  V5 & 0.337934& 0.46 & RRc  & 20.73 & 0.9982 &   1.9 \\
+4418920846732620032 &     & 0.643239& 0.27 & RRab & 21.71 & 0.9963 &   2.0 \\
+4418914863842345856 &  V1 & 0.293229& 0.56 & RRc  & 20.36 & 0.9696 &   2.0 \\
+4418726027016125056 &  V3 & 0.329948& 0.51 & RRc  & 20.65 & 0.9978 &   3.8 \\
+4418725889577171328 &  V4 & 0.286366& 0.59 & RRc  & 20.28 & 0.9915 &   4.5 \\
+4418731829516302848 &     & 0.28164 & 0.3  & RRc  & 18.14 & 0.9297 &   4.8 \\
+4418913218870688768 &  V2 & 0.332473& 0.4  & RRc  & 20.33 & 0.9977 &   5.0 \\
+4418734165978521728 &     & 0.380859& 0.52 & RRc  & 22.75 & 0.9850 &   7.6 \\
+4418724034151291776 & 403 & 0.551705& 1.14 & RRab & 21.23 & 0.9987 &   9.7 \\
+4418732791593809152 & 405 & 0.329670& 0.45 & RRc  & 20.74 & 0.9959 &  11.0 \\
+4419052204012341760 &     & 0.530006& 1.01 & RRab & 20.84 & 0.9960 &  44.5 \\
+4418142117622280192 & 393 & 0.462671& 1.36 & RRab & 19.95 & 0.9060 & 118.7 \\
+6339498589346000768 &     & 0.320110& 0.46 & RRc  & 18.89 & 0.9934 & 217.7 \\
+6339499379619987200 &     & 0.329577& 0.47 & RRc  & 19.35 & 0.9966 & 218.6 \\
+6339478312804685824 &     & 0.465883& 1.12 & RRab & 19.91 & 0.9970 & 242.8 \\
+4421078432143578496 &     & 0.520749& 1.11 & RRab & 21.98 & 0.9981 & 246.1 \\
+6339398155830632448 &     & 0.688316& 0.75 & RRab & 19.82 & 0.9982 & 258.9 \\
+4427253907921168000 &     & 0.496334& 1.35 & RRab & 21.81 & 0.9905 & 362.5 \\
+4427220338456828416 &     & 0.600796& 0.81 & RRab & 21.98 & 0.9974 & 384.4 \\
+4427234700827469952 &     & 0.517911& 1.24 & RRab & 21.94 & 0.9979 & 392.3 \\
+6337302795905934336 &     & 0.519462& 1.02 & RRab & 17.29 & 0.9808 & 404.5 \\
+6337233350579111680 &     & 0.542383& 0.96 & RRab & 18.98 & 0.9962 & 450.6 \\
+4427646704153765632 &     & 0.385552& 0.37 & RRc  & 19.90 & 0.9429 & 456.5 \\
+4424705647988505600 &     & 0.335502& 0.41 & RRc  & 20.87 & 0.9832 & 513.0 \\
+4426221707021159424 &     & 0.533313& 1.14 & RRab & 21.97 & 0.9728 & 550.1 \\
+\bottomrule
+\end{tabular}
+\end{footnotesize}
+\end{table*}
+
 \bibliographystyle{aasjournal}

### 2019-08-22 paper/main.tex
Updates from Overleaf

@@
 Thin stellar streams, formed from the tidal disruption of globular clusters, are important gravitational tools, sensitive to both global and small-scale properties of dark matter.
-The Palomar 5 stellar stream (Pal 5) is an exemplar stream within the Milky Way: Its $\sim 20\degr$ tidal tails connect back to the progenitor cluster, and the stream has been used to study the shape and total mass of, along with substructure in, the dark matter distribution of the Galaxy.
-%SP the above sentence is a bit long
+The Palomar 5 stellar stream (Pal 5) is an exemplar stream within the Milky Way: Its $\sim 20\degr$ tidal tails connect back to the progenitor cluster, and the stream has been used to study the shape, total mass, and substructure fraction of the dark matter distribution of the Galaxy.
 However, most details of the phase-space distribution of the stream are not fully explained, and dynamical models that use the stream for other inferences are therefore incomplete.
@@
 Conversely, of the $\sim$150 known globular clusters, only $\lesssim 20\%$ have purported tidal tails \citep[e.g.,][]{Leon:2000, Kundu:2019}, but even these are mostly low density and tenuous \citep[as might be expected, e.g.,][]{Balbinot:2018}.
+%SP The above sentence makes it sound like ~<20% of the 150 known GCs have streams, is it really that many?
 A prominent exception to these statements is the Palomar 5 (Pal 5) globular cluster and stream.
@@
 The Pal 5 stream---at a heliocentric distance of $d \sim 20~\kpc$---was discovered using early SDSS imaging, which reached well below the main sequence turnoff of the stream over a large area surrounding the cluster \citep{Odenkirchen:2001, Rockosi:2002}.
+%SP: Maybe cite Dotter 2011 for previous distance? 
 Since then, radial velocities have been measured for a handful of giant branch stars in the cluster and along the stream \citep{Odenkirchen:2002, Odenkirchen:2009, Ibata:2017}, and deeper imaging data has been used to map the tails at higher signal-to-noise \citep{Bernard:2016, Ibata:2016, Bonaca:2019}.
@@
 For example, using the sky track and radial velocity information, the stream has been used to measure the enclosed mass of the Milky Way \citep{Kuepper:2015, Bovy:2016}.
+%SP: if you say the "shape and mass", you can cite Pearson 15 too ;)
 Using deep photometry, the density variations along the stream have been used to place limits on the abundance of massive substructures in the Galactic halo \citep{Erkal:2017}.
@@
 For example, the leading and trailing tails have significantly different total star counts and lengths \citep{Dehnen:2004, Bernard:2016}, and the stream track, width, and density vary significantly over the full extent of the stream \citep{Ibata:2016, Bonaca:2019}.
-Given that Pal 5's pericenter is well within the Galactic disk ($r_{\textrm{peri}} \sim 7$--$8~\kpc$), these features could be a sign of perturbations from molecular clouds \citep[e.g.,][]{Amorisco:2016}, interaction with the Galactic bar \citep[e.g.,][]{Pearson:2017}, signatures of past dark matter subhalo encounters \citep[e.g.,][]{Erkal:2017}, or some combination of all of these effects.
+%SP: Might be work stating how big the estimated mass difference is in the trailing, leading vs cluster, as many people seem to be curious about this.
+Given that Pal 5's pericenter is well within the Galactic disk ($r_{\textrm{peri}} \sim 7$--$8~\kpc$), 
+%SP: cite erkal 2017 here (his fig 
+these features could be a sign of perturbations from molecular clouds \citep[e.g.,][]{Amorisco:2016}, interaction with the Galactic bar \citep[e.g.,][]{Pearson:2017}, signatures of past dark matter subhalo encounters \citep[e.g.,][]{Erkal:2017}, or some combination of all of these effects.
 Mapping the stream in all six phase-space dimensions---i.e., combining deep photometry with radial velocity, distance, and proper motion measurements---will enable and motivate improved dynamical models of the stream and more precise constraints on the Galactic mass distribution \citep[e.g.,][]{PriceWhelan:2013}.
@@
 To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu:2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters, here including the RR\typec~stars.
-The $\log{N_{RR}-M_V$ has quite a large scatter so, particularly for such low numbers of RRLs, it will only allow us to give rough limits on the cluster's absolute magnitude.
-Correcting the observed number of 26~RRLs by the median completeness of 92\%, we get a total expected number of 28~RRLs, for which we find an absolute magnitude $M_V=-8.1\pm 1.1$, which correspond to the mode and 16/84th percentiles of the posterior probability PDF.
+The $\log{N_{RR}}-M_V$ relation has a large scatter so, particularly for such low numbers of RRLs, it will only allow us to give rough limits on the cluster's absolute magnitude.
+Correcting the observed number of 26~RRLs by the median completeness of 92\%, we get a total expected number of 28~RRLs, for which we find an absolute magnitude $M_V=-8.1\pm 1.1$, which correspond to the mode and 16/84th percentiles of the posterior probability distribution.
 We therefore find that the initial mass of Pal 5 \todo{...}.
@@
 % The first two stars --403 and 405-- are located at distances of  $9\farcm8$ and $11'$ from the cluster's center, which \citet{Vivas2004} notes places them within the cluster's tidal radius, assuming the value of $r_t\sim16'$ reported by \citet{Odenkirchen:2002}.
-Two of these stars (IDs 403 and 405) are separated by just $\sim10'$ and $11'$ from the cluster's center, which places them within the cluster's Jacobi radius \citep{Vivas:2004}, and an additional star (star 393) outside of the Jacobi radius but at a similar distance could be associated with the stream \citep{Vivas:2006}.
+Two of these stars (IDs 403 and 405) are separated by just $\sim10'$ and $11'$ from the cluster's center, which places them within the cluster's Jacobi radius, and an additional star (star 393) outside of the Jacobi radius but at a similar distance could be associated with the stream \citep{Vivas:2006}.
 More recently, \citet{Ibata:2017} use a prior iteration of the PS1 RRL catalog \citep{Hernitschek:2016} to identify \todo{XX} candidate members of the Pal 5 cluster and tails based on sky position and distance alone.
@@
 
-Using a catalog of RR Lyrae-type stars identified using multi-epoch, multi-band photometry from the Pan-STARRS PS1 survey and cross-matched to \Gaia\ \DR{2}, we XX.
+Using a catalog of RR Lyrae stars identified using multi-epoch, multi-band photometry from the Pan-STARRS PS1 survey and cross-matched to \Gaia\ \DR{2}, we construct a probabilistic model to determine membership with the Pal 5 cluster and stream.
+We find 24 RRLs with membership probability $>0.5$ (13 type~\typeab\ and 11 type~\typec) over $\approx 18\degr$ of the stream.
+From these stars, we infer the distance to the cluster, finding a closer value, $d = 20.9 \pm 0.2~\kpc$, than many past studies, and detect a distance gradient of $\sim 0.4~\kpc~{\rm deg}^{-1}$ between the trailing and leading tail (between $-5^\circ < \phi_1 < 5^\circ$).
+We also, for the first time, detect trends in proper motion along the stream.
+We release the full RRL catalog in the region around Pal 5 (containing distance information from PS1 and astrometry from \Gaia) with our derived membership probabilities.
+These data will be instrumental for improving dynamical models of Pal 5, both for using Pal 5 to constrain the global matter distribution of the Milky Way and in interpreting the observed, complex density structure of the stream.
 
@@
 CM acknowledges funding from the MIA program at Universidad de la Republica, Uruguay, and is grateful for the hospitality and support of the Flatiron Institute where part of this research was carried out.
+This work has made use of data from the European Space Agency (ESA) mission
+{\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed by the {\it Gaia}
+Data Processing and Analysis Consortium (DPAC,
+\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding for the DPAC
+has been provided by national institutions, in particular the institutions
+participating in the {\it Gaia} Multilateral Agreement.
 

### 2019-08-22 paper/main.tex
Updates from Overleaf

@@
 We use a large catalog of RR Lyrae-type stars (RRLs) with astrometric data from the \Gaia\ mission to probabilistically identify RRLs in the Pal 5 stream.
-%SP throughout the manuscript, check if you mean plural when using RRL vs RRLs.
+\sa{throughout the manuscript, check if you mean plural when using RRL vs RRLs.}
 RRLs are useful because they are intrinsically-luminous standard candles and their distances can be inferred with small relative precision ($\sim3\%$).
@@
 Conversely, of the $\sim$150 known globular clusters, only $\lesssim 20\%$ have purported tidal tails \citep[e.g.,][]{Leon:2000, Kundu:2019}, but even these are mostly low density and tenuous \citep[as might be expected, e.g.,][]{Balbinot:2018}.
-%SP The above sentence makes it sound like ~<20% of the 150 known GCs have streams, is it really that many?
+\sa{The above sentence makes it sound like $\lesssim 20\%$ of the 150 known GCs have streams, is it really that many?}
 A prominent exception to these statements is the Palomar 5 (Pal 5) globular cluster and stream.
@@
 The Pal 5 stream---at a heliocentric distance of $d \sim 20~\kpc$---was discovered using early SDSS imaging, which reached well below the main sequence turnoff of the stream over a large area surrounding the cluster \citep{Odenkirchen:2001, Rockosi:2002}.
-%SP: Maybe cite Dotter 2011 for previous distance? 
+\sa{Maybe cite Dotter 2011 for previous distance? }
 Since then, radial velocities have been measured for a handful of giant branch stars in the cluster and along the stream \citep{Odenkirchen:2002, Odenkirchen:2009, Ibata:2017}, and deeper imaging data has been used to map the tails at higher signal-to-noise \citep{Bernard:2016, Ibata:2016, Bonaca:2019}.
@@
 For example, using the sky track and radial velocity information, the stream has been used to measure the enclosed mass of the Milky Way \citep{Kuepper:2015, Bovy:2016}.
-%SP: if you say the "shape and mass", you can cite Pearson 15 too ;)
+\sa{if you say the "shape and mass", you can cite Pearson 15 too ;)}
 Using deep photometry, the density variations along the stream have been used to place limits on the abundance of massive substructures in the Galactic halo \citep{Erkal:2017}.
@@
 For example, the leading and trailing tails have significantly different total star counts and lengths \citep{Dehnen:2004, Bernard:2016}, and the stream track, width, and density vary significantly over the full extent of the stream \citep{Ibata:2016, Bonaca:2019}.
-%SP: Might be work stating how big the estimated mass difference is in the trailing, leading vs cluster, as many people seem to be curious about this.
+\sa{Might be work stating how big the estimated mass difference is in the trailing, leading vs cluster, as many people seem to be curious about this.}
 Given that Pal 5's pericenter is well within the Galactic disk ($r_{\textrm{peri}} \sim 7$--$8~\kpc$), 
-%SP: cite erkal 2017 here (his fig 
+\sa{cite Erkal 2017 here (his fig 6)?}
 these features could be a sign of perturbations from molecular clouds \citep[e.g.,][]{Amorisco:2016}, interaction with the Galactic bar \citep[e.g.,][]{Pearson:2017}, signatures of past dark matter subhalo encounters \citep[e.g.,][]{Erkal:2017}, or some combination of all of these effects.
 Mapping the stream in all six phase-space dimensions---i.e., combining deep photometry with radial velocity, distance, and proper motion measurements---will enable and motivate improved dynamical models of the stream and more precise constraints on the Galactic mass distribution \citep[e.g.,][]{PriceWhelan:2013}.
-However, at its present distance, main sequence stars are barely detected and have large astrometric uncertainties in \Gaia\ \DR{2}, and other stellar tracers (e.g., giant branch stars) are not numerous enough or precise enough distance indicators to provide useful relative distance information along the stream.
+However, at its present distance, main sequence stars are barely detected and have large astrometric uncertainties in \Gaia\ \DR{2}, 
+\sa{The above makes it sound like the ms stars are barely detected in general, maybe flip sentence so it's clear that *in Gaia*....}
+and other stellar tracers (e.g., giant branch stars) are not numerous enough or precise enough distance indicators to provide useful relative distance information along the stream.
+\sa{Maybe here mention that a "velocity" gradient has been detected along the stream (Kuzma 2015)? }
 % Enter: RR Lyrae-type stars.
@@
 At its current distance, Pal~5's horizontal branch is bright enough ($\G\sim17.3$) to be observed by \Gaia.
-Therefore, combining \Gaia~DR2 proper motions with the newer PS1 RRL catalog from \citet{Sesar:2017b} --with improved classification probabilities and light curve periods--  will allows to make precise distance and proper motion measurements along the full extent of the Pal~5 stream.
+Therefore, combining \Gaia~DR2 proper motions with the newer PS1 RRL catalog from \citet{Sesar:2017b} --with improved classification probabilities and light curve periods--  will allow us to make precise distance and proper motion measurements along the full extent of the Pal~5 stream.
 Both, key ingredients for dynamical modelling of the Pal~5 stream...
@@
 \caption{\textbf{Left}: Proper motions of RRLs in the region $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$ with distances in the range $18 < d < 25~\kpc$.
-\textbf{Middle}: The same, but for RRLs within $1~\degr$ of the mean sky track of Pal 5 \citep{Bonaca:2019}.
+The distance selection here is purely illustrative: We do not use a distance cut in our membership model.
+% \sa{Might be worth looking at closer distances (leading arm in a MW potential comes as close as $\sim$15 kpc), especially if you want to find members at a dec of $-15\deg$.}
+% APW: My models only go to 18 at the edge of the sky footprint, but this distance cut is only used for illustrative purposes. I added a sentence to clarify.
+\textbf{Middle}: The same, but for RRLs within $1\degr$ of the mean sky track of Pal 5 \citep{Bonaca:2019}.
 The over-density of stars near $(-2.7, -2.7)$ is the stream, and the two colored points show the Pal 5 cluster proper motions from \citet{Vasiliev:2019} and \citet{Fritz:2015}, respectively.
-\textbf{Right}: The same, but for RRLs excluding $2~\degr$ around the Pal 5 sky track.}
+\textbf{Right}: The same, but for RRLs excluding $2\degr$ around the Pal 5 sky track.}
 \label{fig:pm}
@@
 \end{figure*}
-
-\todo{This part was rewritten to reflect the catalog we ended up using. CM.}
 
@@
 At present, however, these catalogs are subject to more significant and spatially-varying incompleteness in this particular area of the sky due to the \Gaia\ scanning law \citep[see][]{Rimoldini2018}.
+\sa{Mention if DR3 will help?}
 
@@
 \figurename~\ref{fig:pm} demonstrates that Pal 5 stream members appear in the RRL catalog: This figure shows the \Gaia\ \DR{2} proper motions for all RRLs in the sky window defined above within the distance range $18 < d < 25~\kpc$, then the same for RRLs in a tight selection around the observed stream track, and then finally the same for RRLs \emph{excluding} the stream region.
+\sa{Do we want to show this stream track?
+Or state how wide it is?
+(it's inthe  figure, but mention here too).}
 
@@
 Using the \Gaia~DR2 VariClassifier \citet{Holl2018,Rimoldini2018} and Specific Objects Study \citep{Clementini2018} catalogs, we find a median completeness of 92\% with a standard deviation of 16\%  with no evident spatial trend across the field of Figure~\ref{fig:members}.
+\sa{I got a little confused in the section above as Fig 3 is referenced but Fig 2 never mentioned. }
 
@@
 In order to search for new RRL stars in the Pal 5 stellar stream, we construct a probabilistic model of the stream and background stellar distribution (simultaneously) using proper motion $(\mu_\alpha, \mu_\delta)$ and distance $d$ data for all RRL in the sky window $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$, where $(\alpha, \delta)$ are right ascension and declination.
+\sa{Is it worth it to show the whole region in fig 3, and/or maybe extend to even more negative $\delta $?}
 We \emph{do not} include the known spatial extent or morphology of the Pal 5 stream in our membership model for the stream: We instead later use the sky positions to validate the modeling procedure and selection criteria we then use to define a sample of probable Pal 5 stream RRLs.
@@
 We use \project{extreme deconvolution} \citep[\acronym{XD};][]{Bovy:XD} to optimize the parameters of a Gaussian Mixture Model (GMM) representation of the (deconvolved) background RRL density.
+\sa{minor: three sentences in a row start with "we use".}
 We first fit the background model (with \acronym{XD}) using between $3$--$12$ mixture components and evaluate the Bayesian Information Criterion (BIC) for the optimal parameters in each case.
@@
 \caption{RRL stars in the vicinity of the Pal 5 stream (with $\textrm{prob.} > 0.01$), colored by membership probability, and under-plotted with the inferred stream trends used in the membership model.
+\sa{Could be worth emphasizing here too that the probability does not depend on positional info (i.e.
+being close to the stream itself) for people only skimming through figures.}
 \textbf{Upper left}: Sky positions (in the Pal 5 stream coordinate frame) of RRLs, with arrows showing the \Gaia\ \DR{2} proper motion vector direction.
@@
 \includegraphics[width=\textwidth]{members.pdf}
-\caption{Sky positions of RR Lyrae stars with Pal 5 membership probabilities $>0.01$, colored by membership probability (large markers).
-The background greyscale shows the surface density of CMD-filtered main sequence stars from \citet{Bonaca:2019}.}
+\caption{Sky positions of RR Lyra stars with Pal 5 membership probabilities $>0.01$, colored by membership probability (large \sa{circular?} markers).
+The background greyscale shows the surface density of CMD-filtered main sequence stars from \citet{Bonaca:2019}.
+\sa{specify that RA/dec are in degrees?}}
 \label{fig:members}
@@
 
-\figurename~\ref{fig:trackmembers} shows a summary of the inferred membership model properties in sky position (shown in Pal 5 coordinates $\phi_1, \phi_2$), distance, and proper motion components.
-\tablename~\ref{t:rrl_members} shows an abbreviated version of our catalog of RRL in the region around Pal 5, with PS1 and \Gaia data and membership probabilities appended.
+\figurename~\ref{fig:trackmembers} shows a summary of the inferred membership model properties in sky position (shown in Pal 5 coordinates $\phi_1, \phi_2$) \sa{might be worth emphasizing again that this is along the stream of pal 5 and references Ana's figure (people aren't used to seeing the stream in this coord system.}, distance, and proper motion components.
+\tablename~\ref{t:rrl_members} shows an abbreviated version of our catalog of RRL in the region around Pal 5, with PS1 and \Gaia~ data and membership probabilities appended.
 
@@
 
-\subsection{RR Lyrae stars associated with Pal 5 and its tails}
-
+\subsection{RR Lyra stars associated with Pal 5 and its tails}
+%SP: changed RR Lyrae stars to RR Lyra stars a few places, maybe check whole manuscript (otherwise plural twice). 
 We find a total of 24 RR Lyrae stars with membership probability $> 0.5$ that lie within $1\degr$ of the stream track \citep[again using the track from][]{Bonaca:2019}.
@@
 No theoretical models of Pal 5 have predicted tidal debris at such locations, so follow-up spectroscopy to confirm or rule out their origin would be very informative.
+\sa{What about RRLs at more negative $\delta$, wasn't your region larger than the figure extent?}
 
@@
 
-Past distance measurements to the Pal 5 cluster have ranged from $20.9~\kpc$ \citep{Dotter:2011} to $23.2~\kpc$ \citep{Harris:1996}, with a previous measurement from a small number of RR Lyrae stars in between at $22.3~\kpc$ \citep{Vivas:2006}.
-We find a mean Pal 5 cluster distance of $20.9 \pm 0.2~\kpc$ using the PS1 RRLs and the membership model described in \sectionname~\ref{sec:membership}
+Past distance measurements to the Pal 5 cluster have ranged from $20.9~\kpc$ \citep{Dotter:2011} \sa{worth mentioning with which technique?} to $23.2~\kpc$ \citep{Harris:1996}, with a previous measurement from a small number of RR Lyrae stars in between at $22.3~\kpc$ \citep{Vivas:2006}.
+We find a mean Pal 5 cluster distance of $20.9 \pm 0.2~\kpc$ using the PS1 RRLs and the membership model described in \sectionname~\ref{sec:membership}.
 This is closer than most past measurements of the cluster distance, but consistent with the measurement from \citet{Dotter:2011}.
 We find a proper motion $(\mu_\alpha\,\cos\delta, \mu_\delta) = (-2.69, -2.70) \pm (0.04, 0.03)~\masyr$, consistent with the measurement in \citet{Vasiliev:2019}.
+\sa{state that this is larger than most before next sentence?}
 The larger proper motion (relative to \citealt{Fritz:2015} and other earlier measurements) and the closer distance means that the tangential velocity of Pal 5 is $\approx 30~\kms$ larger than most previous models have assumed.
+\sa{which will be crucial/important for future modeling, migth change inferred inclosed mass of kuepper 2015 etc etc.
+More a discussion point I guess, will read.}
 
@@
 While the proper motion in right ascension appears consistent with constant or only a mild gradient over the extent of the stream, we see a significant gradient in the declination component, with a difference of about $\approx 1~\masyr$ from trailing to leading tail.
+\sa{worth mentioning Pearson 2017, fig 2 where this same trend is also seen (flatter gradient in RA proper motion than in dec proper motion?}
 % \todo{Make a table with summary information?}
@@
 
-Assuming a Jacobi radius of $12^\prime$ (computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.4\times10^4~\msun$ and using the Milky Way mass model in \citealt{gala}), we find 15 RRLs in the cluster tails, 11 classified as type~\typeab\ and 4 type~\typec.
+Assuming a Jacobi radius of $12^\prime$ (computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.4\times10^4~\msun$ \sa{ref for mass?} and using the Milky Way mass model in \citealt{gala}), we find 15 RRLs in the cluster tails, 11 classified as type~\typeab\ and 4 type~\typec.
 Combined with the 2 type~\typeab\ and 7 type~\typec\ RRLs in the cluster itself, there are 24 RRLs in total associated to Pal~5 and its tails with membership probabilities larger than 0.5.
@@
 The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram.
-The distribution of the \typeab~RRL clearly indicate Pal~5 to be an Oosterhoff I cluster, consistent with their mean period of $0.565~d$ \todo{(XX) ?}.
+The distribution of the \typeab~RRL clearly indicate that Pal~5 is an Oosterhoff I cluster, consistent with having a mean period of $0.565~d$ \todo{(XX) ?}.
 Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an OoI classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.32$ \citep{TODO}.
@@
 \todo{CM:( any other point to be made about this?)}.
+\sa{as someone who doesn't know much about RRLs, how weird is the above?
+maybe some more discussion/elaboration?
+}
 
@@
 \includegraphics[width=\textwidth]{rrls_PA.pdf}
-\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field.}
+\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field. \sa{some statement on what to take away from the figure... i.e. consistent with Pal 5 being an "x" type GC....}}
 \label{fig:PA_diagram}
@@
 
-The Pal~5 cluster has 5 previous known RRL stars (named V1-V5), all of type \typec, as reported in the \citet{Clement:2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{SawyerHogg:1973}.
+The Pal~5 cluster has 5 previous known RRL stars (named V1--V5), all of type which are \typec, as reported in the \citet{Clement:2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{SawyerHogg:1973}.
 In their analysis of the halo density profile with RRLs, \citet{Vivas:2006} report an overdensity (``Group~6'') that they claim could be associated with Pal~5's tidal tails.
@@
 We find 24 RRLs with membership probability $>0.5$ (13 type~\typeab\ and 11 type~\typec) over $\approx 18\degr$ of the stream.
-From these stars, we infer the distance to the cluster, finding a closer value, $d = 20.9 \pm 0.2~\kpc$, than many past studies, and detect a distance gradient of $\sim 0.4~\kpc~{\rm deg}^{-1}$ between the trailing and leading tail (between $-5^\circ < \phi_1 < 5^\circ$).
+From these stars \sa{you mean only the cluster stars, right?}, we infer the distance to the cluster, finding a closer value, $d = 20.9 \pm 0.2~\kpc$, than many past studies, and detect a distance gradient of $\sim 0.4~\kpc~{\rm deg}^{-1}$ between the trailing and leading tail (between $-5^\circ < \phi_1 < 5^\circ$).
 We also, for the first time, detect trends in proper motion along the stream. 
+\sa{which has been suggested in/reported in/consistent with modeling of the stream...? fig 2, pearson 17}
 We release the full RRL catalog in the region around Pal 5 (containing distance information from PS1 and astrometry from \Gaia) with our derived membership probabilities.
 These data will be instrumental for improving dynamical models of Pal 5, both for using Pal 5 to constrain the global matter distribution of the Milky Way and in interpreting the observed, complex density structure of the stream.
+\sa{not sure where in text, but could be worth mentioning that you don't find stars beyond the current footprint of the stream (i.e.
+that leading stream does appear to be shorter than trailing..., although looking at fig 3, you actually don't find trailin stream RRLs much further from the prog.
+than you find leading arm RRLs.}
 

### 2019-08-23 paper/main.tex
respond to Sa

--- 2019-08-22 paper/main.tex
@@
 We use a large catalog of RR Lyrae-type stars (RRLs) with astrometric data from the \Gaia\ mission to probabilistically identify RRLs in the Pal 5 stream.
-%SP throughout the manuscript, check if you mean plural when using RRL vs RRLs.
 RRLs are useful because they are intrinsically-luminous standard candles and their distances can be inferred with small relative precision ($\sim3\%$).
@@
 Conversely, of the $\sim$150 known globular clusters, only $\lesssim 20\%$ have purported tidal tails \citep[e.g.,][]{Leon:2000, Kundu:2019}, but even these are mostly low density and tenuous \citep[as might be expected, e.g.,][]{Balbinot:2018}.
-%SP The above sentence makes it sound like ~<20% of the 150 known GCs have streams, is it really that many?
+% \sa{The above sentence makes it sound like $\lesssim 20\%$ of the 150 known GCs have streams, is it really that many?}
+% APW: Yep! According to the citations.
 A prominent exception to these statements is the Palomar 5 (Pal 5) globular cluster and stream.
@@
 The Pal 5 stream---at a heliocentric distance of $d \sim 20~\kpc$---was discovered using early SDSS imaging, which reached well below the main sequence turnoff of the stream over a large area surrounding the cluster \citep{Odenkirchen:2001, Rockosi:2002}.
-%SP: Maybe cite Dotter 2011 for previous distance? 
+% \sa{Maybe cite Dotter 2011 for previous distance? }
+% APW: It's just a qualitative statement to give a rough sense of its distance, so I'm not really quoting a measurement
 Since then, radial velocities have been measured for a handful of giant branch stars in the cluster and along the stream \citep{Odenkirchen:2002, Odenkirchen:2009, Ibata:2017}, and deeper imaging data has been used to map the tails at higher signal-to-noise \citep{Bernard:2016, Ibata:2016, Bonaca:2019}.
@@
 For example, using the sky track and radial velocity information, the stream has been used to measure the enclosed mass of the Milky Way \citep{Kuepper:2015, Bovy:2016}.
-%SP: if you say the "shape and mass", you can cite Pearson 15 too ;)
+% \sa{if you say the "shape and mass", you can cite Pearson 15 too ;)}
+% APW: I say "measure" not "rule out" :) - but I'll reassess when I do the shortening and see if it fits
 Using deep photometry, the density variations along the stream have been used to place limits on the abundance of massive substructures in the Galactic halo \citep{Erkal:2017}.
@@
 However, a number of peculiar aspects of the stream still remain unexplained in detail.
-For example, the leading and trailing tails have significantly different total star counts and lengths \citep{Dehnen:2004, Bernard:2016}, and the stream track, width, and density vary significantly over the full extent of the stream \citep{Ibata:2016, Bonaca:2019}.
-%SP: Might be work stating how big the estimated mass difference is in the trailing, leading vs cluster, as many people seem to be curious about this.
+For example, the leading and trailing tails have significantly different total star counts and lengths \citep[both different by a factor of $\sim$2 between leading and trailing;][]{Dehnen:2004, Bernard:2016}, and the stream track, width, and density vary significantly over the full extent of the stream \citep{Ibata:2016, Bonaca:2019}.
+% \sa{Might be work stating how big the estimated mass difference is in the trailing, leading vs cluster, as many people seem to be curious about this.}
+% APW: Added a note above
 Given that Pal 5's pericenter is well within the Galactic disk ($r_{\textrm{peri}} \sim 7$--$8~\kpc$),
-%SP: cite erkal 2017 here (his fig 
+% \sa{cite Erkal 2017 here (his fig 6)?}
+% APW: Again just a rough / qualtitative number to give a sense - I don't really see the need to cite given that this is from my own exploration with the orbit fits!
 these features could be a sign of perturbations from molecular clouds \citep[e.g.,][]{Amorisco:2016}, interaction with the Galactic bar \citep[e.g.,][]{Pearson:2017}, signatures of past dark matter subhalo encounters \citep[e.g.,][]{Erkal:2017}, or some combination of all of these effects.
 Mapping the stream in all six phase-space dimensions---i.e., combining deep photometry with radial velocity, distance, and proper motion measurements---will enable and motivate improved dynamical models of the stream and more precise constraints on the Galactic mass distribution \citep[e.g.,][]{PriceWhelan:2013}.
-However, at its present distance, main sequence stars are barely detected and have large astrometric uncertainties in \Gaia\ \DR{2}, and other stellar tracers (e.g., giant branch stars) are not numerous enough or precise enough distance indicators to provide useful relative distance information along the stream.
+However, with \Gaia\ \DR{2}, at its present distance, main sequence stars are barely detected and have large astrometric uncertainties.
+Other stellar tracers (e.g., giant branch stars) are not numerous enough or precise enough distance indicators to provide useful relative distance information along the stream.
+% \sa{Maybe here mention that a "velocity" gradient has been detected along the stream (Kuzma 2015)? }
+% APW: Good idea, but then it would make sense to talk about the RV data in Ibata+2017. I'll try to fit both into the discussion.
 % Enter: RR Lyrae-type stars.
@@
 At its current distance, Pal~5's horizontal branch is bright enough ($\G\sim17.3$) to be observed by \Gaia.
-Therefore, combining \Gaia~DR2 proper motions with the newer PS1 RRL catalog from \citet{Sesar:2017b} --with improved classification probabilities and light curve periods--  will allows to make precise distance and proper motion measurements along the full extent of the Pal~5 stream.
+Therefore, combining \Gaia~DR2 proper motions with the newer PS1 RRL catalog from \citet{Sesar:2017b} --with improved classification probabilities and light curve periods--  will allow us to make precise distance and proper motion measurements along the full extent of the Pal~5 stream.
 Both, key ingredients for dynamical modelling of the Pal~5 stream...
 %meaning we can measure kinematic properties of the full stream...
-\citep{Ibata:2017} found some RRL in pal 5, but no proper motions.
+\citep{Ibata:2017} found some RRLs in pal 5, but no proper motions.
 This is useful for modeling...
-Many stream candidates found in RRL alone \citep{Mateu:2018}.
+Many stream candidates found in RRLs alone \citep{Mateu:2018}.
 
@@
 \caption{\textbf{Left}: Proper motions of RRLs in the region $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$ with distances in the range $18 < d < 25~\kpc$.
-\textbf{Middle}: The same, but for RRLs within $1~\degr$ of the mean sky track of Pal 5 \citep{Bonaca:2019}.
+The distance selection here is purely illustrative: We do not use a distance cut in our membership model.
+% \sa{Might be worth looking at closer distances (leading arm in a MW potential comes as close as $\sim$15 kpc), especially if you want to find members at a dec of $-15\deg$.}
+% APW: Hm, my models only go to ~18 at the edge of the sky footprint, but this distance cut is only used for illustrative purposes. I added a sentence to clarify.
+\textbf{Middle}: The same, but for RRLs within $1\degr$ of the mean sky track of Pal 5 \citep{Bonaca:2019}.
 The over-density of stars near $(-2.7, -2.7)$ is the stream, and the two colored points show the Pal 5 cluster proper motions from \citet{Vasiliev:2019} and \citet{Fritz:2015}, respectively.
-\textbf{Right}: The same, but for RRLs excluding $2~\degr$ around the Pal 5 sky track.}
+\textbf{Right}: The same, but for RRLs excluding $2\degr$ around the Pal 5 sky track.}
 \label{fig:pm}
@@
 Two all-sky RR Lyrae catalogs have also been released as part of \Gaia~\DR{2} \citep[VariClassifier and Specific Object Studies][]{Holl2018, Rimoldini2018, Clementini2018}.
-At present, however, these catalogs are subject to more significant and spatially-varying incompleteness in this particular area of the sky due to the \Gaia\ scanning law \citep[see][]{Rimoldini2018}.
+At present, however, these catalogs are subject to more significant and spatially-varying incompleteness in this particular area of the sky due to the \Gaia\ scanning law \citep[see][, though many of these issues will go away with future \Gaia\ data releases]{Rimoldini2018}.
+% \sa{Mention if DR3 will help?}
+% APW: done!
 
@@
 We cross-match the full PS1 RRL catalog with \Gaia~DR2 \citep{Gaia:2018}, using a 3" sky separation tolerance, in order to retrieve astrometric information for these stars.
-Although not all PS1 RRLs have been identified as RRLs from \Gaia\ photometry, all of the PS1 RRL are present in the main point source catalog.
-\figurename~\ref{fig:pm} demonstrates that Pal 5 stream members appear in the RRL catalog: This figure shows the \Gaia\ \DR{2} proper motions for all RRLs in the sky window defined above within the distance range $18 < d < 25~\kpc$, then the same for RRLs in a tight selection around the observed stream track, and then finally the same for RRLs \emph{excluding} the stream region.
+Although not all PS1 RRLs have been identified as RRLs from \Gaia\ photometry, all of the PS1 RRLs are present in the main point source catalog.
+\figurename~\ref{fig:pm} demonstrates that Pal 5 stream members appear in the RRL catalog: This figure shows the \Gaia\ \DR{2} proper motions for all RRLs in the sky window defined above within the distance range $18 < d < 25~\kpc$, then the same for RRLs in a tight selection around the observed stream track (within $1\degr$ of the stream track), and then finally the same for RRLs \emph{excluding} the stream region (excluding stars within $2\degr$ of the stream track).
+% \sa{Do we want to show this stream track? Or state how wide it is? (it's inthe  figure, but mention here too).}
+% APW: Noted.
 
@@
 We can also produce an estimate per line of sight using the methodology described in \citet{Rybizki2018}, which requires two independent catalogs to assess the completeness of both in a probabilistic manner.
-Using the \Gaia~DR2 VariClassifier \citet{Holl2018,Rimoldini2018} and Specific Objects Study \citep{Clementini2018} catalogs, we find a median completeness of 92\% with a standard deviation of 16\%  with no evident spatial trend across the field of Figure~\ref{fig:members}.
+Using the \Gaia~DR2 VariClassifier \citet{Holl2018,Rimoldini2018} and Specific Objects Study \citep{Clementini2018} catalogs, we find a median completeness of 92\% with a standard deviation of 16\% with no evident spatial trend across the sky selection window.
+% \sa{I got a little confused in the section above as Fig 3 is referenced but Fig 2 never mentioned. }
+% APW: I moved this paragraph up from the discussion and forgot to edit - thanks!
 
@@
 
-In order to search for new RRL stars in the Pal 5 stellar stream, we construct a probabilistic model of the stream and background stellar distribution (simultaneously) using proper motion $(\mu_\alpha, \mu_\delta)$ and distance $d$ data for all RRL in the sky window $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$, where $(\alpha, \delta)$ are right ascension and declination.
+In order to search for new RRL stars in the Pal 5 stellar stream, we construct a probabilistic model of the stream and background stellar distribution (simultaneously) using proper motion $(\mu_\alpha, \mu_\delta)$ and distance $d$ data for all RRLs in the sky window $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$, where $(\alpha, \delta)$ are right ascension and declination.
+% \sa{Is it worth it to show the whole region in fig 3, and/or maybe extend to even more negative $\delta $?}
+% APW: I don't think so? And what do you mean by the "whole region"?
 We \emph{do not} include the known spatial extent or morphology of the Pal 5 stream in our membership model for the stream: We instead later use the sky positions to validate the modeling procedure and selection criteria we then use to define a sample of probable Pal 5 stream RRLs.
@@
 We use RRL stars in the background region to construct an error-deconvolved density model in the 3D space of proper motion components and distance.
-We use proper motions and uncertainties from \Gaia---using the provided covariance matrix between proper motion components---and distances and uncertainties as derived above (\sectionname~\ref{sec:data}).
-We use \project{extreme deconvolution} \citep[\acronym{XD};][]{Bovy:XD} to optimize the parameters of a Gaussian Mixture Model (GMM) representation of the (deconvolved) background RRL density.
+We use proper motions and uncertainties from \Gaia---using the provided covariance matrix between proper motion components---and distances and uncertainties from the PS1 RRL catalog (\sectionname~\ref{sec:data}).
+We optimize the parameters of a Gaussian Mixture Model (GMM) representation of the (deconvolved) background RRL density using \project{extreme deconvolution} \citep[\acronym{XD};][]{Bovy:XD}.
+% \sa{minor: three sentences in a row start with "we use".}
+% APW: changed one of them
 We first fit the background model (with \acronym{XD}) using between $3$--$12$ mixture components and evaluate the Bayesian Information Criterion (BIC) for the optimal parameters in each case.
@@
 \caption{RRL stars in the vicinity of the Pal 5 stream (with $\textrm{prob.} > 0.01$), colored by membership probability, and under-plotted with the inferred stream trends used in the membership model.
+We note again that the membership probability does not depend on position or proximity to the stream, and only on distance and proper motions.
+% \sa{Could be worth emphasizing here too that the probability does not depend on positional info (i.e. being close to the stream itself) for people only skimming through figures.}
+% APW: added!
 \textbf{Upper left}: Sky positions (in the Pal 5 stream coordinate frame) of RRLs, with arrows showing the \Gaia\ \DR{2} proper motion vector direction.
@@
 \includegraphics[width=\textwidth]{members.pdf}
-\caption{Sky positions of RR Lyrae stars with Pal 5 membership probabilities $>0.01$, colored by membership probability (large markers).
+\caption{Sky positions of RR Lyra stars with Pal 5 membership probabilities $>0.01$, colored by membership probability (large markers).
 The background greyscale shows the surface density of CMD-filtered main sequence stars from \citet{Bonaca:2019}.}
@@
 \figurename~\ref{fig:trackmembers} shows a summary of the inferred membership model properties in sky position (shown in Pal 5 coordinates $\phi_1, \phi_2$), distance, and proper motion components.
-\tablename~\ref{t:rrl_members} shows an abbreviated version of our catalog of RRL in the region around Pal 5, with PS1 and \Gaia data and membership probabilities appended.
+\tablename~\ref{t:rrl_members} shows an abbreviated version of our catalog of RRL in the region around Pal 5, with PS1 and \Gaia~ data and membership probabilities appended.
 
@@
 
-\subsection{RR Lyrae stars associated with Pal 5 and its tails}
-
+\subsection{RR Lyra stars associated with Pal 5 and its tails}
+%SP: changed RR Lyrae stars to RR Lyra stars a few places, maybe check whole manuscript (otherwise plural twice).
 We find a total of 24 RR Lyrae stars with membership probability $> 0.5$ that lie within $1\degr$ of the stream track \citep[again using the track from][]{Bonaca:2019}.
@@
 We find a total of 10 RRLs within the estimated Jacobi radius of the cluster ($\approx 12'$), meaning that 15 RRLs are spread between the leading and trailing tails.
-\figurename~\ref{fig:members} shows the sky positions of all RRL with membership probability $>0.01$ over-plotted on the density of filtered main sequence stars from \citet{Bonaca:2019}.
+\figurename~\ref{fig:members} shows the sky positions of all RRLs with membership probability $>0.01$ over-plotted on the density of filtered main sequence stars from \citet{Bonaca:2019}.
 We note that there are a few RRLs outside of the main stream track that appear to be at a consistent distance with consistent proper motions (e.g., the two stars near $\alpha = 230\degr$, $\delta = 0\degr$) that are not included in the tally above.
 No theoretical models of Pal 5 have predicted tidal debris at such locations, so follow-up spectroscopy to confirm or rule out their origin would be very informative.
+% \sa{What about RRLs at more negative $\delta$, wasn't your region larger than the figure extent?}
+% APW: There aren't any stars with prob > 0.01 below dec ~ -11 near the stream, so I trimmed the axis.
 
@@
 Past distance measurements to the Pal 5 cluster have ranged from $20.9~\kpc$ \citep{Dotter:2011} to $23.2~\kpc$ \citep{Harris:1996}, with a previous measurement from a small number of RR Lyrae stars in between at $22.3~\kpc$ \citep{Vivas:2006}.
-We find a mean Pal 5 cluster distance of $20.9 \pm 0.2~\kpc$ using the PS1 RRLs and the membership model described in \sectionname~\ref{sec:membership}
+% \sa{worth mentioning with which technique?}
+We find a mean Pal 5 cluster distance of $20.9 \pm 0.2~\kpc$ using the PS1 RRLs and the membership model described in \sectionname~\ref{sec:membership}.
 This is closer than most past measurements of the cluster distance, but consistent with the measurement from \citet{Dotter:2011}.
 We find a proper motion $(\mu_\alpha\,\cos\delta, \mu_\delta) = (-2.69, -2.70) \pm (0.04, 0.03)~\masyr$, consistent with the measurement in \citet{Vasiliev:2019}.
+% \sa{state that this is larger than most before next sentence?}
 The larger proper motion (relative to \citealt{Fritz:2015} and other earlier measurements) and the closer distance means that the tangential velocity of Pal 5 is $\approx 30~\kms$ larger than most previous models have assumed.
+This is a noteworthy update to consider for future stream modeling efforts and will likely change the inferred Milky Way parameters as compared to past models \citep[e.g.,][]{Kuepper:2015, Bovy:2016}.
 
@@
 While the proper motion in right ascension appears consistent with constant or only a mild gradient over the extent of the stream, we see a significant gradient in the declination component, with a difference of about $\approx 1~\masyr$ from trailing to leading tail.
-% \todo{Make a table with summary information?}
+% \sa{worth mentioning Pearson 2017, fig 2 where this same trend is also seen (flatter gradient in RA proper motion than in dec proper motion?}
+% APW: What were mu_1 and mu_2? Were those proper motions in RA/Dec?
 
@@
 
-Assuming a Jacobi radius of $12^\prime$ (computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.4\times10^4~\msun$ and using the Milky Way mass model in \citealt{gala}), we find 15 RRLs in the cluster tails, 11 classified as type~\typeab\ and 4 type~\typec.
-Combined with the 2 type~\typeab\ and 7 type~\typec\ RRLs in the cluster itself, there are 24 RRLs in total associated to Pal~5 and its tails with membership probabilities larger than 0.5.
+Assuming a Jacobi radius of $11'$ (computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.2\times10^4~\msun$ \citep{Kuepper:2015} and using the Milky Way mass model in \citealt{gala}), we find 16 RRLs in the cluster tails, 11 classified as type~\typeab\ and 5 type~\typec.
+Combined with the 2 type~\typeab\ and 6 type~\typec\ RRLs in the cluster itself, there are 24 RRLs in total associated to Pal~5 and its tails with membership probabilities larger than 0.5.
 In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.2$.
@@
 The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram.
-The distribution of the \typeab~RRL clearly indicate Pal~5 to be an Oosterhoff I cluster, consistent with their mean period of $0.565~d$ \todo{(XX) ?}.
+The distribution of the \typeab~RRLs clearly indicate that Pal~5 is an Oosterhoff I cluster, consistent with having a mean period of $0.565~d$ \todo{(XX) ?}.
 Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an OoI classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.32$ \citep{TODO}.
@@
 From these stars, we infer the distance to the cluster, finding a closer value, $d = 20.9 \pm 0.2~\kpc$, than many past studies, and detect a distance gradient of $\sim 0.4~\kpc~{\rm deg}^{-1}$ between the trailing and leading tail (between $-5^\circ < \phi_1 < 5^\circ$).
-We also, for the first time, detect trends in proper motion along the stream.
+% \sa{you mean only the cluster stars, right?}
+% APW: no, I mean all! Because of our quadratic model, even stars in the nearby part of the stream technically inform the cluster position
+We also, for the first time, detect trends in proper motion along the stream, predicted by models of the stream \citep[e.g.,][]{Pearson:2017}.
+% \sa{which has been suggested in/reported in/consistent with modeling of the stream...? fig 2, pearson 17}
+% APW: Not really worth saying consistent or not because of the distance difference, but I added the citation!
 We release the full RRL catalog in the region around Pal 5 (containing distance information from PS1 and astrometry from \Gaia) with our derived membership probabilities.
 These data will be instrumental for improving dynamical models of Pal 5, both for using Pal 5 to constrain the global matter distribution of the Milky Way and in interpreting the observed, complex density structure of the stream.
+% \sa{not sure where in text, but could be worth mentioning that you don't find stars beyond the current footprint of the stream (i.e.
+that leading stream does appear to be shorter than trailing..., although looking at fig 3, you actually don't find trailin stream RRLs much further from the prog.
+than you find leading arm RRLs.}
 

### 2019-08-23 paper/main.tex
Updates from Overleaf

@@
 Two all-sky RR Lyrae catalogs have also been released as part of \Gaia~\DR{2} \citep[VariClassifier and Specific Object Studies][]{Holl2018, Rimoldini2018, Clementini2018}.
-At present, however, these catalogs are subject to more significant and spatially-varying incompleteness in this particular area of the sky due to the \Gaia\ scanning law \citep[see][, though many of these issues will go away with future \Gaia\ data releases]{Rimoldini2018}.
+At present, however, these catalogs are subject to more significant and spatially-varying incompleteness in this particular area of the sky due to the \Gaia\ scanning law \citep[see][though many of these issues will go away with future \Gaia\ data releases]{Rimoldini2018}.
 % \sa{Mention if DR3 will help?}
@@
 % APW: I don't think so? And what do you mean by the "whole region"?
+%SP meant whether you found any RRLs below $\delta<-10$ (now found answer below so never mind)
 We \emph{do not} include the known spatial extent or morphology of the Pal 5 stream in our membership model for the stream: We instead later use the sky positions to validate the modeling procedure and selection criteria we then use to define a sample of probable Pal 5 stream RRLs.
@@
 % APW: no, I mean all! Because of our quadratic model, even stars in the nearby part of the stream technically inform the cluster position
+%SP: cool, does the distance estimate change at all if you only use cluster stars?
 We also, for the first time, detect trends in proper motion along the stream, predicted by models of the stream \citep[e.g.,][]{Pearson:2017}.

### 2019-08-26 paper/main.tex
Updates from Overleaf

@@
 % Results
-By building a probabilistic model of the Pal 5 cluster and stream in proper motion and distance, we find 25 RRLs consistent with being members of the cluster (10) and stream (15).
+By building a probabilistic model of the Pal 5 cluster and stream in proper motion and distance, we find 24 RRLs consistent with being members of the cluster (9) and stream (15).
 Using these RRLs, we detect gradients in distance and proper motion along the stream, and provide an updated measurement of the distance to the Pal 5 cluster using the RRLs.
@@
 
-% Based on the prior existence of  paucity of type~\typeab\ RRL stars of in the cluster, Pal~5  would have been expected to be classified as an Oosterhoff type II cluster (ref-XX, no one really seems to say this explicitly), at odds with the cluster's known spectroscopic metallicity of $\FeH=-1.32$ \citep{TODO}, as OoII clusters are on average more metal-poor than OoI clusters.
-
-Assuming a Jacobi radius of $11'$ (computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.2\times10^4~\msun$ \citep{Kuepper:2015} and using the Milky Way mass model in \citealt{gala}), we find 16 RRLs in the cluster tails, 11 classified as type~\typeab\ and 5 type~\typec.
-Combined with the 2 type~\typeab\ and 6 type~\typec\ RRLs in the cluster itself, there are 24 RRLs in total associated to Pal~5 and its tails with membership probabilities larger than 0.5.
-In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.2$.
+Assuming a Jacobi radius of $11'$---computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.2\times10^4~\msun$ \citep{Kuepper:2015} and using the Milky Way mass model in \citet{gala}---we find 10 RRLs in the cluster itself (3 type~\typeab\ and 7 type~\typec) and 15 in the cluster tails (11 type~\typeab\ and 4 type~\typec), all  with membership probabilities larger than 0.5.
+In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.27$, the typical ratio found Oosterhoff I populations \citep{Smith1995}.
 \figurename~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
 The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram.
-The distribution of the \typeab~RRLs clearly indicate that Pal~5 is an Oosterhoff I cluster, consistent with having a mean period of $0.565~d$ \todo{(XX) ?}.
-Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an OoI classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.32$ \citep{TODO}.
-Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars.
-This type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017} and the Sagittarius dwarf galaxy, but not in any other dwarf spheroidal galaxy.
+The distribution of the \typeab~RRLs clearly indicate that Pal~5 is an Oosterhoff I cluster, consistent with having a mean period of $0.53~d$.
+Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an Oosterhoff I classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.32$ \citep{TODO}.
+Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars, again consistent with Pal~5's metallicity,  as this type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017}.
+
+%.
+and the Sagittarius dwarf galaxy, but not in any other dwarf spheroidal galaxy.
 \todo{CM:( any other point to be made about this?)}.
@@
 maybe some more discussion/elaboration?
-}
+} \cm{well, it's not weird at all.
+I've added a sentence saying so.}
 
@@
 \includegraphics[width=\textwidth]{rrls_PA.pdf}
-\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field. \sa{some statement on what to take away from the figure... i.e. consistent with Pal 5 being an "x" type GC....}}
+\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field. The distribution of the RRLs found in the cluster and it's tails is consistent with an Oosterhoff type I. }
 \label{fig:PA_diagram}
@@
 %SP: cool, does the distance estimate change at all if you only use cluster stars?
+% APW: they are consistent
 We also, for the first time, detect trends in proper motion along the stream, predicted by models of the stream \citep[e.g.,][]{Pearson:2017}.

### 2019-08-26 paper/main.tex

@@
 For example, using the sky track and radial velocity information, the stream has been used to measure the enclosed mass of the Milky Way \citep{Kuepper:2015, Bovy:2016}.
-% \sa{if you say the "shape and mass", you can cite Pearson 15 too ;)}
-% APW: I say "measure" not "rule out" :) - but I'll reassess when I do the shortening and see if it fits
 Using deep photometry, the density variations along the stream have been used to place limits on the abundance of massive substructures in the Galactic halo \citep{Erkal:2017}.
@@
 
-Assuming a Jacobi radius of $11'$---computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.2\times10^4~\msun$ \citep{Kuepper:2015} and using the Milky Way mass model in \citet{gala}---we find 10 RRLs in the cluster itself (3 type~\typeab\ and 7 type~\typec) and 15 in the cluster tails (11 type~\typeab\ and 4 type~\typec), all  with membership probabilities larger than 0.5.
-In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.27$, the typical ratio found Oosterhoff I populations \citep{Smith1995}.
+Assuming a Jacobi radius of $11'$---computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.2\times10^4~\msun$ \citep{Kuepper:2015} and using the Milky Way mass model in \citet{gala}---we find 9 RRLs in the cluster itself (2 type~\typeab\ and 7 type~\typec) and 15 in the cluster's tails (11 type~\typeab\ and 4 type~\typec), all  with membership probabilities larger than 0.5.
+In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.2$, a typical ratio found Oosterhoff I populations \citep{Smith1995}.
 \figurename~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
 The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram.
-The distribution of the \typeab~RRLs clearly indicate that Pal~5 is an Oosterhoff I cluster, consistent with having a mean period of $0.53~d$.
+The distribution of the \typeab~RRLs clearly indicate that Pal~5 is an Oosterhoff I cluster, consistent with having a mean period of $0.54~d$.
 Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an Oosterhoff I classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.32$ \citep{TODO}.
@@
 
-The Pal~5 cluster has 5 previous known RRL stars (named V1--V5), all of type which are \typec, as reported in the \citet{Clement:2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{SawyerHogg:1973}.
+The Pal~5 cluster has 5 previously known RRL stars (named V1--V5), all of type which are \typec, as reported in the \citet{Clement:2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{SawyerHogg:1973}.
 In their analysis of the halo density profile with RRLs, \citet{Vivas:2006} report an overdensity (``Group~6'') that they claim could be associated with Pal~5's tidal tails.
+Two of these stars (IDs 403 and 405) are separated by just $\sim10'$ and $11'$ from the cluster's center, which places them within the cluster's Jacobi radius, and an additional star (star 393) outside of the Jacobi radius but at a similar distance could be associated with the stream \citep{Vivas:2006}.
+We confirm the three stars are highly probable members.
+More recently, \citet{Ibata:2017} use a prior iteration of the PS1 RRL catalog \citep{Hernitschek:2016}, which contained \emph{plausible} RRLs, to identify 6 and 12 candidate members of the Pal 5 cluster and tails, respectively, based on sky position and distance alone.
+However, without a published catalog, we cannot compare our sample to theirs.
+
 % The QUEST IDs of the stars in their Group~6 are: 403, 405, 407, 411, 416 and 422.
 % The first two stars --403 and 405-- are located at distances of  $9\farcm8$ and $11'$ from the cluster's center, which \citet{Vivas2004} notes places them within the cluster's tidal radius, assuming the value of $r_t\sim16'$ reported by \citet{Odenkirchen:2002}.
-Two of these stars (IDs 403 and 405) are separated by just $\sim10'$ and $11'$ from the cluster's center, which places them within the cluster's Jacobi radius, and an additional star (star 393) outside of the Jacobi radius but at a similar distance could be associated with the stream \citep{Vivas:2006}.
-More recently, \citet{Ibata:2017} use a prior iteration of the PS1 RRL catalog \citep{Hernitschek:2016} to identify \todo{XX} candidate members of the Pal 5 cluster and tails based on sky position and distance alone.
-However, without a published catalog, we cannot compare our sample to theirs.
 
@@
 
-Star 422 is reported as an \rrc~in \citet{Vivas2004}, with a period of 0.61925~d and $V$ amplitude 0.51~mag.
+%Star 422 is reported as an \rrc~in \citet{Vivas2004}, with a period of 0.61925~d and $V$ amplitude 0.51~mag.
 In PS1 it is classified as an \rrc~with a period of 0.3832276153~d.
@@
 
-Star \verb+source_id=4418731829516302848+ is reported in PS1 and \Gaia~as an \rrab~with a period around 0.646 and a similarly low amplitude $A_G=0.2$ and $A_g=0.3$ in both surveys.
+Star \verb+source_id=4418731829516302848+ is reported in PS1 and \Gaia~as an \rrab~with a period around 0.646~d and similarly low amplitudes $A_G=0.2$ and $A_g=0.3$ in both surveys.
 This star is reported in CRTS \citep{Drake2014} (ID CSS\_J151623.2-000831) as an \rrc~with a period 0.28164~d.
@@
 \toprule
-\Gaia~DR2               & Other & Period & Amp-$g$ & Type & D & member & sep  \\
+\Gaia~DR2               & Other & Period & Amp-$g$ & Type & D & member & separation  \\
 \verb+source_id+        & ID   & (d)    & (mag)   &     &(kpc)& prob & (") \\
 \midrule
-4418920808077110784 &  V5 & 0.337934& 0.46 & RRc  & 20.73 & 0.9982 &   1.9 \\
-4418920846732620032 &     & 0.643239& 0.27 & RRab & 21.71 & 0.9963 &   2.0 \\
-4418914863842345856 &  V1 & 0.293229& 0.56 & RRc  & 20.36 & 0.9696 &   2.0 \\
-4418726027016125056 &  V3 & 0.329948& 0.51 & RRc  & 20.65 & 0.9978 &   3.8 \\
-4418725889577171328 &  V4 & 0.286366& 0.59 & RRc  & 20.28 & 0.9915 &   4.5 \\
-4418731829516302848 &     & 0.28164 & 0.3  & RRc  & 18.14 & 0.9297 &   4.8 \\
-4418913218870688768 &  V2 & 0.332473& 0.4  & RRc  & 20.33 & 0.9977 &   5.0 \\
-4418734165978521728 &     & 0.380859& 0.52 & RRc  & 22.75 & 0.9850 &   7.6 \\
-4418724034151291776 & 403 & 0.551705& 1.14 & RRab & 21.23 & 0.9987 &   9.7 \\
-4418732791593809152 & 405 & 0.329670& 0.45 & RRc  & 20.74 & 0.9959 &  11.0 \\
-4419052204012341760 &     & 0.530006& 1.01 & RRab & 20.84 & 0.9960 &  44.5 \\
-4418142117622280192 & 393 & 0.462671& 1.36 & RRab & 19.95 & 0.9060 & 118.7 \\
-6339498589346000768 &     & 0.320110& 0.46 & RRc  & 18.89 & 0.9934 & 217.7 \\
-6339499379619987200 &     & 0.329577& 0.47 & RRc  & 19.35 & 0.9966 & 218.6 \\
-6339478312804685824 &     & 0.465883& 1.12 & RRab & 19.91 & 0.9970 & 242.8 \\
-4421078432143578496 &     & 0.520749& 1.11 & RRab & 21.98 & 0.9981 & 246.1 \\
-6339398155830632448 &     & 0.688316& 0.75 & RRab & 19.82 & 0.9982 & 258.9 \\
-4427253907921168000 &     & 0.496334& 1.35 & RRab & 21.81 & 0.9905 & 362.5 \\
-4427220338456828416 &     & 0.600796& 0.81 & RRab & 21.98 & 0.9974 & 384.4 \\
-4427234700827469952 &     & 0.517911& 1.24 & RRab & 21.94 & 0.9979 & 392.3 \\
-6337302795905934336 &     & 0.519462& 1.02 & RRab & 17.29 & 0.9808 & 404.5 \\
-6337233350579111680 &     & 0.542383& 0.96 & RRab & 18.98 & 0.9962 & 450.6 \\
-4427646704153765632 &     & 0.385552& 0.37 & RRc  & 19.90 & 0.9429 & 456.5 \\
-4424705647988505600 &     & 0.335502& 0.41 & RRc  & 20.87 & 0.9832 & 513.0 \\
-4426221707021159424 &     & 0.533313& 1.14 & RRab & 21.97 & 0.9728 & 550.1 \\
+4418920808077110784 & V5 & 0.337934 & 0.46 & RRc & 20.7 & 1.00 & 2.0 \\
+4418920846732620032 &    & 0.643239 & 0.27 & RRab & 21.7 & 1.00 & 2.0 \\
+4418914863842345856 & V1 & 0.293229 & 0.56 & RRc & 20.4 & 0.97 & 2.0 \\
+4418726027016125056 & V3 & 0.329948 & 0.51 & RRc & 20.7 & 1.00 & 3.9 \\
+4418725889577171328 & V4 & 0.286366 & 0.59 & RRc & 20.3 & 0.99 & 4.5 \\
+4418913218870688768 & V2 & 0.332473 & 0.40 & RRc & 20.3 & 1.00 & 5.1 \\
+4418734165978521728 &    & 0.380859 & 0.52 & RRc & 22.8 & 0.92 & 7.7 \\
+4418724034151291776 &403 & 0.551705 & 1.14 & RRab & 21.2 & 1.00 & 9.8 \\
+4418732791593809152 &405 & 0.329670 & 0.45 & RRc & 20.7 & 1.00 & 11.0 \\
+4419052204012341760 &    & 0.530006 & 1.01 & RRab & 20.8 & 1.00 & 44.5 \\
+4418142117622280192 &393 & 0.462671 & 1.36 & RRab & 20.0 & 0.92 & 118.8 \\
+6339498589346000768 &    & 0.320110 & 0.46 & RRc & 18.9 & 0.99 & 217.8 \\
+6339499379619987200 &    & 0.329577 & 0.47 & RRc & 19.4 & 1.00 & 218.6 \\
+6339478312804685824 &    & 0.465883 & 1.12 & RRab & 19.9 & 1.00 & 242.9 \\
+4421078432143578496 &    & 0.520749 & 1.11 & RRab & 22.0 & 1.00 & 246.1 \\
+6339398155830632448 &    & 0.688316 & 0.75 & RRab & 19.8 & 1.00 & 258.9 \\
+4427253907921168000 &    & 0.496334 & 1.35 & RRab & 21.8 & 0.99 & 362.6 \\
+4427220338456828416 &    & 0.600796 & 0.81 & RRab & 22.0 & 1.00 & 384.4 \\
+4427234700827469952 &    & 0.517911 & 1.24 & RRab & 21.9 & 1.00 & 392.4 \\
+6337302795905934336 &    & 0.519462 & 1.02 & RRab & 17.3 & 0.89 & 404.6 \\
+6337233350579111680 &    & 0.542383 & 0.96 & RRab & 19.0 & 0.93 & 450.7 \\
+4427646704153765632 &    & 0.385552 & 0.37 & RRc & 19.9 & 0.81 & 456.6 \\
+4424705647988505600 &    & 0.335502 & 0.41 & RRc & 20.9 & 0.99 & 513.0 \\
+4426221707021159424 &    & 0.533313 & 1.14 & RRab & 22.0 & 0.93 & 550.1 \\
 \bottomrule

### 2019-08-29 paper/main.tex

@@
              Princeton University, Princeton, NJ 08544, USA}
-% \email{aprice-whelan@flatironinstitute.org}
-% \correspondingauthor{Adrian M. Price-Whelan}
+\email{aprice-whelan@flatironinstitute.org}
+\correspondingauthor{Adrian M. Price-Whelan}
 
@@
 \includegraphics[width=\textwidth]{members.pdf}
-\caption{Sky positions of RR Lyra stars with Pal 5 membership probabilities $>0.01$, colored by membership probability (large markers).
-The background greyscale shows the surface density of CMD-filtered main sequence stars from \citet{Bonaca:2019}.}
+\caption{Sky positions of RR Lyrae stars with Pal 5 membership probabilities $>0.01$, colored by membership probability (large markers).
+The background greyscale shows the surface density of CMD-filtered main sequence stars from \citet{Bonaca:2019}.
+Note that the membership probability does not incorporate sky position information, but a clear over-density of high-probability RRL members do trace the stream on the sky.}
 \label{fig:members}
@@
 CM acknowledges funding from the MIA program at Universidad de la Republica, Uruguay, and is grateful for the hospitality and support of the Flatiron Institute where part of this research was carried out.
+
 This work has made use of data from the European Space Agency (ESA) mission

### 2019-09-02 paper/main.tex
Updates from Overleaf

@@
 
-\shorttitle{RR Lyrae in the Pal 5 stream}
+\shorttitle{RRL in the Pal 5 stream}
 \shortauthors{People et al.}
@@
 
-\title{Kinematics of the Palomar 5 stellar stream from RR Lyrae stars}
+\title{Kinematics of the Palomar 5 stellar stream from RRLs}
 
@@
 RR Lyrae stars (RRLs) are pulsating horizontal branch giants intrinsically more luminous than main sequence turn-off stars.
-Although about $\sim 10,000$ time less numerous than main sequence stars, RRLs are extremely useful probes of the Galactic halo as they are well established as standard candles; they can be observed to large distances \citep{Medina:2018,Sesar:2017c}; and are reliably identified based on their photometric variability, with little to no contamination from other types of variable stars \citep[e.g.]{Holl:2018,Drake:2017,Mateu:2012}.
-Their distances can be estimated based on photometric data alone, with errors that go from $\sim8\%$ in the optical to as low as 3\% in the infrared \citep{Neeley:2017, Sesar:2017a}.
+Although about $\sim 10,000$ time less numerous than main sequence stars, RRLs are extremely useful probes of the Galactic halo as they are well established as standard candles; they can be observed to large distances \citep{Medina:2018,Sesar:2017c}; and are reliably identified based on their photometric variability, with little to no contamination from other types of variable stars \citep[e.g.][]{Holl:2018,Drake:2017,Mateu:2012}.
+Their distances can be estimated based on photometric data alone, with errors $\sim8\%$ in the optical, and can be as low as 3\% in the infrared \citep{Neeley:2017}.
+These properties have made RRLs a key tracer of structure and substructure across the Galaxy, having been used to trace the density profile of the halo, thick disk and bulge \citep[e.g.][and reference therein]{Iorio:2018,Mateu:2018b,Kunder:2008}, to identify new streams and accretion events \citep[e.g.]{Duffau:2006,Sesar:2010,Mateu:2018,Iorio:2019} and even to argue against the extragalactic origin of overdensities in the disk \citep{Mateu:2009,Price-Whelan:2015}.
 
@@
 At its current distance, Pal~5's horizontal branch is bright enough ($\G\sim17.3$) to be observed by \Gaia.
-Therefore, combining \Gaia~DR2 proper motions with the newer PS1 RRL catalog from \citet{Sesar:2017b} --with improved classification probabilities and light curve periods--  will allow us to make precise distance and proper motion measurements along the full extent of the Pal~5 stream.
-Both, key ingredients for dynamical modelling of the Pal~5 stream...
-%meaning we can measure kinematic properties of the full stream...
-\citep{Ibata:2017} found some RRLs in pal 5, but no proper motions.
-This is useful for modeling...
-Many stream candidates found in RRLs alone \citep{Mateu:2018}.
+Therefore, combining \Gaia~DR2 proper motions with the newer PS1 RRL catalog from \citet{Sesar:2017b} --with improved classification probabilities and light curve periods--  will allow us to make the first precise distance and proper motion measurements along the full extent of the Pal~5 stream.
+Both will prove to be key ingredients for dynamical modelling of the Pal~5 stream.
 
-%---yes, I rewrite things many times. C.---
-%RR Lyrae stars (RRLs) are pulsating horizontal branch giants intrinsically more luminous than main sequence stars, but about $\sim 10,000$ time less numerous.
-At its current distance, Pal~5's horizontal branch is bright enough ($\G\sim17.3$) to be observed by \Gaia, so it's full extent can be traced and it's kinematic properties measured using RRLs.
-These stars can be reliably identified based on their photometric variability, with little to no contamination from other types of variable stars \citep[e.g.]{Holl2018,Drake2017,Mateu2012}, and are well known standard candles.
-Their distances can be estimated based on photometric data alone by means of their period-luminosity relation, with distance errors that go from $\sim8\%$ in the optical to as low as 3\% in the infrared \citep{Neeley:2017, Sesar:2017a}.
-Previous studies have found RRLs in Pal5's tails: In the QUEST survey, \citet{Vivas2005} found 2 RRLs; while  \citep{Ibata:2017} found 12 RRLs using the catalog of PS1 candidate RRLs from \citet{Hernitschek2015}.
-Combining \Gaia~DR2 proper motions with the newer PS1 RRL catalog from \citet{Sesar2017b} --with improved classification probabilities and light curve periods--  will allow for precise distance and proper motion measurements along the full extent of the Pal~5 stream.
 In this Article, we construct a probabilistic model for kinematic properties of RRLs in the vicinity of the Pal 5 stream to determine membership probabilities and thus map the distance and proper motion trends in the stream.
-In \sectionname~\ref{sec:data}, we describe the source RR Lyrae catalog we use to search for Pal 5 members.
+In \sectionname~\ref{sec:data}, we describe the source RRL catalog we use to search for Pal 5 members.
 In \sectionname~\ref{sec:membership}, we implement a probabilistic membership model to identify candidate members of the stream using distance and astrometric information.
-In \sectionname~\ref{sec:results}, we ...
+In \sectionname~\ref{sec:results}, we discuss the population and kinematics of the RRL members found and use these to provide an estimate of the system's stellar mass and luminosity.  
+We present our conclusions in \sectionname~\ref{sec:conclusions}.
 
@@
 
-Two all-sky RR Lyrae catalogs have also been released as part of \Gaia~\DR{2} \citep[VariClassifier and Specific Object Studies][]{Holl2018, Rimoldini2018, Clementini2018}.
+Two all-sky RRL catalogs have also been released as part of \Gaia~\DR{2} \citep[VariClassifier and Specific Object Studies][]{Holl2018, Rimoldini2018, Clementini2018}.
 At present, however, these catalogs are subject to more significant and spatially-varying incompleteness in this particular area of the sky due to the \Gaia\ scanning law \citep[see][though many of these issues will go away with future \Gaia\ data releases]{Rimoldini2018}.
@@
 
-For our present work, therefore, we use the superset of 68\,085 stars consisting of the 61\,795 \emph{bona fide} PS1 RRLs, plus the 6\,459 non-\emph{bona fide} PS1 stars also reported in the SOS \Gaia~\DR{2} catalog.
+For our present work, therefore, we use the superset of 68\,254 stars consisting of the 61\,795 \emph{bona fide} PS1 RRLs, plus the 6\,459 non-\emph{bona fide} PS1 stars also reported in the SOS \Gaia~\DR{2} catalog.
 Out of these, 68\,085 RRLs (99.8\%) have astrometric information available in \Gaia~\DR{2} and all---by construction---have photometric data and pulsation periods from PS1.
@@
 
-We cross-match the full PS1 RRL catalog with \Gaia~DR2 \citep{Gaia:2018}, using a 3" sky separation tolerance, in order to retrieve astrometric information for these stars.
+We cross-match the full PS1 RRL catalog with \Gaia~DR2 \citep{Gaia:2018}, using a 1" sky separation tolerance, in order to retrieve astrometric information for these stars.
 Although not all PS1 RRLs have been identified as RRLs from \Gaia\ photometry, all of the PS1 RRLs are present in the main point source catalog.
-\figurename~\ref{fig:pm} demonstrates that Pal 5 stream members appear in the RRL catalog: This figure shows the \Gaia\ \DR{2} proper motions for all RRLs in the sky window defined above within the distance range $18 < d < 25~\kpc$, then the same for RRLs in a tight selection around the observed stream track (within $1\degr$ of the stream track), and then finally the same for RRLs \emph{excluding} the stream region (excluding stars within $2\degr$ of the stream track).
+\figurename~\ref{fig:pm} demonstrates that Pal 5 stream members appear in the RRL catalog: This figure shows the \Gaia\ \DR{2} proper motions for all RRLs in the \textcolor{red}{selected} sky window within the distance range $18 < d < 25~\kpc$, then the same for RRLs in a tight selection around the observed stream track (within $1\degr$ of the stream track, \textcolor{red}{see \sectionname~\ref{sec:membership} for details}), and then finally the same for RRLs \emph{excluding} the stream region (excluding stars within $2\degr$ of the stream track).
 % \sa{Do we want to show this stream track? Or state how wide it is? (it's inthe  figure, but mention here too).}
@@
 They estimate the mean completeness to be 92\% for \rrab~and 79\% for \rrc, up to a distance of 40~kpc, well beyond the distance of Pal~5 and its tails.
-At the average ratio of 3 \rrc~stars per every 10 \rrab~\citep{Layden1995}, this represents a mean completeness of 89\%.
-We can also produce an estimate per line of sight using the methodology described in \citet{Rybizki2018}, which requires two independent catalogs to assess the completeness of both in a probabilistic manner.
-Using the \Gaia~DR2 VariClassifier \citet{Holl2018,Rimoldini2018} and Specific Objects Study \citep{Clementini2018} catalogs, we find a median completeness of 92\% with a standard deviation of 16\% with no evident spatial trend across the sky selection window.
+At the average ratio of 3 \rrc~stars per every 10 \rrab~\citep{Layden:1995}, this represents a mean completeness of 89\%.
+We can also produce an estimate per line of sight using the methodology described in \citet{Rybizki:2018}, which requires two independent catalogs to assess the completeness of both in a probabilistic manner.
+Comparing PS1 against the \Gaia~DR2 catalogue, comprised of the outer join of the VariClassifier \citep{Holl2018,Rimoldini2018} and Specific Objects Study \citep{Clementini2018} catalogs, we find a median completeness of 92\% with a standard deviation of 16\% with no evident spatial trend across the sky selection window.
 % \sa{I got a little confused in the section above as Fig 3 is referenced but Fig 2 never mentioned. }
 % APW: I moved this paragraph up from the discussion and forgot to edit - thanks!
-
-% \subsection{RRL Distances}
-
-% We assume a single intrinsic magnitude $M_G=+0.63$ for all RRLs, corresponding to the mean halo metallicity $\FeH=-1.5$ according to the calibration by \citet{Muraveva2018}, and compute distances to all stars based on their $G$-band intensity-averaged magnitudes: we use \verb+int_average_g+ for stars in VC+SOS and the intensity-averaged $g$ reported by \citet{Sesar2017b} converted to the G-band using the transformation of Eq.~XX from \citet{Evans2018}.
-To correct for the effect of dust extinction in the apparent magnitudes, we obtain the colour excess $E(B-V)$ for each RRL from the \citet{Schlegel1998} maps, and compute the V-band extinction as $A_V=2.742E(B-V)$ assuming an $R_V$=3.1 \citep{Schlafly2011}.
-Finally, the extinction is transformed to the $G$ band using the calibration from Carrasco (2018, private communication):
-% \begin{equation}
-% \begin{split}
-% f(\Gbp-\Grp, A_V) \equiv & a + b \ (\Gbp-\Grp) + c \ (\Gbp-\Grp)^2 \\
-%                          & +\ d \  A_V + e\  A_V \ (\Gbp-\Grp)
-% \end{split}
-% \end{equation}
-% \begin{equation}
-%  A_G = f(\Gbp-\Grp, A_V) \cdot A_V
-% \end{equation}
-% where $a=0.97883$, $b=-0.14365$, $c=0.011077$, $d=0.034842$, and $e=-0.0041448$; assuming a mean colour $BP-RP=0.7$ for RRLs.
-
-% \todo{Describe uncertainties and justification of relative precision vs. absolute error}
 
@@
 \includegraphics[width=\textwidth]{members.pdf}
-\caption{Sky positions of RR Lyrae stars with Pal 5 membership probabilities $>0.01$, colored by membership probability (large markers).
+\caption{Sky positions of RRLs with Pal 5 membership probabilities $>0.01$ \todo{CM: APW check this number, I think it might be 0.1?}, colored by membership probability \sout{(large markers)}.
 The background greyscale shows the surface density of CMD-filtered main sequence stars from \citet{Bonaca:2019}.
@@
 
-\subsection{RR Lyra stars associated with Pal 5 and its tails}
+\subsection{RRLs associated with Pal 5 and its tails}
 %SP: changed RR Lyrae stars to RR Lyra stars a few places, maybe check whole manuscript (otherwise plural twice).
-We find a total of 24 RR Lyrae stars with membership probability $> 0.5$ that lie within $1\degr$ of the stream track \citep[again using the track from][]{Bonaca:2019}.
+%CM: Sarah, I see your point, but I've never seen it written like that in the literature. Anyway, since we're using RRLs as an abbreviation, I've changed all instances to this in the whole manuscript.
+We find a total of 24 RRLs with membership probability $> 0.5$ that lie within $1\degr$ of the stream track \citep[again using the track from][]{Bonaca:2019}.
 Of those, 13 are classified as type \typeab\ and 11 are classified as type \typec\ RRLs.
-We find a total of 10 RRLs within the estimated Jacobi radius of the cluster ($\approx 12'$), meaning that 15 RRLs are spread between the leading and trailing tails.
+We find a total of 9 RRLs within the estimated Jacobi radius of the cluster ($\approx 12'$), meaning that 15 RRLs are spread between the leading and trailing tails.
 \figurename~\ref{fig:members} shows the sky positions of all RRLs with membership probability $>0.01$ over-plotted on the density of filtered main sequence stars from \citet{Bonaca:2019}.
-We note that there are a few RRLs outside of the main stream track that appear to be at a consistent distance with consistent proper motions (e.g., the two stars near $\alpha = 230\degr$, $\delta = 0\degr$) that are not included in the tally above.
+We note that there are a few RRLs outside of the main stream track that appear to be at a consistent distance with consistent proper motions (e.g., the two stars near $\alpha = 230\degr$, $\delta = 0\degr$ in \figurename~\ref{fig:members}) that are not included in the tally above.
 No theoretical models of Pal 5 have predicted tidal debris at such locations, so follow-up spectroscopy to confirm or rule out their origin would be very informative.
@@
 
-Past distance measurements to the Pal 5 cluster have ranged from $20.9~\kpc$ \citep{Dotter:2011} to $23.2~\kpc$ \citep{Harris:1996}, with a previous measurement from a small number of RR Lyrae stars in between at $22.3~\kpc$ \citep{Vivas:2006}.
+Past distance measurements to the Pal 5 cluster have ranged from $20.9~\kpc$ \citep{Dotter:2011} to $23.2~\kpc$ \citep{Harris:1996}, with a previous measurement from a small number of RRLs in between at $22.3~\kpc$ \citep{Vivas:2006}.
 % \sa{worth mentioning with which technique?}
@@
 
-\subsection{RR Lyrae population }
+\subsection{Properties of the RRL population}
 
 Assuming a Jacobi radius of $11'$---computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.2\times10^4~\msun$ \citep{Kuepper:2015} and using the Milky Way mass model in \citet{gala}---we find 9 RRLs in the cluster itself (2 type~\typeab\ and 7 type~\typec) and 15 in the cluster's tails (11 type~\typeab\ and 4 type~\typec), all  with membership probabilities larger than 0.5.
-In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.2$, a typical ratio found Oosterhoff I populations \citep{Smith1995}.
+In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.2$, a typical ratio found Oosterhoff I populations \citep{Smith:1995}.
 \figurename~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
@@
 The distribution of the \typeab~RRLs clearly indicate that Pal~5 is an Oosterhoff I cluster, consistent with having a mean period of $0.54~d$.
-Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an Oosterhoff I classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.32$ \citep{TODO}.
-Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars, again consistent with Pal~5's metallicity,  as this type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017}.
+Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an Oosterhoff I classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.4$ \citep{Dotter:2011}.
+Out of our newly found \rrab\ RRLs, two are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars, again consistent with Pal~5's metallicity,  as this type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017}.
 
@@
 \includegraphics[width=\textwidth]{rrls_PA.pdf}
-\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field. The distribution of the RRLs found in the cluster and it's tails is consistent with an Oosterhoff type I. }
+\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field membership probabilities $>0.1$ (small markers) inside the Pal~5 track (large markers), colored by membership probability.
+The distribution of the RRLs found in the cluster and it's tails is consistent with an Oosterhoff type I (dashed line).
+}
 \label{fig:PA_diagram}
@@
 
-%We can provide an estimate on the cluster's luminosity based on the observed number of RRLs. For that, the first step is to assess how complete our observed sample is.
-
-%we obtain the completeness map shown in Figure~\ref{fig:completeness_map}, for RRLs of type \typeab~and \typec.
-The mean completeness estimated this way is 92\% with a standard deviation of 16\% and, as the figure shows, there is no evident spatial trend.
-The two estimates are remarkably consistent and, in what follows, we will adopt a completeness of 90\%-XX.
-
 To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu:2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters, here including the RR\typec~stars.
-The $\log{N_{RR}}-M_V$ relation has a large scatter so, particularly for such low numbers of RRLs, it will only allow us to give rough limits on the cluster's absolute magnitude.
-Correcting the observed number of 26~RRLs by the median completeness of 92\%, we get a total expected number of 28~RRLs, for which we find an absolute magnitude $M_V=-8.1\pm 1.1$, which correspond to the mode and 16/84th percentiles of the posterior probability distribution.
-We therefore find that the initial mass of Pal 5 \todo{...}.
-
-% \begin{figure}
-% \begin{center}
-% \includegraphics[width=\columnwidth]{ps1_completeness.pdf}
-% \caption{PS1 RRL estimated completeness. The Pal 5 stream's track is shown with the solid line.}\label{fig:completeness_map}
-% \end{center}
-% \end{figure}
-
-% \todo{CM: estimate luminosity based on number of RRLs. Compare to estimate from $M_V-\FeH$ relation. Derive stellar mass from stellar luminosity. Compare to dynamical mass estimates.}
+The $\log{N_{RR}}-M_V$ relation has a large scatter so, particularly for such low numbers of RRLs, it will only allow us to give rough limits on the cluster's absolute magnitude and, therefore, of its luminosity.
+Correcting the observed number of 24~RRLs by the median completeness of 92\%, we get a total expected number of 26~RRLs, for which we find a total luminosity $L_V=3.9_{-2.5}^{+7.4}\times 10^4 \mathrm{L}_\sun$.%, which correspond to the mode and 16/84th percentiles of the posterior probability distribution.
+We then estimate a stellar mass of $0.8_{-0.5}^{+1.5}\times 10^5 \msun$, by assuming a V-band mass-to-light ratio $M/L=2.$ corresponding to a simple stellar population with the age (12 Gyr) and metallicity ($\FeH=-1.4$) of the cluster \citep{Dotter:2011}, computed from the \citet{Bruzual:2003} models \footnote{We have used Z=0.0004 ($\FeH=-1.6$), the nearest to Pal~5’s metallicity available in the models.}.
+Our mass estimate is larger than that of \citet{Ibata:2017}, who report a current mass of $1.2\pm 0.4 \times10^4\msun$ for the cluster and tails; although their estimate for the initial mass ($4.7\pm 1.5\times10^4\msun$) of the system is at the lower end of our confidence limits.
 
@@
 
-\subsection{Previously known Pal 5 RR Lyrae}
+\subsection{Previously known Pal 5 RRLs}
 
-The Pal~5 cluster has 5 previously known RRL stars (named V1--V5), all of type which are \typec, as reported in the \citet{Clement:2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{SawyerHogg:1973}.
-In their analysis of the halo density profile with RRLs, \citet{Vivas:2006} report an overdensity (``Group~6'') that they claim could be associated with Pal~5's tidal tails.
-Two of these stars (IDs 403 and 405) are separated by just $\sim10'$ and $11'$ from the cluster's center, which places them within the cluster's Jacobi radius, and an additional star (star 393) outside of the Jacobi radius but at a similar distance could be associated with the stream \citep{Vivas:2006}.
+The Pal~5 cluster has 5 previously known RRL stars (named V1--V5), all of type which are \typec, as reported in the \citet{Clement:2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{Kinman:1962}.
+In their analysis of the halo density profile with QUEST RRLs, \citet{Vivas:2006} report an overdensity (``Group~6'') that they claim could be associated with Pal~5's tidal tails.
+Two of these stars (IDs 403 and 405) are separated by just $\sim10'$ and $11'$ from the cluster's center, which places them within the cluster's Jacobi radius; and an additional star (star 393) outside of the Jacobi radius, but at a similar distance, that could be associated with the stream \citep{Vivas:2006}.
 We confirm the three stars are highly probable members.
+\citet{Wu:2005} reports another RRL star (133) as possibly associated, which we also find is a highly probable member.
 More recently, \citet{Ibata:2017} use a prior iteration of the PS1 RRL catalog \citep{Hernitschek:2016}, which contained \emph{plausible} RRLs, to identify 6 and 12 candidate members of the Pal 5 cluster and tails, respectively, based on sky position and distance alone.
@@
 
-% The QUEST IDs of the stars in their Group~6 are: 403, 405, 407, 411, 416 and 422.
-% The first two stars --403 and 405-- are located at distances of  $9\farcm8$ and $11'$ from the cluster's center, which \citet{Vivas2004} notes places them within the cluster's tidal radius, assuming the value of $r_t\sim16'$ reported by \citet{Odenkirchen:2002}.
-
-% In summary,  previous studies report 7 known RRLs associated to the cluster itself (1 \typeab~and 6 \typec) plus 5 more reported as possibly associated to the tails (4 \typeab~and 2 \typec).
-
 \section{Conclusions} \label{sec:conclusions}
 
-Using a catalog of RR Lyrae stars identified using multi-epoch, multi-band photometry from the Pan-STARRS PS1 survey and cross-matched to \Gaia\ \DR{2}, we construct a probabilistic model to determine membership with the Pal 5 cluster and stream.
+Using a catalog of RRLs identified using multi-epoch, multi-band photometry from the Pan-STARRS PS1 survey and cross-matched to \Gaia\ \DR{2}, we construct a probabilistic model to determine membership with the Pal 5 cluster and stream.
 We find 24 RRLs with membership probability $>0.5$ (13 type~\typeab\ and 11 type~\typec) over $\approx 18\degr$ of the stream.
@@
 
-Star \verb+source_id=6339498589346000768+ is reported both in PS1 and \Gaia~as an \rrc~, with a period 0.4714315854~d in PS1 and 0.320110123108~d in \Gaia~SOS, the two are a pair of 1~d aliases \citep[see e.g.]{Lafler1965}.
+Star \verb+source_id=6339498589346000768+ is reported both in PS1 and \Gaia~as an \rrc~, with a period 0.4714315854~d in PS1 and 0.320110123108~d in \Gaia~SOS, the two are a pair of 1~d aliases \citep[see e.g.]{Lafler:1965}.
 In what follows we adopt the shorter period reported in \Gaia~as the correct one as it is the more probable one for an \rrc~star.
@@
 Star \verb+source_id=4418731829516302848+ is reported in PS1 and \Gaia~as an \rrab~with a period around 0.646~d and similarly low amplitudes $A_G=0.2$ and $A_g=0.3$ in both surveys.
-This star is reported in CRTS \citep{Drake2014} (ID CSS\_J151623.2-000831) as an \rrc~with a period 0.28164~d.
+This star is reported in CRTS \citep{Drake:2017} (ID CSS\_J151623.2-000831) as an \rrc~with a period 0.28164~d.
 We adopt the CRTS period for this star in what follows, as it is based on many more observations ($\gtrsim200$) than \Gaia~and PS1's (about a dozen per filter in each case) and the \rrc~classification and short period are more likely for such a short amplitude star.
@@
 \begin{footnotesize}
+\begin{threeparttable}[t]
 \begin{tabular}{llllllll}
@@
 \Gaia~DR2               & Other & Period & Amp-$g$ & Type & D & member & separation  \\
-\verb+source_id+        & ID   & (d)    & (mag)   &     &(kpc)& prob & (") \\
+\verb+source_id+        & ID\tnote{*}   & (d)    & (mag)   &     &(kpc)& prob & (") \\
 \midrule
-4418920808077110784 & V5 & 0.337934 & 0.46 & RRc & 20.7 & 1.00 & 2.0 \\
+4418920808077110784 & V5 (K62) & 0.337934 & 0.46 & RRc & 20.7 & 1.00 & 2.0 \\
 4418920846732620032 &          & 0.643239 & 0.27 & RRab & 21.7 & 1.00 & 2.0 \\
-4418914863842345856 & V1 & 0.293229 & 0.56 & RRc & 20.4 & 0.97 & 2.0 \\
-4418726027016125056 & V3 & 0.329948 & 0.51 & RRc & 20.7 & 1.00 & 3.9 \\
-4418725889577171328 & V4 & 0.286366 & 0.59 & RRc & 20.3 & 0.99 & 4.5 \\
-4418913218870688768 & V2 & 0.332473 & 0.40 & RRc & 20.3 & 1.00 & 5.1 \\
+4418914863842345856 & V1 (K62) & 0.293229 & 0.56 & RRc & 20.4 & 0.97 & 2.0 \\
+4418726027016125056 & V3 (K62) & 0.329948 & 0.51 & RRc & 20.7 & 1.00 & 3.9 \\
+4418725889577171328 & V4 (K62) & 0.286366 & 0.59 & RRc & 20.3 & 0.99 & 4.5 \\
+4418913218870688768 & V2 (K62) & 0.332473 & 0.40 & RRc & 20.3 & 1.00 & 5.1 \\
 4418734165978521728 &          & 0.380859 & 0.52 & RRc & 22.8 & 0.92 & 7.7 \\
-4418724034151291776 &403 & 0.551705 & 1.14 & RRab & 21.2 & 1.00 & 9.8 \\
-4418732791593809152 &405 & 0.329670 & 0.45 & RRc & 20.7 & 1.00 & 11.0 \\
-4419052204012341760 &    & 0.530006 & 1.01 & RRab & 20.8 & 1.00 & 44.5 \\
-4418142117622280192 &393 & 0.462671 & 1.36 & RRab & 20.0 & 0.92 & 118.8 \\
+4418724034151291776 & 403 (V04)& 0.551705 & 1.14 & RRab & 21.2 & 1.00 & 9.8 \\
+4418732791593809152 & 405 (V04)& 0.329670 & 0.45 & RRc & 20.7 & 1.00 & 11.0 \\
+4419052204012341760 & 133 (W05)& 0.530006 & 1.01 & RRab & 20.8 & 1.00 & 44.5 \\
+4418142117622280192 & 393 (V04)& 0.462671 & 1.36 & RRab & 20.0 & 0.92 & 118.8 \\
 6339498589346000768 &          & 0.320110 & 0.46 & RRc & 18.9 & 0.99 & 217.8 \\
@@
 \end{tabular} 
+\begin{tablenotes}
+\centering
+     \item[*] References: K62=\citet{Kinman:1962}, V04=\citet{Vivas:2004}, W05=\citet{Wu:2005}
+\end{tablenotes}
+\end{threeparttable}%
 \end{footnotesize}

### 2019-09-02 paper/main.tex
remove some addressed comments

@@
+% TODO:
+% - Check whether I use p > 0.01 or p > 0.1 for sky plot figure
+% - Over-plot the sky track borders on MSTO plot
+% - Add PM and distance dispersion as parameters to fit, or cite out to the values I fix them to
+% - Compute the autocorrelation time for emcee instead, thin by the acor length
+%
+
 % \documentclass[modern]{aastex63}
@@
 
-\shorttitle{RRL in the Pal 5 stream}
+\shorttitle{RR Lyrae stars in the Pal 5 stream}
 \shortauthors{People et al.}
@@
 
-\title{Kinematics of the Palomar 5 stellar stream from RRLs}
+\title{Kinematics of the Palomar 5 stellar stream from RR Lyrae stars}
 
@@
 Conversely, of the $\sim$150 known globular clusters, only $\lesssim 20\%$ have purported tidal tails \citep[e.g.,][]{Leon:2000, Kundu:2019}, but even these are mostly low density and tenuous \citep[as might be expected, e.g.,][]{Balbinot:2018}.
-% \sa{The above sentence makes it sound like $\lesssim 20\%$ of the 150 known GCs have streams, is it really that many?}
-% APW: Yep! According to the citations.
 A prominent exception to these statements is the Palomar 5 (Pal 5) globular cluster and stream.
@@
 The Pal 5 stream---at a heliocentric distance of $d \sim 20~\kpc$---was discovered using early SDSS imaging, which reached well below the main sequence turnoff of the stream over a large area surrounding the cluster \citep{Odenkirchen:2001, Rockosi:2002}.
-% \sa{Maybe cite Dotter 2011 for previous distance? }
-% APW: It's just a qualitative statement to give a rough sense of its distance, so I'm not really quoting a measurement
 Since then, radial velocities have been measured for a handful of giant branch stars in the cluster and along the stream \citep{Odenkirchen:2002, Odenkirchen:2009, Ibata:2017}, and deeper imaging data has been used to map the tails at higher signal-to-noise \citep{Bernard:2016, Ibata:2016, Bonaca:2019}.
@@
 For example, the leading and trailing tails have significantly different total star counts and lengths \citep[both different by a factor of $\sim$2 between leading and trailing;][]{Dehnen:2004, Bernard:2016}, and the stream track, width, and density vary significantly over the full extent of the stream \citep{Ibata:2016, Bonaca:2019}.
-% \sa{Might be work stating how big the estimated mass difference is in the trailing, leading vs cluster, as many people seem to be curious about this.}
-% APW: Added a note above
 Given that Pal 5's pericenter is well within the Galactic disk ($r_{\textrm{peri}} \sim 7$--$8~\kpc$),
-% \sa{cite Erkal 2017 here (his fig 6)?}
-% APW: Again just a rough / qualtitative number to give a sense - I don't really see the need to cite given that this is from my own exploration with the orbit fits!
 these features could be a sign of perturbations from molecular clouds \citep[e.g.,][]{Amorisco:2016}, interaction with the Galactic bar \citep[e.g.,][]{Pearson:2017}, signatures of past dark matter subhalo encounters \citep[e.g.,][]{Erkal:2017}, or some combination of all of these effects.
@@
 Other stellar tracers (e.g., giant branch stars) are not numerous enough or precise enough distance indicators to provide useful relative distance information along the stream.
-% \sa{Maybe here mention that a "velocity" gradient has been detected along the stream (Kuzma 2015)? }
-% APW: Good idea, but then it would make sense to talk about the RV data in Ibata+2017. I'll try to fit both into the discussion.
-% Enter: RR Lyrae-type stars.
 
@@
 Their distances can be estimated based on photometric data alone, with errors $\sim8\%$ in the optical, and can be as low as 3\% in the infrared \citep{Neeley:2017}.
-These properties have made RRLs a key tracer of structure and substructure across the Galaxy, having been used to trace the density profile of the halo, thick disk and bulge \citep[e.g.][and reference therein]{Iorio:2018,Mateu:2018b,Kunder:2008}, to identify new streams and accretion events \citep[e.g.]{Duffau:2006,Sesar:2010,Mateu:2018,Iorio:2019} and even to argue against the extragalactic origin of overdensities in the disk \citep{Mateu:2009,Price-Whelan:2015}.
+These properties have made RRLs a key tracer of structure and substructure across the Galaxy, having been used to trace the density profile of the halo, thick disk and bulge \citep[e.g.][and reference therein]{Iorio:2018,Mateu:2018b,Kunder:2008}, to identify new streams and accretion events \citep[e.g.]{Duffau:2006,Sesar:2010,Mateu:2018,Iorio:2019} and even to argue against the extragalactic origin of overdensities in the disk \citep{Mateu:2009, Price-Whelan:2015}.
 
@@
 The distance selection here is purely illustrative: We do not use a distance cut in our membership model.
-% \sa{Might be worth looking at closer distances (leading arm in a MW potential comes as close as $\sim$15 kpc), especially if you want to find members at a dec of $-15\deg$.}
-% APW: Hm, my models only go to ~18 at the edge of the sky footprint, but this distance cut is only used for illustrative purposes. I added a sentence to clarify.
 \textbf{Middle}: The same, but for RRLs within $1\degr$ of the mean sky track of Pal 5 \citep{Bonaca:2019}.
@@
 Two all-sky RRL catalogs have also been released as part of \Gaia~\DR{2} \citep[VariClassifier and Specific Object Studies][]{Holl2018, Rimoldini2018, Clementini2018}.
-At present, however, these catalogs are subject to more significant and spatially-varying incompleteness in this particular area of the sky due to the \Gaia\ scanning law \citep[see][though many of these issues will go away with future \Gaia\ data releases]{Rimoldini2018}.
-% \sa{Mention if DR3 will help?}
-% APW: done!
+At present, however, these catalogs are subject to more significant and spatially-varying incompleteness in this particular area of the sky due to the \Gaia\ scanning law \citep[see][though many of these issues will improve with future \Gaia\ data releases]{Rimoldini2018}.
 
@@
 \figurename~\ref{fig:pm} demonstrates that Pal 5 stream members appear in the RRL catalog: This figure shows the \Gaia\ \DR{2} proper motions for all RRLs in the \textcolor{red}{selected} sky window within the distance range $18 < d < 25~\kpc$, then the same for RRLs in a tight selection around the observed stream track (within $1\degr$ of the stream track, \textcolor{red}{see \sectionname~\ref{sec:membership} for details}), and then finally the same for RRLs \emph{excluding} the stream region (excluding stars within $2\degr$ of the stream track).
-% \sa{Do we want to show this stream track? Or state how wide it is? (it's inthe  figure, but mention here too).}
-% APW: Noted.
 
@@
 Comparing PS1 against the \Gaia~DR2 catalogue, comprised of the outer join of the VariClassifier \citep{Holl2018,Rimoldini2018} and Specific Objects Study \citep{Clementini2018} catalogs, we find a median completeness of 92\% with a standard deviation of 16\% with no evident spatial trend across the sky selection window.
-% \sa{I got a little confused in the section above as Fig 3 is referenced but Fig 2 never mentioned. }
-% APW: I moved this paragraph up from the discussion and forgot to edit - thanks!
 
@@
 In order to search for new RRL stars in the Pal 5 stellar stream, we construct a probabilistic model of the stream and background stellar distribution (simultaneously) using proper motion $(\mu_\alpha, \mu_\delta)$ and distance $d$ data for all RRLs in the sky window $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$, where $(\alpha, \delta)$ are right ascension and declination.
-% \sa{Is it worth it to show the whole region in fig 3, and/or maybe extend to even more negative $\delta $?}
-% APW: I don't think so? And what do you mean by the "whole region"?
-%SP meant whether you found any RRLs below $\delta<-10$ (now found answer below so never mind)
 We \emph{do not} include the known spatial extent or morphology of the Pal 5 stream in our membership model for the stream: We instead later use the sky positions to validate the modeling procedure and selection criteria we then use to define a sample of probable Pal 5 stream RRLs.
@@
 We optimize the parameters of a Gaussian Mixture Model (GMM) representation of the (deconvolved) background RRL density using \project{extreme deconvolution} \citep[\acronym{XD};][]{Bovy:XD}.
-% \sa{minor: three sentences in a row start with "we use".}
-% APW: changed one of them
 We first fit the background model (with \acronym{XD}) using between $3$--$12$ mixture components and evaluate the Bayesian Information Criterion (BIC) for the optimal parameters in each case.
@@
 We note again that the membership probability does not depend on position or proximity to the stream, and only on distance and proper motions.
-% \sa{Could be worth emphasizing here too that the probability does not depend on positional info (i.e. being close to the stream itself) for people only skimming through figures.}
-% APW: added!
 \textbf{Upper left}: Sky positions (in the Pal 5 stream coordinate frame) of RRLs, with arrows showing the \Gaia\ \DR{2} proper motion vector direction.
@@
 \includegraphics[width=\textwidth]{members.pdf}
-\caption{Sky positions of RRLs with Pal 5 membership probabilities $>0.01$ \todo{CM: APW check this number, I think it might be 0.1?}, colored by membership probability \sout{(large markers)}.
+\caption{Sky positions of RRLs with Pal 5 membership probabilities $>0.01$ \todo{CM: APW check this number, I think it might be 0.1?}, colored by membership probability (markers).
 The background greyscale shows the surface density of CMD-filtered main sequence stars from \citet{Bonaca:2019}.
@@
 We assume that our prior is separable in all of our parameters: We use a uniform prior for the mixture weight, $f$, such that $p(f) = \mathcal{U}(0, 1)$, Gaussian priors on the mean proper motion and distance using the Pal 5 cluster proper motion measurements from \citet{Vasiliev:2019} and the Pal 5 distance from \citet{Kuepper:2015}.
-We use broad uniform priors on the linear and quadratic polynomial coefficients defined in \equationname~\ref{eq:meanpoly}.
+We use broad uniform priors (i.e. $-100 < b, c < 100$ for each coefficient) on the linear and quadratic polynomial coefficients defined in \equationname~\ref{eq:meanpoly}.
 We use the affine-invariant ensemble MCMC sampler \texttt{emcee} \citep{emcee} to sample from the posterior probability distribution given above.
@@
 No theoretical models of Pal 5 have predicted tidal debris at such locations, so follow-up spectroscopy to confirm or rule out their origin would be very informative.
-% \sa{What about RRLs at more negative $\delta$, wasn't your region larger than the figure extent?}
-% APW: There aren't any stars with prob > 0.01 below dec ~ -11 near the stream, so I trimmed the axis.
 
@@
 We find a proper motion $(\mu_\alpha\,\cos\delta, \mu_\delta) = (-2.69, -2.70) \pm (0.04, 0.03)~\masyr$, consistent with the measurement in \citet{Vasiliev:2019}.
-% \sa{state that this is larger than most before next sentence?}
 The larger proper motion (relative to \citealt{Fritz:2015} and other earlier measurements) and the closer distance means that the tangential velocity of Pal 5 is $\approx 30~\kms$ larger than most previous models have assumed.
@@
 From these stars, we infer the distance to the cluster, finding a closer value, $d = 20.9 \pm 0.2~\kpc$, than many past studies, and detect a distance gradient of $\sim 0.4~\kpc~{\rm deg}^{-1}$ between the trailing and leading tail (between $-5^\circ < \phi_1 < 5^\circ$).
-% \sa{you mean only the cluster stars, right?}
-% APW: no, I mean all! Because of our quadratic model, even stars in the nearby part of the stream technically inform the cluster position
-%SP: cool, does the distance estimate change at all if you only use cluster stars?
-% APW: they are consistent
 We also, for the first time, detect trends in proper motion along the stream, predicted by models of the stream \citep[e.g.,][]{Pearson:2017}.
-% \sa{which has been suggested in/reported in/consistent with modeling of the stream...? fig 2, pearson 17}
-% APW: Not really worth saying consistent or not because of the distance difference, but I added the citation!
 We release the full RRL catalog in the region around Pal 5 (containing distance information from PS1 and astrometry from \Gaia) with our derived membership probabilities.

### 2019-09-02 paper/main.tex
orcid

@@
 % - Compute the autocorrelation time for emcee instead, thin by the acor length
-%
 
@@
 \shorttitle{RR Lyrae stars in the Pal 5 stream}
-\shortauthors{People et al.}
+\shortauthors{Price-Whelan et al.}
 \newcommand{\sa}[1]{{\color{teal} SP: #1}}
@@
 
-\author{Giuliano~Iorio}
+\author[0000-0003-0293-503X]{Giuliano~Iorio}
 \affiliation{Institute of Astronomy, University of Cambridge, Madingley Road, Cambridge CB3 0HA, UK}
@@
 
-\author{Vasily~Belokurov}
+\author[0000-0002-0038-9584]{Vasily~Belokurov}
 \affiliation{Institute of Astronomy, University of Cambridge, Madingley Road, Cambridge CB3 0HA, UK}
@@
 Although not all PS1 RRLs have been identified as RRLs from \Gaia\ photometry, all of the PS1 RRLs are present in the main point source catalog.
-\figurename~\ref{fig:pm} demonstrates that Pal 5 stream members appear in the RRL catalog: This figure shows the \Gaia\ \DR{2} proper motions for all RRLs in the \textcolor{red}{selected} sky window within the distance range $18 < d < 25~\kpc$, then the same for RRLs in a tight selection around the observed stream track (within $1\degr$ of the stream track, \textcolor{red}{see \sectionname~\ref{sec:membership} for details}), and then finally the same for RRLs \emph{excluding} the stream region (excluding stars within $2\degr$ of the stream track).
+\figurename~\ref{fig:pm} demonstrates that Pal 5 stream members appear in the RRL catalog: This figure shows the \Gaia\ \DR{2} proper motions for all RRLs in the selected sky window within the distance range $18 < d < 25~\kpc$, then the same for RRLs in a tight selection around the observed stream track (within $1\degr$ of the stream track, see \sectionname~\ref{sec:membership} for details), and then finally the same for RRLs \emph{excluding} the stream region (excluding stars within $2\degr$ of the stream track).
 

### 2019-09-03 paper/main.tex
Updates from Overleaf

@@
 % - Add PM and distance dispersion as parameters to fit, or cite out to the values I fix them to
+%     - APW: I tried running with these as parameters, but they are unconstrained, so doesn't help. I'll just set the dispersions to small (<uncertainties) values and explain.
 % - Compute the autocorrelation time for emcee instead, thin by the acor length
@@
 \includegraphics[width=\textwidth]{members.pdf}
-\caption{Sky positions of RRLs with Pal 5 membership probabilities $>0.01$ \todo{CM: APW check this number, I think it might be 0.1?}, colored by membership probability (markers).
+\caption{Sky positions of RRLs with Pal 5 membership probabilities $>0.1$, colored by membership probability (markers).
 The background greyscale shows the surface density of CMD-filtered main sequence stars from \citet{Bonaca:2019}.
@@
 We find a total of 9 RRLs within the estimated Jacobi radius of the cluster ($\approx 12'$), meaning that 15 RRLs are spread between the leading and trailing tails.
-\figurename~\ref{fig:members} shows the sky positions of all RRLs with membership probability $>0.01$ over-plotted on the density of filtered main sequence stars from \citet{Bonaca:2019}.
+\figurename~\ref{fig:members} shows the sky positions of all RRLs with membership probability $>0.1$ over-plotted on the density of filtered main sequence stars from \citet{Bonaca:2019}.
 We note that there are a few RRLs outside of the main stream track that appear to be at a consistent distance with consistent proper motions (e.g., the two stars near $\alpha = 230\degr$, $\delta = 0\degr$ in \figurename~\ref{fig:members}) that are not included in the tally above.

### 2019-09-03 paper/main.tex
model in pm12

@@
-% TODO:
-% - Check whether I use p > 0.01 or p > 0.1 for sky plot figure
-% - Over-plot the sky track borders on MSTO plot
-% - Add PM and distance dispersion as parameters to fit, or cite out to the values I fix them to
-%     - APW: I tried running with these as parameters, but they are unconstrained, so doesn't help. I'll just set the dispersions to small (<uncertainties) values and explain.
-% - Compute the autocorrelation time for emcee instead, thin by the acor length
-
 % \documentclass[modern]{aastex63}
@@
 \newcommand{\sa}[1]{{\color{teal} SP: #1}}
+
+% Measurements:
+\newcommand{\cld}{\ensuremath{20.6~\kpc}}
+\newcommand{\clderr}{\ensuremath{20.6 \pm 0.2~\kpc}}
+\newcommand{\NRRL}{27}    % N total, prob > 0.5
+\newcommand{\NRRab}{16}   % N type ab, total
+\newcommand{\NRRc}{11}    % N type c, total
+\newcommand{\NRRcl}{9}    % N in cluster
+\newcommand{\NRRtails}{18} % N in tails
 
@@
 % Results
-By building a probabilistic model of the Pal 5 cluster and stream in proper motion and distance, we find 24 RRLs consistent with being members of the cluster (9) and stream (15).
-Using these RRLs, we detect gradients in distance and proper motion along the stream, and provide an updated measurement of the distance to the Pal 5 cluster using the RRLs.
+By building a probabilistic model of the Pal 5 cluster and stream in proper motion and distance, we find \NRRL\ RRLs consistent with being members of the cluster (\NRRcl) and stream (\NRRtails).
+Using these RRLs, we detect gradients in distance and proper motion along the stream, and provide an updated measurement of the distance to the Pal 5 cluster using the RRLs, $d = \clderr$.
 % Conclusions
@@
 RR Lyrae stars (RRLs) are pulsating horizontal branch giants intrinsically more luminous than main sequence turn-off stars.
-Although about $\sim 10,000$ time less numerous than main sequence stars, RRLs are extremely useful probes of the Galactic halo as they are well established as standard candles; they can be observed to large distances \citep{Medina:2018,Sesar:2017c}; and are reliably identified based on their photometric variability, with little to no contamination from other types of variable stars \citep[e.g.][]{Holl:2018,Drake:2017,Mateu:2012}.
+Although about $\sim 10,000$ time less numerous than main sequence stars, RRLs are extremely useful probes of the Galactic halo as they are well-established as standard candles; they can be observed to large distances \citep{Medina:2018,Sesar:2017c}; and are reliably identified based on their photometric variability, with little to no contamination from other types of variable stars \citep[e.g.,][]{Holl:2018,Drake:2017,Mateu:2012}.
 Their distances can be estimated based on photometric data alone, with errors $\sim8\%$ in the optical, and can be as low as 3\% in the infrared \citep{Neeley:2017}.
-These properties have made RRLs a key tracer of structure and substructure across the Galaxy, having been used to trace the density profile of the halo, thick disk and bulge \citep[e.g.][and reference therein]{Iorio:2018,Mateu:2018b,Kunder:2008}, to identify new streams and accretion events \citep[e.g.]{Duffau:2006,Sesar:2010,Mateu:2018,Iorio:2019} and even to argue against the extragalactic origin of overdensities in the disk \citep{Mateu:2009, Price-Whelan:2015}.
+These properties have made RRLs a key tracer of structure and substructure across the Galaxy, having been used to trace the density profile of the halo, thick disk and bulge \citep[e.g.,][and reference therein]{Iorio:2018,Mateu:2018b,Kunder:2008}, to identify new streams and accretion events \citep[e.g.]{Duffau:2006,Sesar:2010,Mateu:2018,Iorio:2019} and even to argue against the extragalactic origin of overdensities in the disk \citep{Mateu:2009, Price-Whelan:2015}.
 
-Previous studies have found RRLs in Pal5's tails: the first two were found by \citet{Vivas:2001} in the QUEST survey; and more recently \citep{Ibata:2017} found about ten new RRLs using the catalog of PS1 candidate RRLs from \citet{Hernitschek:2016}.
+Previous studies have found RRLs in Pal5's tails: the first two were found by \citet{Vivas:2001} in the QUEST survey, and more recently \citep{Ibata:2017} found about ten new RRLs using a catalog of PS1 candidate RRLs from \citet{Hernitschek:2016}.
 At its current distance, Pal~5's horizontal branch is bright enough ($\G\sim17.3$) to be observed by \Gaia.
-Therefore, combining \Gaia~DR2 proper motions with the newer PS1 RRL catalog from \citet{Sesar:2017b} --with improved classification probabilities and light curve periods--  will allow us to make the first precise distance and proper motion measurements along the full extent of the Pal~5 stream.
+Therefore, combining \Gaia~DR2 proper motions with the newer PS1 RRL catalog from \citet{Sesar:2017b}---with improved classification probabilities and light curve periods---enables us to make the first precise distance and proper motion measurements along the full extent of the Pal~5 stream.
 Both will prove to be key ingredients for dynamical modelling of the Pal~5 stream.
@@
 Although not all PS1 RRLs have been identified as RRLs from \Gaia\ photometry, all of the PS1 RRLs are present in the main point source catalog.
-\figurename~\ref{fig:pm} demonstrates that Pal 5 stream members appear in the RRL catalog: This figure shows the \Gaia\ \DR{2} proper motions for all RRLs in the selected sky window within the distance range $18 < d < 25~\kpc$, then the same for RRLs in a tight selection around the observed stream track (within $1\degr$ of the stream track, see \sectionname~\ref{sec:membership} for details), and then finally the same for RRLs \emph{excluding} the stream region (excluding stars within $2\degr$ of the stream track).
+\figurename~\ref{fig:pm} demonstrates that Pal 5 stream members appear in the RRL catalog: This figure shows the \Gaia\ \DR{2} proper motions for all RRLs in the selected sky window (indicated in the caption) within the distance range $18 < d < 25~\kpc$, then the same for RRLs in a tight selection around the observed stream track (see \sectionname~\ref{sec:membership} for details), and then finally the same for RRLs \emph{excluding} the stream region.
 
@@
 
-In order to search for new RRL stars in the Pal 5 stellar stream, we construct a probabilistic model of the stream and background stellar distribution (simultaneously) using proper motion $(\mu_\alpha, \mu_\delta)$ and distance $d$ data for all RRLs in the sky window $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$, where $(\alpha, \delta)$ are right ascension and declination.
+In order to search for new RRL stars in the Pal 5 stellar stream, we construct a probabilistic model of the stream and background stellar distribution (simultaneously) using proper motion $(\mu_1, \mu_2) = (\mu_{\phi_1}\cos\phi_2, \mu_{\phi_2})$ and distance $d$ data for all RRLs in the sky window $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$, where $(\alpha, \delta)$ are right ascension and declination and $\phi_1, \phi_2$ are a rotated spherical coordinate system aligned with the stream \citep{Bonaca:2019}.
 We \emph{do not} include the known spatial extent or morphology of the Pal 5 stream in our membership model for the stream: We instead later use the sky positions to validate the modeling procedure and selection criteria we then use to define a sample of probable Pal 5 stream RRLs.
@@
 
-To start, we use the sky track and width of Pal 5 from \citet{Bonaca:2019} to mask out a $\sim4^\circ$-wide region around the stream---and an extrapolation of the known extent of the stream---in order to define a ``background'' region that should be devoid of Pal 5 member stars.
+To start, we use the sky track and width of Pal 5 from \citet{Bonaca:2019} to mask out a wide region around the stream---and an extrapolation of the known extent of the stream---in order to define a ``background'' region that should be devoid of Pal 5 member stars (see purple lines in \figurename~\ref{fig:members}).
 We use RRL stars in the background region to construct an error-deconvolved density model in the 3D space of proper motion components and distance.
@@
 \textbf{Upper right}: Distance as a function of Pal 5 longitude, $\phi_1$, for the RRLs.
-The shaded (orange) band shows the 16--84th percentile confidence region for the inferred distance trend of the stream, and the solid line shows the median posterior sample, all assuming a quadratic function of $\phi_1$.
-\textbf{Lower panels}: Same as upper right, but for the proper motion in right ascension, $\mu_\alpha$, and declination, $\mu_\delta$.
+The shaded (orange) band shows the 5--95th percentile confidence region for the inferred distance trend of the stream, and the solid line shows the median posterior sample, all assuming a quadratic function of $\phi_1$.
+\textbf{Lower panels}: Same as upper right, but for the proper motion components in the Pal 5 coordinate system, $\mu_1$ and $\mu_2$.
 }
@@
 
-To represent the Pal 5 stream, we use a single Gaussian component with a fixed dispersion of $0.05~\masyr$ for the proper motion components ($\approx 5~\kms$ at the distance of Pal 5) and $0.2~\kpc$ in distance.
-However, we allow the mean to vary independently in each component as a function of $\phi_1$, a longitude coordinate (along the stream) as defined in \citet{Bonaca:2019}.
-Based on the Pal 5 stream models from \citet{Bonaca:2019}, over the range of $\phi_1 \in (-20^\circ, 15^\circ)$, the mean stream trends in $\mu_\alpha$, $\mu_\delta$, and distance can be well-approximated (within our observational uncertainties) by a quadratic function in $\phi_1$.
+To represent the Pal 5 stream, we use a single Gaussian component with a fixed dispersion of $0.05~\masyr$ for the proper motion components and $0.2~\kpc$ in distance.
+We have tried including the dispersions as additional parameters in our model but find that they are unconstrained; We therefore fix these values as mentioned to be smaller than the typical uncertainties.
+We allow the mean of the Gaussian component to vary independently in each component as a function of $\phi_1$.
+Based on the Pal 5 stream models from \citet{Bonaca:2019}, over the range of $\phi_1 \in (-20^\circ, 15^\circ)$, the mean stream trends in $\mu_1, \mu_2$, and distance $d$ can be well-approximated (within our observational uncertainties) by a quadratic function in $\phi_1$.
 The mean of our stream component, $\bs{x}$, is therefore given by
@@
     \bs{x} = \begin{pmatrix}
-        \mu_\alpha^* + b_\alpha \, \phi_1 + c_\alpha \, \phi_1^2\\
-        \mu_\delta^* + b_\delta \, \phi_1 + c_\delta \, \phi_1^2\\
-        d^* + b_d \, \phi_1 + c_d \, \phi_1 \label{eq:meanpoly}
+        \mu_1^* + b_1 \, (\phi_1 - x_1) + c_1 \, (\phi_1 - x_1)^2\\
+        \mu_2^* + b_2 \, (\phi_1 - x_2) + c_2 \, (\phi_1 - x_2)^2\\
+        d^* + b_d \, (\phi_1 - x_d) + c_d \, (\phi_1 - x_d)^2
+        \label{eq:meanpoly}
     \end{pmatrix}
 \end{equation}
-where $\bs{\theta} = (\mu_\alpha^*, \mu_\delta^*, d^*, b_\alpha, b_\delta, b_d, c_\alpha, c_\delta, c_d)$ are free parameters.
+where $\bs{\theta} = (\mu_1^*, \mu_2^*, d^*, b_1, b_2, b_d, c_1, c_2, c_d, x_1, x_2, x_d)$ are free parameters.
 
-Given the above background model, and this model for the stream track, the full likelihood for an RRL star, $n$, with data $D_n = (\mu_\alpha, \mu_\delta, d)_n$ is represented as a mixture of the two components with mixture weight $f$,
+Given the above background model, and this model for the stream track, the full likelihood for an RRL star, $n$, with data $D_n = (\mu_1, \mu_2, d)_n$ is represented as a mixture of the two components with mixture weight $f$,
 \begin{equation}
@@
 The background greyscale shows the surface density of CMD-filtered main sequence stars from \citet{Bonaca:2019}.
-Note that the membership probability does not incorporate sky position information, but a clear over-density of high-probability RRL members do trace the stream on the sky.}
+Note that the membership probability does not incorporate sky position information, but a clear over-density of high-probability RRL members do trace the stream on the sky.
+Purple lines outline the stream track window defined in \citet{Bonaca:2019} and used here to exclude stream stars when constructing the background model.}
 \label{fig:members}
@@
 We assume that our prior is separable in all of our parameters: We use a uniform prior for the mixture weight, $f$, such that $p(f) = \mathcal{U}(0, 1)$, Gaussian priors on the mean proper motion and distance using the Pal 5 cluster proper motion measurements from \citet{Vasiliev:2019} and the Pal 5 distance from \citet{Kuepper:2015}.
-We use broad uniform priors (i.e. $-100 < b, c < 100$ for each coefficient) on the linear and quadratic polynomial coefficients defined in \equationname~\ref{eq:meanpoly}.
+We use broad uniform priors (i.e.
+$-100 < b, c < 100$ for each coefficient, and $-20^\circ < x < 15^\circ$ for each polynomial centroid) for the trend parameters defined in \equationname~\ref{eq:meanpoly}.
 We use the affine-invariant ensemble MCMC sampler \texttt{emcee} \citep{emcee} to sample from the posterior probability distribution given above.
-We run \texttt{emcee} with 80 walkers for an initial 1024 steps to ``burn-in'' the sampler, then reset and restart the sampler for another 2048 steps.
-We compute the Gelman-Rubin convergence statistic $\mathcal{R}$ \citep{Gelman:1992} to verify that $\mathcal{R} < 1.1$ for all walkers and assume that the chains have converged.
-We then thin the chains by saving only every 128th step, leaving us with 1,280 posterior samples in our membership model parameters, $\bs{\theta}$.
-For each star, we also compute posterior membership probabilities using the samples following \citet{DFM:blog}.
+We run \texttt{emcee} with 104 walkers for an initial 1024 steps to ``burn-in'' the sampler, then reset and restart the sampler for another 65536 steps.
+We estimate the maximum (over parameters) autocorrelation length using \texttt{emcee}, $\tau_{\rm max} \approx 1200$, and thin the chains by keeping only every 1200th step, leaving us with 5616 posterior samples in our membership model parameters, $\bs{\theta}$.
+For each star, we use these samples to compute posterior membership probabilities following \citet{DFM:blog}.
 
@@
 \subsection{RRLs associated with Pal 5 and its tails}
-%SP: changed RR Lyrae stars to RR Lyra stars a few places, maybe check whole manuscript (otherwise plural twice).
-%CM: Sarah, I see your point, but I've never seen it written like that in the literature. Anyway, since we're using RRLs as an abbreviation, I've changed all instances to this in the whole manuscript.
-We find a total of 24 RRLs with membership probability $> 0.5$ that lie within $1\degr$ of the stream track \citep[again using the track from][]{Bonaca:2019}.
-Of those, 13 are classified as type \typeab\ and 11 are classified as type \typec\ RRLs.
-We find a total of 9 RRLs within the estimated Jacobi radius of the cluster ($\approx 12'$), meaning that 15 RRLs are spread between the leading and trailing tails.
+We find a total of \NRRL\ RRLs with membership probability $> 0.5$ that lie within the stream track \citep[again using the track from][see lines in Figure~\ref{fig:members}]{Bonaca:2019}.
+Of those, \NRRab\ are classified as type \typeab\ and \NRRc\ are classified as type \typec\ RRLs.
+We find a total of \NRRcl\ RRLs within the estimated Jacobi radius of the cluster ($\approx 11'$), meaning that \NRRtails\ RRLs are spread between the leading and trailing tails.
 \figurename~\ref{fig:members} shows the sky positions of all RRLs with membership probability $>0.1$ over-plotted on the density of filtered main sequence stars from \citet{Bonaca:2019}.
@@
 Past distance measurements to the Pal 5 cluster have ranged from $20.9~\kpc$ \citep{Dotter:2011} to $23.2~\kpc$ \citep{Harris:1996}, with a previous measurement from a small number of RRLs in between at $22.3~\kpc$ \citep{Vivas:2006}.
-% \sa{worth mentioning with which technique?}
-We find a mean Pal 5 cluster distance of $20.9 \pm 0.2~\kpc$ using the PS1 RRLs and the membership model described in \sectionname~\ref{sec:membership}.
-This is closer than most past measurements of the cluster distance, but consistent with the measurement from \citet{Dotter:2011}.
-We find a proper motion $(\mu_\alpha\,\cos\delta, \mu_\delta) = (-2.69, -2.70) \pm (0.04, 0.03)~\masyr$, consistent with the measurement in \citet{Vasiliev:2019}.
-The larger proper motion (relative to \citealt{Fritz:2015} and other earlier measurements) and the closer distance means that the tangential velocity of Pal 5 is $\approx 30~\kms$ larger than most previous models have assumed.
+We find a mean Pal 5 cluster distance of \clderr\ using the PS1 RRLs and the membership model described in \sectionname~\ref{sec:membership}.
+This is closer to the sun than most past measurements of the cluster distance, but consistent with the measurement from \citet{Dotter:2011}.
+We find a proper motion $(\mu_{\phi_1}\cos\phi_2, \mu_{\phi_2}) = (3.78, 0.71) \pm (0.02, 0.03)~\masyr$ or $(\mu_\alpha\,\cos\delta, \mu_\delta) = (-2.75, -2.69)~\masyr$, consistent with the measurement in \citet{Vasiliev:2019} within our uncertainties.
+The larger proper motion (relative to \citealt{Fritz:2015} and other earlier measurements) and the updated distance means that the tangential velocity of Pal 5 is $\approx 16~\kms$ larger than most previous models have assumed.
 This is a noteworthy update to consider for future stream modeling efforts and will likely change the inferred Milky Way parameters as compared to past models \citep[e.g.,][]{Kuepper:2015, Bovy:2016}.
 
-We also detect a distance trend from the closer leading tail ($d \approx 19~\kpc$ at $\phi_1 \approx 5\degr$) to the more distant trailing tail ($d \approx 22~\kpc$ at $\phi_1 \approx -5\degr$).
+We also detect a distance trend from the closer leading tail ($d \approx 19.4~\kpc$ at $\phi_1 \approx 5\degr$) to the more distant trailing tail ($d \approx 21.3~\kpc$ at $\phi_1 \approx -5\degr$).
 While the proper motion in right ascension appears consistent with constant or only a mild gradient over the extent of the stream, we see a significant gradient in the declination component, with a difference of about $\approx 1~\masyr$ from trailing to leading tail.
@@
 
-Assuming a Jacobi radius of $11'$---computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.2\times10^4~\msun$ \citep{Kuepper:2015} and using the Milky Way mass model in \citet{gala}---we find 9 RRLs in the cluster itself (2 type~\typeab\ and 7 type~\typec) and 15 in the cluster's tails (11 type~\typeab\ and 4 type~\typec), all  with membership probabilities larger than 0.5.
-In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.2$, a typical ratio found Oosterhoff I populations \citep{Smith:1995}.
+Assuming a Jacobi radius of $11'$---computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.2\times10^4~\msun$ \citep{Kuepper:2015} and using the Milky Way mass model in \citet{gala}---we find \NRRcl\ RRLs in the cluster itself (3 type~\typeab\ and 6 type~\typec) and \NRRtails\ in the cluster's tails (13 type~\typeab\ and 5 type~\typec), all with membership probabilities larger than 0.5.
+In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.5$, \todo{APW: still typical?} a typical ratio found Oosterhoff I populations \citep{Smith:1995}.
 \figurename~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
@@
 
+\todo{APW: update numbers with new catalog}
 To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu:2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters, here including the RR\typec~stars.
@@
 Using a catalog of RRLs identified using multi-epoch, multi-band photometry from the Pan-STARRS PS1 survey and cross-matched to \Gaia\ \DR{2}, we construct a probabilistic model to determine membership with the Pal 5 cluster and stream.
-We find 24 RRLs with membership probability $>0.5$ (13 type~\typeab\ and 11 type~\typec) over $\approx 18\degr$ of the stream.
-From these stars, we infer the distance to the cluster, finding a closer value, $d = 20.9 \pm 0.2~\kpc$, than many past studies, and detect a distance gradient of $\sim 0.4~\kpc~{\rm deg}^{-1}$ between the trailing and leading tail (between $-5^\circ < \phi_1 < 5^\circ$).
+We find \NRRL\ RRLs with membership probability $>0.5$ (\NRRab\ type~\typeab\ and \NRRc\ type~\typec) over $\approx 18\degr$ of the stream.
+From these stars, we infer the distance to the cluster \clderr\ and detect a distance gradient of $\sim 0.2~\kpc~{\rm deg}^{-1}$ between the trailing and leading tail (between $-5^\circ < \phi_1 < 5^\circ$).
 We also, for the first time, detect trends in proper motion along the stream, predicted by models of the stream \citep[e.g.,][]{Pearson:2017}.
 We release the full RRL catalog in the region around Pal 5 (containing distance information from PS1 and astrometry from \Gaia) with our derived membership probabilities.
-These data will be instrumental for improving dynamical models of Pal 5, both for using Pal 5 to constrain the global matter distribution of the Milky Way and in interpreting the observed, complex density structure of the stream.
+We note that past models of the Pal 5 stream have typically used larger cluster distances and slower proper motion values, meaning that any past model streams will be offset from the measurements and analysis presented here \citep[e.g.,][]{Kuepper:2015, Erkal:2017, Pearson:2017}.
+These data will be instrumental for improving and updating dynamical models of Pal 5, both for using Pal 5 to constrain the global matter distribution of the Milky Way and in interpreting the observed, complex density structure of the stream.
 % \sa{not sure where in text, but could be worth mentioning that you don't find stars beyond the current footprint of the stream (i.e.

### 2019-09-04 paper/main.tex
Updates from Overleaf

@@
 \newcommand{\NRRL}{27}     % N total, prob > 0.5
-\newcommand{\NRRab}{16}   % N type ab, total
-\newcommand{\NRRc}{11}    % N type c, total
-\newcommand{\NRRcl}{9}    % N in cluster
-\newcommand{\NRRtails}{18} % N in tails
+% \newcommand{\NRRab}{16}    % N type ab, total
+% \newcommand{\NRRc}{11}     % N type c, total
+% \newcommand{\NRRcl}{9}     % N in cluster
+% \newcommand{\NRRtails}{18} % N in tails
+
+\newcommand{\NRRab}{15}    % N type ab, total
+\newcommand{\NRRc}{12}     % N type c, total
+\newcommand{\NRRcl}{10}     % N in cluster
+\newcommand{\NRRtails}{17} % N in tails
 
@@
 We use the affine-invariant ensemble MCMC sampler \texttt{emcee} \citep{emcee} to sample from the posterior probability distribution given above.
-We run \texttt{emcee} with 104 walkers for an initial 1024 steps to ``burn-in'' the sampler, then reset and restart the sampler for another 65536 steps.
-We estimate the maximum (over parameters) autocorrelation length using \texttt{emcee}, $\tau_{\rm max} \approx 1200$, and thin the chains by keeping only every 1200th step, leaving us with 5616 posterior samples in our membership model parameters, $\bs{\theta}$.
+We run \texttt{emcee} with 104 walkers for an initial 1024 steps to ``burn-in'' the sampler, then reset and restart the sampler for another 131072 steps.
+We estimate the maximum (over parameters) autocorrelation length using \texttt{emcee}, $\tau_{\rm max} \approx 2200$, and thin the chains by keeping only every 2200th step, leaving us with 6196 posterior samples in our membership model parameters, $\bs{\theta}$.
 For each star, we use these samples to compute posterior membership probabilities following \citet{DFM:blog}.
@@
 
-Assuming a Jacobi radius of $11'$---computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.2\times10^4~\msun$ \citep{Kuepper:2015} and using the Milky Way mass model in \citet{gala}---we find \NRRcl\ RRLs in the cluster itself (3 type~\typeab\ and 6 type~\typec) and \NRRtails\ in the cluster's tails (13 type~\typeab\ and 5 type~\typec), all with membership probabilities larger than 0.5.
-In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.5$, \todo{APW: still typical?} a typical ratio found Oosterhoff I populations \citep{Smith:1995}.
+% \todo{APW: still typical?} 
+%CM: yes, Nab/Nc has a long tail towards large values 
+Assuming a Jacobi radius of $11'$---computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.2\times10^4~\msun$ \citep{Kuepper:2015} and using the Milky Way mass model in \citet{gala}---we find \NRRcl\ RRLs in the cluster itself (2 type~\typeab\ and 8 type~\typec) and \NRRtails\ in the cluster's tails (13 type~\typeab\ and 4 type~\typec), all with membership probabilities larger than 0.5.
+In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.25$,a typical ratio found Oosterhoff I populations \citep{Smith:1995}.
 \figurename~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
@@
 Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an Oosterhoff I classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.4$ \citep{Dotter:2011}.
-Out of our newly found \rrab\ RRLs, two are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars, again consistent with Pal~5's metallicity,  as this type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017}.
+Interestingly, there seems to be a strong segregation of the RRL types: all but two of the \rrab~are found in the cluster tails and two thirds of the \rrc~are found in the cluster itself.
+Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars, again consistent with Pal~5's metallicity,  as this type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017}.
 
@@
 
-\todo{APW: update numbers with new catalog}
+%\todo{APW: update numbers with new catalog}
 To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu:2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters, here including the RR\typec~stars.
 The $\log{N_{RR}}-M_V$ relation has a large scatter so, particularly for such low numbers of RRLs, it will only allow us to give rough limits on the cluster's absolute magnitude and, therefore, of its luminosity.
-Correcting the observed number of 24~RRLs by the median completeness of 92\%, we get a total expected number of 26~RRLs, for which we find a total luminosity $L_V=3.9_{-2.5}^{+7.4}\times 10^4 \mathrm{L}_\sun$.%, which correspond to the mode and 16/84th percentiles of the posterior probability distribution.
-We then estimate a stellar mass of $0.8_{-0.5}^{+1.5}\times 10^5 \msun$, by assuming a V-band mass-to-light ratio $M/L=2.$ corresponding to a simple stellar population with the age (12 Gyr) and metallicity ($\FeH=-1.4$) of the cluster \citep{Dotter:2011}, computed from the \citet{Bruzual:2003} models \footnote{We have used Z=0.0004 ($\FeH=-1.6$), the nearest to Pal~5’s metallicity available in the models.}.
+Correcting the observed number of 27~RRLs by the median completeness of 92\%, we get a total expected number of 29~RRLs, for which we find a total luminosity $L_V=4.6_{-3.0}^{+8.8}\times 10^4 \mathrm{L}_\sun$.
+We then estimate a stellar mass of $0.9_{-0.6}^{+1.8}\times 10^5 \msun$, by assuming a V-band mass-to-light ratio $M/L=2$ corresponding to a simple stellar population with the age (12 Gyr) and metallicity ($\FeH=-1.4$) of the cluster \citep{Dotter:2011}, computed from the \citet{Bruzual:2003} models \footnote{We have used Z=0.0004 ($\FeH=-1.6$), the nearest to the cluster's metallicity available in the models.}.
 Our mass estimate is larger than that of \citet{Ibata:2017}, who report a current mass of $1.2\pm 0.4 \times10^4\msun$ for the cluster and tails; although their estimate for the initial mass ($4.7\pm 1.5\times10^4\msun$) of the system is at the lower end of our confidence limits.
@@
 \software{
-    Astropy \citep{astropy, astropy:2018},
+    astropy \citep{astropy, astropy:2018},
     emcee \citep{emcee},
     gala \citep{gala},
-    IPython \citep{ipython}
+    IPython \citep{ipython},
+    matplotlib \citep{mpl},
+    numpy \citep{numpy},
+    scipy \citep{scipy}
 }
@@
 4418920808077110784 & V5 (K62) & 0.337934 & 0.46 & RRc & 20.7 & 1.00 & 2.0 \\
-4418920846732620032 &          & 0.643239 & 0.27 & RRab & 21.7 & 1.00 & 2.0 \\
-4418914863842345856 & V1 (K62) & 0.293229 & 0.56 & RRc & 20.4 & 0.97 & 2.0 \\
+4418920846732620032 &          & 0.643239 & 0.27 & RRab & 21.7 & 0.99 & 2.0 \\
+4418914863842345856 & V1 (K62) & 0.293229 & 0.56 & RRc & 20.4 & 0.95 & 2.0 \\
 4418726027016125056 & V3 (K62) & 0.329948 & 0.51 & RRc & 20.7 & 1.00 & 3.9 \\
-4418725889577171328 & V4 (K62) & 0.286366 & 0.59 & RRc & 20.3 & 0.99 & 4.5 \\
+4418725889577171328 & V4 (K62) & 0.286366 & 0.59 & RRc & 20.3 & 1.00 & 4.5 \\
+4418731829516302848 &          & 0.281640 & 0.30 & RRc & 18.1 & 0.65 & 4.9 \\
 4418913218870688768 & V2 (K62) & 0.332473 & 0.40 & RRc & 20.3 & 1.00 & 5.1 \\
-4418734165978521728 &          & 0.380859 & 0.52 & RRc & 22.8 & 0.92 & 7.7 \\
+4418734165978521728 &          & 0.380859 & 0.52 & RRc & 22.8 & 0.91 & 7.7 \\
 4418724034151291776 & 403 (V04)& 0.551705 & 1.14 & RRab & 21.2 & 1.00 & 9.8 \\
-4418732791593809152 & 405 (V04)& 0.329670 & 0.45 & RRc & 20.7 & 1.00 & 11.0 \\
-4419052204012341760 & 133 (W05)& 0.530006 & 1.01 & RRab & 20.8 & 1.00 & 44.5 \\
-4418142117622280192 & 393 (V04)& 0.462671 & 1.36 & RRab & 20.0 & 0.92 & 118.8 \\
+4418732791593809152 & 405 (V04) & 0.329670 & 0.45 & RRc & 20.7 & 1.00 & 11.0 \\
+4419052204012341760 & 133 (W05) & 0.530006 & 1.01 & RRab & 20.8 & 1.00 & 44.5 \\
+4418142117622280192 & 393 (V04) & 0.462671 & 1.36 & RRab & 20.0 & 0.91 & 118.8 \\
 6339498589346000768 &  & 0.320110 & 0.46 & RRc & 18.9 & 0.99 & 217.8 \\
@@
 4427234700827469952 &  & 0.517911 & 1.24 & RRab & 21.9 & 1.00 & 392.4 \\
-6337302795905934336 &          & 0.519462 & 1.02 & RRab & 17.3 & 0.89 & 404.6 \\
-6337233350579111680 &          & 0.542383 & 0.96 & RRab & 19.0 & 0.93 & 450.7 \\
-4427646704153765632 &          & 0.385552 & 0.37 & RRc & 19.9 & 0.81 & 456.6 \\
-4424705647988505600 &          & 0.335502 & 0.41 & RRc & 20.9 & 0.99 & 513.0 \\
-4426221707021159424 &          & 0.533313 & 1.14 & RRab & 22.0 & 0.93 & 550.1 \\
+6337233350579111680 &  & 0.542383 & 0.96 & RRab & 19.0 & 0.98 & 450.7 \\
+4427646704153765632 &  & 0.385552 & 0.37 & RRc & 19.9 & 0.87 & 456.6 \\
+4424705647988505600 &  & 0.335502 & 0.41 & RRc & 20.9 & 1.00 & 513.0 \\
+4426221707021159424 &  & 0.533313 & 1.14 & RRab & 22.0 & 0.99 & 550.1 \\
+4450058431917932672 &  & 0.712626 & 1.15 & RRab & 21.1 & 0.57 & 857.8 \\
+4439927909737905280 &  & 0.724974 & 0.91 & RRab & 22.2 & 0.85 & 1124.3 \\
+4445607604551159040 &  & 0.456636 & 1.40 & RRab & 22.3 & 0.81 & 1394.3 \\
 \bottomrule

### 2019-09-04 paper/main.tex
whitespace in paper and update noteobok
Updates from Overleaf

@@
 \newcommand{\sa}[1]{{\color{teal} SP: #1}}
+\newcommand{\ab}[1]{{\color{magenta} AB: #1}}
 
@@
 
-\figurename~\ref{fig:trackmembers} shows a summary of the inferred membership model properties in sky position (shown in Pal 5 coordinates $\phi_1, \phi_2$), distance, and proper motion components.
+\figurename~\ref{fig:trackmembers} shows a summary of the inferred membership model properties in sky position (shown in Pal 5 coordinates $\phi_1, \phi_2$), distance, $d$, and proper motion components.
 \tablename~\ref{t:rrl_members} shows an abbreviated version of our catalog of RRL in the region around Pal 5, with PS1 and \Gaia~ data and membership probabilities appended.
@@
 No theoretical models of Pal 5 have predicted tidal debris at such locations, so follow-up spectroscopy to confirm or rule out their origin would be very informative.
+\ab{mention RRL overdensities along the tails don't correspond to MS overdensities?}
 
@@
 
+\ab{this section is mostly about the distances -- include in the subtitle?}
 Past distance measurements to the Pal 5 cluster have ranged from $20.9~\kpc$ \citep{Dotter:2011} to $23.2~\kpc$ \citep{Harris:1996}, with a previous measurement from a small number of RRLs in between at $22.3~\kpc$ \citep{Vivas:2006}.
@@
 This is closer to the sun than most past measurements of the cluster distance, but consistent with the measurement from \citet{Dotter:2011}.
+\ab{can we work this point a bit more?
+I thought it was mainly the HST CMD measurement from Aaron that was an outlier, and everything else at a larger distance.
+then I think in Nitya + Tobias paper they found that to match pms, they require a smaller distance, and now we find it in PS1 RRL.
+I don't think we need to recount the whole history in a letter, but it might be helpful to comment on the difference wrt Kathy's RRL result}
 We find a proper motion $(\mu_{\phi_1}\cos\phi_2, \mu_{\phi_2}) = (3.78, 0.71) \pm (0.02, 0.03)~\masyr$ or $(\mu_\alpha\,\cos\delta, \mu_\delta) = (-2.75, -2.69)~\masyr$, consistent with the measurement in \citet{Vasiliev:2019} within our uncertainties.
@@
 \figurename~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
+\ab{can we provide a sentence for what it (physically) means to be an oosterhoff I population?}
 The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram.
@@
 Interestingly, there seems to be a strong segregation of the RRL types: all but two of the \rrab~are found in the cluster tails and two thirds of the \rrc~are found in the cluster itself.
+\ab{this is interesting indeed -- can we provide a hypothesis?}
 Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars, again consistent with Pal~5's metallicity,  as this type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017}.
@@
 \caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field membership probabilities $>0.1$ (small markers) inside the Pal~5 track (large markers), colored by membership probability.
-The distribution of the RRLs found in the cluster and it's tails is consistent with an Oosterhoff type I (dashed line).
+The distribution of the RRLs found in the cluster and its tails is consistent with an Oosterhoff type I (dashed line).
 }

### 2019-09-05 paper/main.tex
add bonaca and update distance info

@@
 \documentclass[twocolumn]{aastex63}
+
+% typography
+\usepackage[T1]{fontenc}
+\setlength{\parindent}{1.\baselineskip}
 
@@
 
-\begin{document}
+\begin{document}\sloppy\sloppypar\raggedbottom\frenchspacing % trust Hogg
 
@@
              Simons Foundation, 162 Fifth Avenue, New York, NY 10010, USA}
+
+\author[0000-0002-7846-9787]{Ana Bonaca}
+\affiliation{Center for Astrophysics | Harvard \& Smithsonian, 60 Garden St, Cambridge, MA 02138, USA}
 
@@
 We find a total of \NRRcl\ RRLs within the estimated Jacobi radius of the cluster ($\approx 11'$), meaning that \NRRtails\ RRLs are spread between the leading and trailing tails.
-\figurename~\ref{fig:members} shows the sky positions of all RRLs with membership probability $>0.1$ over-plotted on the density of filtered main sequence stars from \citet{Bonaca:2019}.
-We note that there are a few RRLs outside of the main stream track that appear to be at a consistent distance with consistent proper motions (e.g., the two stars near $\alpha = 230\degr$, $\delta = 0\degr$ in \figurename~\ref{fig:members}) that are not included in the tally above.
+\figurename~\ref{fig:members} shows the sky positions of all RRLs with membership probability $>0.1$ over-plotted on the density of filtered main sequence stars from \citet{Bonaca:2019}:
+The RRLs trace the stream track defined by the main sequence stars, but interestingly the density of RRLs does not seem to match the density variations in the under-plotted density map.
+In particular, the bulk of the stream RRLs seem to be spread farther out (in angle) from the cluster relative to the main sequence stars, a possible manifestation of the internal dynamics of the initial cluster \citep[e.g., mass segregation][]{Koch:2004}.
+
+We note also that there are a few RRLs outside of the main stream track that appear to be at a consistent distance with consistent proper motions (e.g., the two stars near $\alpha = 230\degr$, $\delta = 0\degr$ in \figurename~\ref{fig:members}) that are not included in the tally above.
 No theoretical models of Pal 5 have predicted tidal debris at such locations, so follow-up spectroscopy to confirm or rule out their origin would be very informative.
-\ab{mention RRL overdensities along the tails don't correspond to MS overdensities?}
 
-\subsection{Kinematics of the cluster and stream}
+\subsection{Distance and kinematics of the cluster and stream}
 
-\ab{this section is mostly about the distances -- include in the subtitle?}
-Past distance measurements to the Pal 5 cluster have ranged from $20.9~\kpc$ \citep{Dotter:2011} to $23.2~\kpc$ \citep{Harris:1996}, with a previous measurement from a small number of RRLs in between at $22.3~\kpc$ \citep{Vivas:2006}.
 We find a mean Pal 5 cluster distance of \clderr\ using the PS1 RRLs and the membership model described in \sectionname~\ref{sec:membership}.
-This is closer to the sun than most past measurements of the cluster distance, but consistent with the measurement from \citet{Dotter:2011}.
-\ab{can we work this point a bit more?
-I thought it was mainly the HST CMD measurement from Aaron that was an outlier, and everything else at a larger distance.
-then I think in Nitya + Tobias paper they found that to match pms, they require a smaller distance, and now we find it in PS1 RRL.
-I don't think we need to recount the whole history in a letter, but it might be helpful to comment on the difference wrt Kathy's RRL result}
+While past distance measurements to Pal 5 are often reported ranging from $\sim 22.4~\kpc$ to $>23~\kpc$ (e.g., \citealt{Odenkirchen:2009, Erkal:2017}, using measurements from \citealt{Harris:1996, Dotter:2011}), these values are computed from distance modul that have not been corrected for dust extinction (\citealt{Harris:1996}, and see footnote 5 in \citealt{Fritz:2015}).
+Using the recalibrated \citet{Schlegel:1998} dustmap from \citet{Schlafly:2011}, and assuming $R_V=3.1$, we find in the region around Pal 5, $A_V = 0.17~\textrm{mag}$.
+Re-deriving distances based on published distance moduli, we find $d_{\rm Pal 5} = 21~\kpc$ \citep{Harris:1996} and $d_{\rm Pal 5} = 20.7~\kpc$ \citep{Dotter:2011}.
+Assuming a modest distance uncertainty of $\approx 5\%$ for these isochronal distances, our measurement is consistent with these past measurements.
+
+Using a small number of RRLs then thought to be associated with Pal 5, \citet{Vivas:2006} measured a distance to the cluster of $d = 22.3~\kpc$.
+However, this used a now imprecise $M_V$--$\feh$ relation to compute the distance modulus.
+Adopting $\feh = -1.35$ \citep{Ishigaki:2016} and using the updated RRL $M_V$--$\feh$ relation from \citet{Muraveva:2018}, we find instead that this past RRL-based distance to Pal 5 is instead $d = 21.2 \pm 0.6~\kpc$.
+
 We find a proper motion $(\mu_{\phi_1}\cos\phi_2, \mu_{\phi_2}) = (3.78, 0.71) \pm (0.02, 0.03)~\masyr$ or $(\mu_\alpha\,\cos\delta, \mu_\delta) = (-2.75, -2.69)~\masyr$, consistent with the measurement in \citet{Vasiliev:2019} within our uncertainties.
@@
     astropy \citep{astropy, astropy:2018},
+    dustmaps \citep{dustmaps},
     emcee \citep{emcee},

### 2019-09-05 paper/main.tex
Updates from Overleaf

@@
 However, this used a now imprecise $M_V$--$\feh$ relation to compute the distance modulus.
-Adopting $\feh = -1.35$ \citep{Ishigaki:2016} and using the updated RRL $M_V$--$\feh$ relation from \citet{Muraveva:2018}, we find instead that this past RRL-based distance to Pal 5 is instead $d = 21.2 \pm 0.6~\kpc$.
+Adopting $\feh = -1.35$ \citep{Ishigaki:2016} and using the updated RRL $M_V$--$\feh$ relation from \citet{Muraveva:2018}, we find that this past RRL-based distance to Pal 5 is instead $d_{\rm Pal 5} = 21.2 \pm 0.6~\kpc$.
+To summarize: Our inferred distance to the Pal 5 cluster, $d_{\rm Pal 5} = \clderr$, is consistent with (but more precise than) all recent distance measurements, once corrected for dust extinction and updated calibrations of the horizontal branch absolute magnitude.
 
@@
 Assuming a Jacobi radius of $11'$---computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.2\times10^4~\msun$ \citep{Kuepper:2015} and using the Milky Way mass model in \citet{gala}---we find \NRRcl\ RRLs in the cluster itself (2 type~\typeab\ and 8 type~\typec) and \NRRtails\ in the cluster's tails (13 type~\typeab\ and 4 type~\typec), all with membership probabilities larger than 0.5.
-In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.25$,a typical ratio found Oosterhoff I populations \citep{Smith:1995}.
+In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.25$, a typical ratio found in Oosterhoff I populations \citep{Smith:1995},  comprised mainly of stars still in the zero-age horizontal branch.
 \figurename~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
-\ab{can we provide a sentence for what it (physically) means to be an oosterhoff I population?}
 The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram.
@@
 Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an Oosterhoff I classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.4$ \citep{Dotter:2011}.
-Interestingly, there seems to be a strong segregation of the RRL types: all but two of the \rrab~are found in the cluster tails and two thirds of the \rrc~are found in the cluster itself.
-\ab{this is interesting indeed -- can we provide a hypothesis?}
 Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars, again consistent with Pal~5's metallicity,  as this type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017}.
 
-%.
-and the Sagittarius dwarf galaxy, but not in any other dwarf spheroidal galaxy.
-\todo{CM:( any other point to be made about this?)}.
-\sa{as someone who doesn't know much about RRLs, how weird is the above?
-maybe some more discussion/elaboration?
-} \cm{well, it's not weird at all.
-I've added a sentence saying so.}
+Interestingly, there seems to be a strong segregation of the RRL types: all but two of the \NRRab\ \rrab~are found in the cluster tails and two thirds of the \NRRc\ \rrc~are found in the cluster itself.
+At fixed metallicity and helium abundance, position along the horizontal branch is determined by (present-day) mass, with less massive stars ending up at lower temperatures as type \typeab~stars.
+The dichotomy of RRL types between the stream and cluster is therefore suggestive of being a signature of mass segregation: The (slightly) higher mass \rrc\ stars seem to be preferentially more bound to the cluster.
+However, the mass difference between \rrab\ and \rrc\ stars is expected to be small, $\lesssim 0.1~\msun$, and so unlikely to have a significantly different dynamical history---this remains an intriguing puzzle.
 
@@
 
-%\todo{APW: update numbers with new catalog}
-To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu:2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters, here including the RR\typec~stars.
+To estimate the expected total absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu:2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters, here including the RR\typec~stars.
 The $\log{N_{RR}}-M_V$ relation has a large scatter so, particularly for such low numbers of RRLs, it will only allow us to give rough limits on the cluster's absolute magnitude and, therefore, of its luminosity.
@@
 \acknowledgments
+
 This work was performed in part during the Gaia19 workshop and the 2019 Santa Barbara Gaia Sprint (also supported by the Heising-Simons Foundation), both hosted by the Kavli Institute for Theoretical Physics at the University of California, Santa Barbara.
@@
 }
-\clearpage
+
 \appendix

### 2019-09-05 paper/main.tex
few tweaks

@@
 We find a mean Pal 5 cluster distance of \clderr\ using the PS1 RRLs and the membership model described in \sectionname~\ref{sec:membership}.
-While past distance measurements to Pal 5 are often reported ranging from $\sim 22.4~\kpc$ to $>23~\kpc$ (e.g., \citealt{Odenkirchen:2009, Erkal:2017}, using measurements from \citealt{Harris:1996, Dotter:2011}), these values are computed from distance modul that have not been corrected for dust extinction (\citealt{Harris:1996}, and see footnote 5 in \citealt{Fritz:2015}).
+While past distance measurements to Pal 5 are often reported ranging from $\sim 22.4~\kpc$ to $>23~\kpc$ (e.g., \citealt{Odenkirchen:2009, Erkal:2017}, using measurements from \citealt{Harris:1996, Dotter:2011}), these values are computed from distance moduli that have not been corrected for dust extinction (\citealt{Harris:1996}, and see footnote 5 in \citealt{Fritz:2015}).
 Using the recalibrated \citet{Schlegel:1998} dustmap from \citet{Schlafly:2011}, and assuming $R_V=3.1$, we find in the region around Pal 5, $A_V = 0.17~\textrm{mag}$.
@@
 We find a proper motion $(\mu_{\phi_1}\cos\phi_2, \mu_{\phi_2}) = (3.78, 0.71) \pm (0.02, 0.03)~\masyr$ or $(\mu_\alpha\,\cos\delta, \mu_\delta) = (-2.75, -2.69)~\masyr$, consistent with the measurement in \citet{Vasiliev:2019} within our uncertainties.
-The larger proper motion (relative to \citealt{Fritz:2015} and other earlier measurements) and the updated distance means that the tangential velocity of Pal 5 is $\approx 16~\kms$ larger than most previous models have assumed.
+The larger (absolute) proper motion (relative to \citealt{Fritz:2015} and other earlier measurements) and the updated distance means that the tangential velocity of Pal 5 is $\approx 16~\kms$ larger than most previous models have assumed.
 This is a noteworthy update to consider for future stream modeling efforts and will likely change the inferred Milky Way parameters as compared to past models \citep[e.g.,][]{Kuepper:2015, Bovy:2016}.
@@
 Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an Oosterhoff I classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.4$ \citep{Dotter:2011}.
-Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars, again consistent with Pal~5's metallicity,  as this type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017}.
+Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_g>0.75$~mag) stars, again consistent with Pal~5's metallicity, as this type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017}.
 
@@
 Correcting the observed number of 27~RRLs by the median completeness of 92\%, we get a total expected number of 29~RRLs, for which we find a total luminosity $L_V=4.6_{-3.0}^{+8.8}\times 10^4 \mathrm{L}_\sun$.
-We then estimate a stellar mass of $0.9_{-0.6}^{+1.8}\times 10^5 \msun$, by assuming a V-band mass-to-light ratio $M/L=2$ corresponding to a simple stellar population with the age (12 Gyr) and metallicity ($\FeH=-1.4$) of the cluster \citep{Dotter:2011}, computed from the \citet{Bruzual:2003} models \footnote{We have used Z=0.0004 ($\FeH=-1.6$), the nearest to the cluster's metallicity available in the models.}.
-Our mass estimate is larger than that of \citet{Ibata:2017}, who report a current mass of $1.2\pm 0.4 \times10^4\msun$ for the cluster and tails; although their estimate for the initial mass ($4.7\pm 1.5\times10^4\msun$) of the system is at the lower end of our confidence limits.
+We then estimate a stellar mass of $0.9_{-0.6}^{+1.8}\times 10^5 \msun$, by assuming a V-band mass-to-light ratio $M/L=2$ corresponding to a simple stellar population with the age (12 Gyr) and metallicity ($\FeH=-1.4$) of the cluster \citep{Dotter:2011}, computed from the \citet{Bruzual:2003} models.\footnote{We have used Z=0.0004 ($\FeH=-1.6$), the nearest to the cluster's metallicity available in the models.}
+Our mass estimate is larger than that of \citet{Ibata:2017}, who report a current mass of $1.2\pm 0.4 \times10^4\msun$ for the cluster and tails, although their estimate for the initial mass ($4.7\pm 1.5\times10^4\msun$) of the system is at the lower end of our confidence limits.
 
@@
 
-The Pal~5 cluster has 5 previously known RRL stars (named V1--V5), all of type which are \typec, as reported in the \citet{Clement:2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{Kinman:1962}.
+The Pal~5 cluster has 5 previously known RRL stars (named V1--V5), all of type which are type-\typec, as reported in the \citet{Clement:2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{Kinman:1962}.
 In their analysis of the halo density profile with QUEST RRLs, \citet{Vivas:2006} report an overdensity (``Group~6'') that they claim could be associated with Pal~5's tidal tails.
@@
 We note that past models of the Pal 5 stream have typically used larger cluster distances and slower proper motion values, meaning that any past model streams will be offset from the measurements and analysis presented here \citep[e.g.,][]{Kuepper:2015, Erkal:2017, Pearson:2017}.
-These data will be instrumental for improving and updating dynamical models of Pal 5, both for using Pal 5 to constrain the global matter distribution of the Milky Way and in interpreting the observed, complex density structure of the stream.
+These data will be instrumental for improving and updating dynamical models of Pal 5, both for using Pal 5 to constrain the global matter distribution of the Milky Way and in interpreting the observed complex density structure of the stream.
 % \sa{not sure where in text, but could be worth mentioning that you don't find stars beyond the current footprint of the stream (i.e.
@@
 
+\clearpage
 \appendix

### 2019-09-30 paper/main.tex
showing stars in jacobi radius differently

@@
 \newcommand{\ab}[1]{{\color{magenta} AB: #1}}
+\newcommand{\changes}[1]{{\textbf{#1}}}
 
@@
 
-Stellar streams are useful objects because they almost \citep[see][]{Sanders:2013} delineate the past and future trajectory of their progenitor system, offering information about orbits that can be used to infer the mass distribution from the single kinematic snapshot of the Galaxy that we are afforded \citep[e.g.,][]{Johnston:1999, Sanders:2013, PriceWhelan:2014, Bonaca:2018}.
+Stellar streams are useful objects because they almost \citep[see][]{Sanders:2013} delineate the past and future trajectory of their progenitor system, offering information about orbits that can be used to infer the mass distribution from the single kinematic snapshot of the Galaxy that we are afforded \citep[e.g.,][]{Johnston:1999, Sanders:2013, PriceWhelan:2014, Bonaca:2018, Malhan:2019, Erkal:2019}.
 % Binney:2008, Eyre:2009,
@@
 
+\begin{figure*}[t!]
+\begin{center}
+\includegraphics[width=0.95\textwidth, trim=0 20px 0 0]{proper-motion.pdf}
+\caption{\textbf{Left}: Proper motions of RRLs in the region $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$ with distances in the range $18 < d < 25~\kpc$.
+The distance selection here is purely illustrative: We do not use a distance cut in our membership model.
+\textbf{Middle}: The same, but for RRLs within $1\degr$ of the mean sky track of Pal 5 \citep{Bonaca:2019}.
+The square (blue) markers indicate RRLs that lie within the Jacobi radius of the cluster (see \sectionname~\ref{sec:results}).
+The over-density of stars near $(-2.7, -2.7)$ is the cluster and stream, and the two colored points show the Pal 5 cluster proper motions from \citet{Vasiliev:2019} and \citet{Fritz:2015}, respectively.
+\textbf{Right}: The same, but for RRLs excluding $2\degr$ around the Pal 5 sky track.}
+\label{fig:pm}
+\end{center}
+\end{figure*}
+
 In this Article, we construct a probabilistic model for kinematic properties of RRLs in the vicinity of the Pal 5 stream to determine membership probabilities and thus map the distance and proper motion trends in the stream.
@@
 \section{Data} \label{sec:data}
-
-\begin{figure*}[t!]
-\begin{center}
-\includegraphics[width=\textwidth]{proper-motion.pdf}
-\caption{\textbf{Left}: Proper motions of RRLs in the region $215^\circ < \alpha < 255^\circ$, $-15^\circ < \delta < 10^\circ$ with distances in the range $18 < d < 25~\kpc$.
-The distance selection here is purely illustrative: We do not use a distance cut in our membership model.
-\textbf{Middle}: The same, but for RRLs within $1\degr$ of the mean sky track of Pal 5 \citep{Bonaca:2019}.
-The over-density of stars near $(-2.7, -2.7)$ is the stream, and the two colored points show the Pal 5 cluster proper motions from \citet{Vasiliev:2019} and \citet{Fritz:2015}, respectively.
-\textbf{Right}: The same, but for RRLs excluding $2\degr$ around the Pal 5 sky track.}
-\label{fig:pm}
-\end{center}
-\end{figure*}
 
@@
 Although not all PS1 RRLs have been identified as RRLs from \Gaia\ photometry, all of the PS1 RRLs are present in the main point source catalog.
-\figurename~\ref{fig:pm} demonstrates that Pal 5 stream members appear in the RRL catalog: This figure shows the \Gaia\ \DR{2} proper motions for all RRLs in the selected sky window (indicated in the caption) within the distance range $18 < d < 25~\kpc$, then the same for RRLs in a tight selection around the observed stream track (see \sectionname~\ref{sec:membership} for details), and then finally the same for RRLs \emph{excluding} the stream region.
+\figurename~\ref{fig:pm} demonstrates that Pal 5 cluster and stream members appear in the RRL catalog: This figure shows the \Gaia\ \DR{2} proper motions for all RRLs in the selected sky window (indicated in the caption) within the distance range $18 < d < 25~\kpc$, then the same for RRLs in a tight selection around the observed stream track (see \sectionname~\ref{sec:membership} for details), and then finally the same for RRLs \emph{excluding} the stream region.
 

### 2019-09-30 paper/main.tex
more response to ref

@@
 % Results
-By building a probabilistic model of the Pal 5 cluster and stream in proper motion and distance, we find \NRRL\ RRLs consistent with being members of the cluster (\NRRcl) and stream (\NRRtails).
+By building a probabilistic model of the Pal 5 \changes{cluster and stream} in proper motion and distance, we find \NRRL\ RRLs consistent with being members of the cluster (\NRRcl) and stream (\NRRtails).
 Using these RRLs, we detect gradients in distance and proper motion along the stream, and provide an updated measurement of the distance to the Pal 5 cluster using the RRLs, $d = \clderr$.
@@
 To start, we use the sky track and width of Pal 5 from \citet{Bonaca:2019} to mask out a wide region around the stream---and an extrapolation of the known extent of the stream---in order to define a ``background'' region that should be devoid of Pal 5 member stars (see purple lines in \figurename~\ref{fig:members}).
-We use RRL stars in the background region to construct an error-deconvolved density model in the 3D space of proper motion components and distance.
+\changes{We use 1,301 RRL stars in the background region (the full rectangular sky window excluding the stream track) to construct an error-deconvolved density model in the 3D space of proper motion components and distance.}
 We use proper motions and uncertainties from \Gaia---using the provided covariance matrix between proper motion components---and distances and uncertainties from the PS1 RRL catalog (\sectionname~\ref{sec:data}).
@@
 \includegraphics[width=\textwidth]{tracks.pdf}
-\caption{RRL stars in the vicinity of the Pal 5 stream (with $\textrm{prob.} > 0.01$), colored by membership probability, and under-plotted with the inferred stream trends used in the membership model.
+\caption{RRL stars in the vicinity of the Pal 5 stream colored by membership probability and under-plotted with the inferred stream trends used in the membership model.
 We note again that the membership probability does not depend on position or proximity to the stream, and only on distance and proper motions.
+\changes{Stars with probability $<0.1$ are plotted with a smaller marker in all panels.}
 \textbf{Upper left}: Sky positions (in the Pal 5 stream coordinate frame) of RRLs, with arrows showing the \Gaia\ \DR{2} proper motion vector direction.
@@
 \includegraphics[width=\textwidth]{members.pdf}
-\caption{Sky positions of RRLs with Pal 5 membership probabilities $>0.1$, colored by membership probability (markers).
+\caption{Sky positions of RRLs with Pal 5 colored by membership probability, \changes{but stars with probability $<0.1$ are plotted with a smaller marker.}
 The background greyscale shows the surface density of CMD-filtered main sequence stars from \citet{Bonaca:2019}.
@@
 where $p(\bs{\theta}, f)$ is the prior probability distribution over the parameters.
-We assume that our prior is separable in all of our parameters: We use a uniform prior for the mixture weight, $f$, such that $p(f) = \mathcal{U}(0, 1)$, Gaussian priors on the mean proper motion and distance using the Pal 5 cluster proper motion measurements from \citet{Vasiliev:2019} and the Pal 5 distance from \citet{Kuepper:2015}.
+We assume that our prior is separable in all of our parameters.
+We adopt a uniform prior for the mixture weight, $f$, such that $p(f) = \mathcal{U}(0, 1)$.
+\changes{We use Gaussian priors on the mean proper motion and distance using the Pal 5 cluster proper motion measurements from \citet{Vasiliev:2019}, $\bs{\mu} = (-2.728, -2.687) \pm (0.022	0.025)~\masyr$ with a correlation coefficient $\rho_\mu = -0.39$, and the Pal 5 distance from \citet{Kuepper:2015}, $\bs{d} = 23.6 \pm 0.8~\kpc$.
+However, we inflate the distance errorbar by a factor of 5 and the proper motion errorbars by a factor of 10 to soften the impact of this prior on our analysis.}
+
 We use broad uniform priors (i.e.
@@
 Correcting the observed number of 27~RRLs by the median completeness of 92\%, we get a total expected number of 29~RRLs, for which we find a total luminosity $L_V=4.6_{-3.0}^{+8.8}\times 10^4 \mathrm{L}_\sun$.
-We then estimate a stellar mass of $0.9_{-0.6}^{+1.8}\times 10^5 \msun$, by assuming a V-band mass-to-light ratio $M/L=2$ corresponding to a simple stellar population with the age (12 Gyr) and metallicity ($\FeH=-1.4$) of the cluster \citep{Dotter:2011}, computed from the \citet{Bruzual:2003} models.\footnote{We have used Z=0.0004 ($\FeH=-1.6$), the nearest to the cluster's metallicity available in the models.}
+We then estimate a stellar mass of $0.9_{-0.6}^{+1.8}\times 10^5 \msun$, by assuming a V-band mass-to-light ratio $M/L \approx 2$ \citep{Baumgardt:2019} corresponding to a simple stellar population with the age (12 Gyr) and metallicity ($\FeH=-1.4$) of the cluster \citep{Dotter:2011}, computed from the \citet{Bruzual:2003} models.\footnote{We have used Z=0.0004 ($\FeH=-1.6$), the nearest to the cluster's metallicity available in the models.}
 Our mass estimate is larger than that of \citet{Ibata:2017}, who report a current mass of $1.2\pm 0.4 \times10^4\msun$ for the cluster and tails, although their estimate for the initial mass ($4.7\pm 1.5\times10^4\msun$) of the system is at the lower end of our confidence limits.

### 2019-09-30 paper/main.tex
Updates from Overleaf

--- 2019-09-05 paper/main.tex
@@
 We find a mean Pal 5 cluster distance of \clderr\ using the PS1 RRLs and the membership model described in \sectionname~\ref{sec:membership}.
-While past distance measurements to Pal 5 are often reported ranging from $\sim 22.4~\kpc$ to $>23~\kpc$ (e.g., \citealt{Odenkirchen:2009, Erkal:2017}, using measurements from \citealt{Harris:1996, Dotter:2011}), these values are computed from distance moduli that have not been corrected for dust extinction (\citealt{Harris:1996}, and see footnote 5 in \citealt{Fritz:2015}).
+While past distance measurements to Pal 5 are often reported ranging from $\sim 22.4~\kpc$ to $>23~\kpc$ (e.g., \citealt{Odenkirchen:2009, Erkal:2017}, using measurements from \citealt{Harris:1996, Dotter:2011}), these values are computed from distance modul that have not been corrected for dust extinction (\citealt{Harris:1996}, and see footnote 5 in \citealt{Fritz:2015}).
 Using the recalibrated \citet{Schlegel:1998} dustmap from \citet{Schlafly:2011}, and assuming $R_V=3.1$, we find in the region around Pal 5, $A_V = 0.17~\textrm{mag}$.
@@
 We find a proper motion $(\mu_{\phi_1}\cos\phi_2, \mu_{\phi_2}) = (3.78, 0.71) \pm (0.02, 0.03)~\masyr$ or $(\mu_\alpha\,\cos\delta, \mu_\delta) = (-2.75, -2.69)~\masyr$, consistent with the measurement in \citet{Vasiliev:2019} within our uncertainties.
-The larger (absolute) proper motion (relative to \citealt{Fritz:2015} and other earlier measurements) and the updated distance means that the tangential velocity of Pal 5 is $\approx 16~\kms$ larger than most previous models have assumed.
+The larger proper motion (relative to \citealt{Fritz:2015} and other earlier measurements) and the updated distance means that the tangential velocity of Pal 5 is $\approx 16~\kms$ larger than most previous models have assumed.
 This is a noteworthy update to consider for future stream modeling efforts and will likely change the inferred Milky Way parameters as compared to past models \citep[e.g.,][]{Kuepper:2015, Bovy:2016}.
@@
 Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an Oosterhoff I classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.4$ \citep{Dotter:2011}.
-Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_g>0.75$~mag) stars, again consistent with Pal~5's metallicity, as this type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017}.
+Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars, again consistent with Pal~5's metallicity,  as this type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017}.
 
@@
 Correcting the observed number of 27~RRLs by the median completeness of 92\%, we get a total expected number of 29~RRLs, for which we find a total luminosity $L_V=4.6_{-3.0}^{+8.8}\times 10^4 \mathrm{L}_\sun$.
-We then estimate a stellar mass of $0.9_{-0.6}^{+1.8}\times 10^5 \msun$, by assuming a V-band mass-to-light ratio $M/L=2$ corresponding to a simple stellar population with the age (12 Gyr) and metallicity ($\FeH=-1.4$) of the cluster \citep{Dotter:2011}, computed from the \citet{Bruzual:2003} models.\footnote{We have used Z=0.0004 ($\FeH=-1.6$), the nearest to the cluster's metallicity available in the models.}
-Our mass estimate is larger than that of \citet{Ibata:2017}, who report a current mass of $1.2\pm 0.4 \times10^4\msun$ for the cluster and tails, although their estimate for the initial mass ($4.7\pm 1.5\times10^4\msun$) of the system is at the lower end of our confidence limits.
+We then estimate a stellar mass of $0.9_{-0.6}^{+1.8}\times 10^5 \msun$, by assuming a V-band mass-to-light ratio $M/L=2$ corresponding to a simple stellar population with the age (12 Gyr) and metallicity ($\FeH=-1.4$) of the cluster \citep{Dotter:2011}, computed from the \citet{Bruzual:2003} models \footnote{We have used Z=0.0004 ($\FeH=-1.6$), the nearest to the cluster's metallicity available in the models.}.
+Our mass estimate is larger than that of \citet{Ibata:2017}, who report a current mass of $1.2\pm 0.4 \times10^4\msun$ for the cluster and tails; although their estimate for the initial mass ($4.7\pm 1.5\times10^4\msun$) of the system is at the lower end of our confidence limits.
 
@@
 
-The Pal~5 cluster has 5 previously known RRL stars (named V1--V5), all of type which are type-\typec, as reported in the \citet{Clement:2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{Kinman:1962}.
+The Pal~5 cluster has 5 previously known RRL stars (named V1--V5), all of type which are \typec, as reported in the \citet{Clement:2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{Kinman:1962}.
 In their analysis of the halo density profile with QUEST RRLs, \citet{Vivas:2006} report an overdensity (``Group~6'') that they claim could be associated with Pal~5's tidal tails.
@@
 We note that past models of the Pal 5 stream have typically used larger cluster distances and slower proper motion values, meaning that any past model streams will be offset from the measurements and analysis presented here \citep[e.g.,][]{Kuepper:2015, Erkal:2017, Pearson:2017}.
-These data will be instrumental for improving and updating dynamical models of Pal 5, both for using Pal 5 to constrain the global matter distribution of the Milky Way and in interpreting the observed complex density structure of the stream.
+These data will be instrumental for improving and updating dynamical models of Pal 5, both for using Pal 5 to constrain the global matter distribution of the Milky Way and in interpreting the observed, complex density structure of the stream.
 % \sa{not sure where in text, but could be worth mentioning that you don't find stars beyond the current footprint of the stream (i.e.
@@
 CM acknowledges funding from the MIA program at Universidad de la Republica, Uruguay, and is grateful for the hospitality and support of the Flatiron Institute where part of this research was carried out.
+CM thanks the PAPIIT program from DGAPA/UNAM for their support
+through grant IG100319.
 
@@
 
-\clearpage
 \appendix

### 2019-09-30 paper/main.tex
fix changes

--- 2019-09-30 paper/main.tex
@@
 % Results
-By building a probabilistic model of the Pal 5 \changes{cluster and stream} in proper motion and distance, we find \NRRL\ RRLs consistent with being members of the cluster (\NRRcl) and stream (\NRRtails).
+By building a probabilistic model of the Pal 5 cluster and stream in proper motion and distance, we find \NRRL\ RRLs consistent with being members of the cluster (\NRRcl) and stream (\NRRtails).
 Using these RRLs, we detect gradients in distance and proper motion along the stream, and provide an updated measurement of the distance to the Pal 5 cluster using the RRLs, $d = \clderr$.
@@
 Although not all PS1 RRLs have been identified as RRLs from \Gaia\ photometry, all of the PS1 RRLs are present in the main point source catalog.
-\figurename~\ref{fig:pm} demonstrates that Pal 5 cluster and stream members appear in the RRL catalog: This figure shows the \Gaia\ \DR{2} proper motions for all RRLs in the selected sky window (indicated in the caption) within the distance range $18 < d < 25~\kpc$, then the same for RRLs in a tight selection around the observed stream track (see \sectionname~\ref{sec:membership} for details), and then finally the same for RRLs \emph{excluding} the stream region.
+\figurename~\ref{fig:pm} demonstrates that Pal 5 \changes{cluster and stream} members appear in the RRL catalog: This figure shows the \Gaia\ \DR{2} proper motions for all RRLs in the selected sky window (indicated in the caption) within the distance range $18 < d < 25~\kpc$, then the same for RRLs in a tight selection around the observed stream track (see \sectionname~\ref{sec:membership} for details), and then finally the same for RRLs \emph{excluding} the stream region.
 

### 2019-09-30 paper/main.tex
Updates from Overleaf

--- 2019-09-30 paper/main.tex
@@
 \citet{Wu:2005} reports another RRL star (133) as possibly associated, which we also find is a highly probable member.
-More recently, \citet{Ibata:2017} use a prior iteration of the PS1 RRL catalog \citep{Hernitschek:2016}, which contained \emph{plausible} RRLs, to identify 6 and 12 candidate members of the Pal 5 cluster and tails, respectively, based on sky position and distance alone.
+\changes{In summary, nine of the RRL stars in our sample were identified previously as candidate members of the Pal 5 cluster or stream (see the ``Other ID'' column in \tablename~\ref{t:rrl_members}).} More recently, \citet{Ibata:2017} use a prior iteration of the PS1 RRL catalog \citep{Hernitschek:2016}, which contained \emph{plausible} RRLs, to identify 6 and 12 candidate members of the Pal 5 cluster and tails, respectively, based on sky position and distance alone.
 However, without a published catalog, we cannot compare our sample to theirs.

### 2019-09-30 paper/main.tex

@@
 Note that the membership probability does not incorporate sky position information, but a clear over-density of high-probability RRL members do trace the stream on the sky.
-Purple lines outline the stream track window defined in \citet{Bonaca:2019} and used here to exclude stream stars when constructing the background model.}
+Purple lines outline the stream track window defined in \citet{Bonaca:2019} and used here to exclude stream stars when constructing the background model.
+The inset panel shows a $1.5^\circ$ by $1.5^\circ$ zoom-in around the Pal 5 cluster, with the Jacobi radius of the cluster shown as a large red circle.
+}
 \label{fig:members}

### 2019-10-01 paper/main.tex
marker consistency

@@
 \textbf{Middle}: The same, but for RRLs within $1\degr$ of the mean sky track of Pal 5 \citep{Bonaca:2019}.
-The square (blue) markers indicate RRLs that lie within the Jacobi radius of the cluster (see \sectionname~\ref{sec:results}).
+\changes{The open circle markers indicate RRLs that lie within the Jacobi radius of the cluster (see \sectionname~\ref{sec:results}) that pass this simple selection.}
 The over-density of stars near $(-2.7, -2.7)$ is the cluster and stream, and the two colored points show the Pal 5 cluster proper motions from \citet{Vasiliev:2019} and \citet{Fritz:2015}, respectively.
@@
 We note again that the membership probability does not depend on position or proximity to the stream, and only on distance and proper motions.
-\changes{Stars with probability $<0.1$ are plotted with a smaller marker in all panels.}
+\changes{In all panels, RRLs within the stream sky track (see \figurename~\ref{fig:members}) are plotted as star markers, and RRLs with membership probability $<0.1$ are plotted with a smaller marker.}
 \textbf{Upper left}: Sky positions (in the Pal 5 stream coordinate frame) of RRLs, with arrows showing the \Gaia\ \DR{2} proper motion vector direction.
@@
 \includegraphics[width=\textwidth]{members.pdf}
-\caption{Sky positions of RRLs with Pal 5 colored by membership probability, \changes{but stars with probability $<0.1$ are plotted with a smaller marker.}
+\caption{Sky positions of RRLs with Pal 5 colored by membership probability.
 The background greyscale shows the surface density of CMD-filtered main sequence stars from \citet{Bonaca:2019}.
@@
 Purple lines outline the stream track window defined in \citet{Bonaca:2019} and used here to exclude stream stars when constructing the background model.
+\changes{As with \figurename~\ref{fig:trackmembers}, RRLs within the sky track are shown as star markers, and RRLs with probability $<0.1$ are plotted with a smaller marker.}
 The inset panel shows a $1.5^\circ$ by $1.5^\circ$ zoom-in around the Pal 5 cluster, with the Jacobi radius of the cluster shown as a large red circle.

### 2019-10-01 paper/main.tex
new table

@@
     \bs{x} = \begin{pmatrix}
-        \mu_1^* + b_1 \, (\phi_1 - x_1) + c_1 \, (\phi_1 - x_1)^2\\
-        \mu_2^* + b_2 \, (\phi_1 - x_2) + c_2 \, (\phi_1 - x_2)^2\\
-        d^* + b_d \, (\phi_1 - x_d) + c_d \, (\phi_1 - x_d)^2
+        a_{\mu_1} + b_{\mu_1} \, (\phi_1 - x_{\mu_1}) + c_{\mu_1} \, (\phi_1 - x_{\mu_1})^2\\
+        a_{\mu_2} + b_{\mu_2} \, (\phi_1 - x_{\mu_2}) + c_{\mu_2} \, (\phi_1 - x_{\mu_2})^2\\
+        a_{d} + b_d \, (\phi_1 - x_d) + c_d \, (\phi_1 - x_d)^2
         \label{eq:meanpoly}
@@
 \end{equation}
-where $\bs{\theta} = (\mu_1^*, \mu_2^*, d^*, b_1, b_2, b_d, c_1, c_2, c_d, x_1, x_2, x_d)$ are free parameters.
+where $\bs{\theta} = (a_{\mu_1}, a_{\mu_2}, a_d, b_{\mu_1}, b_{\mu_2}, b_d, c_{\mu_1}, c_{\mu_2}, c_d, x_{\mu_1}, x_{\mu_2}, x_d)$ are free parameters.
 
@@
 \tablename~\ref{t:rrl_members} shows an abbreviated version of our catalog of RRL in the region around Pal 5, with PS1 and \Gaia~ data and membership probabilities appended.
+\changes{\tablename~\ref{t:trendpars} contains the maximum a posteriori trend parameter values from the above analysis.}
 
@@
 
+\begin{table}
+\begin{center}
+\caption{Maximum a posteriori proper motion and distance trend parameters (see \sectionname~\ref{sec:membership}).
+The constant terms have units $\masyr$ and $\kpc$ for proper motion and distance parameters, respectively.
+The angular units are in degrees such that, e.g., the linear coeff $b_d$ has units of $\kpc~\textrm{deg}^{-1}$.}\label{t:trendpars}
+\begin{footnotesize}
+\begin{tabular}{c c c c c}
+\toprule
+y & $x_{\textrm{y}}$ & $a_\textrm{y}$ & $b_\textrm{y}$ & $c_\textrm{y}$\\
+param. name & reference $\phi_1$ & constant & linear coeff. & quadratic coeff.\\
+\midrule
+$\mu_1$ &  $-1.072$ & 3.740  & $4.102 \times 10^{-2}$  & $-6.423 \times 10^{-4}$ \\
+$\mu_2$ & $-10.954$ & 0.686  & $-2.826 \times 10^{-2}$ & $2.832 \times 10^{-3}$ \\
+$d$     & $-16.081$ & 22.022 & $9.460 \times 10^{-3}$  & $-6.327 \times 10^{-3}$\\
+\bottomrule
+\end{tabular}
+\end{footnotesize}
+\end{center}
+\end{table}
+
 \begin{table*}[t]

### 2019-10-01 paper/main.tex
Updates from Overleaf

@@
 Interestingly, there seems to be a strong segregation of the RRL types: all but two of the \NRRab\ \rrab~are found in the cluster tails and two thirds of the \NRRc\ \rrc~are found in the cluster itself.
-At fixed metallicity and helium abundance, position along the horizontal branch is determined by (present-day) mass, with less massive stars ending up at lower temperatures as type \typeab~stars.
-The dichotomy of RRL types between the stream and cluster is therefore suggestive of being a signature of mass segregation: The (slightly) higher mass \rrc\ stars seem to be preferentially more bound to the cluster.
-However, the mass difference between \rrab\ and \rrc\ stars is expected to be small, $\lesssim 0.1~\msun$, and so unlikely to have a significantly different dynamical history---this remains an intriguing puzzle.
+\changes{
+This could be a stochastic effect given the relatively small number of stars, but it seems unlikely given the observed ratios of ab-to-c stars in globular clusters: At the very least, \rrab~stars should account for about half the number of \rrc's \citep[only 5 out of >150 globulars have more than twice as many type~\typec\ than type \typeab\  RRLs, see][]{Clement:2017,Clement:2001,Catelan:2009}.
+Given the remarkably high completeness of our catalogue, any additional RRL stars would have been detected.
+Another possibility is that this segregation has a dynamical origin: At fixed metallicity and helium abundance, position along the horizontal branch is determined by (present-day) mass, with less massive stars ending up at higher temperatures as type \typec~stars.
+However, if the RRL type segregation had a dynamical origin, the relevant mass should be the one the stars had prior to being ejected from the cluster.} In either case, any of these possible mass differences (either at the main-sequence or the zero-age horizontal branch) are expected to be very small ($\lesssim 0.1~\msun$) and so unlikely to have a significantly different dynamical history---this remains an intriguing puzzle.
 

### 2019-10-01 paper/main.tex
gray
Updates from Overleaf

@@
 
-We also detect a distance trend from the closer leading tail ($d \approx 19.4~\kpc$ at $\phi_1 \approx 5\degr$) to the more distant trailing tail ($d \approx 21.3~\kpc$ at $\phi_1 \approx -5\degr$).
+We also detect a distance trend from the closer leading tail \changes{($d \approx 19.4~\kpc$ and $\approx$$18~\kpc$ at $\phi_1 \approx 5\degr$ and $10\degr$, respectively) to the more distant trailing tail ($d \approx 21.3~\kpc$ and $\approx$$21.9~\kpc$ at $\phi_1 \approx -5\degr$ and $-10\degr$, respectively)}.
 While the proper motion in right ascension appears consistent with constant or only a mild gradient over the extent of the stream, we see a significant gradient in the declination component, with a difference of about $\approx 1~\masyr$ from trailing to leading tail.
@@
 
+\begin{figure*}[t]
+\begin{center}
+\includegraphics[width=0.85\textwidth]{rrls_PA.pdf} 
+\caption{Amplitude ($g$-band) versus Period for all RRLs in the Pal~5 field.
+\changes{Stars with membership probabilities $>0.1$ are colored by membership probability, with those inside the Pal~5 track shown with large (star) markers.
+RRLs with membership probabilities $<0.1$ are shown as gray markers.} The distribution of the RRLs found in the cluster and its tails is consistent with an Oosterhoff type I (dashed line).
+}
+\label{fig:PA_diagram}
+\end{center}
+\end{figure*}
+
 Interestingly, there seems to be a strong segregation of the RRL types: all but two of the \NRRab\ \rrab~are found in the cluster tails and two thirds of the \NRRc\ \rrc~are found in the cluster itself.
@@
 However, if the RRL type segregation had a dynamical origin, the relevant mass should be the one the stars had prior to being ejected from the cluster.} In either case, any of these possible mass differences (either at the main-sequence or the zero-age horizontal branch) are expected to be very small ($\lesssim 0.1~\msun$) and so unlikely to have a significantly different dynamical history---this remains an intriguing puzzle.
-
-\begin{figure*}[t]
-\begin{center}
-\includegraphics[width=\textwidth]{rrls_PA.pdf}
-\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field membership probabilities $>0.1$ (small markers) inside the Pal~5 track (large markers), colored by membership probability.
-The distribution of the RRLs found in the cluster and its tails is consistent with an Oosterhoff type I (dashed line).
-}
-\label{fig:PA_diagram}
-\end{center}
-\end{figure*}
 

### 2019-10-03 paper/main.tex
add dai reference

@@
 The detailed but still limited kinematic data for Pal 5 has made it a canonical example of tidal destruction and a useful tool for constraining Galactic structure.
-For example, using the sky track and radial velocity information, the stream has been used to measure the enclosed mass of the Milky Way \citep{Kuepper:2015, Bovy:2016}.
+For example, using the sky track and radial velocity information, the stream has been used to measure the enclosed mass of the Milky Way \citep{Kuepper:2015, Bovy:2016, Dai:2018}.
 Using deep photometry, the density variations along the stream have been used to place limits on the abundance of massive substructures in the Galactic halo \citep{Erkal:2017}.
@@
 \caption{Sky positions of RRLs with Pal 5 colored by membership probability.
-The background greyscale shows the surface density of CMD-filtered main sequence stars from \citet{Bonaca:2019}.
+The background grayscale shows the surface density of CMD-filtered main sequence stars from \citet{Bonaca:2019}.
 Note that the membership probability does not incorporate sky position information, but a clear over-density of high-probability RRL members do trace the stream on the sky.
@@
 Adopting $\feh = -1.35$ \citep{Ishigaki:2016} and using the updated RRL $M_V$--$\feh$ relation from \citet{Muraveva:2018}, we find that this past RRL-based distance to Pal 5 is instead $d_{\rm Pal 5} = 21.2 \pm 0.6~\kpc$.
+\changes{Interestingly, \citet{Dai:2018} also independently infer a nearer distance to the cluster around $d_{\rm Pal 5} \approx 21~\kpc$ from dynamical modeling of the stream.}
 To summarize: Our inferred distance to the Pal 5 cluster, $d_{\rm Pal 5} = \clderr$, is consistent with (but more precise than) all recent distance measurements, once corrected for dust extinction and updated calibrations of the horizontal branch absolute magnitude.
